---
title: "Choice data"
author: "Lennart Oelschläger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choice data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains how to handle choice data in RprobitB. We use the [Train data set from the mlogit package](https://cran.r-project.org/package=mlogit) for illustration.

```{r setup}
library(RprobitB)
```


## Characteristics and requirements of choice data

Choice data basically consists of

1. an identifier for each decider and optionally for each choice situation,

2. the choices,

3. alternative and decider specific covariates.


Consequently, RprobitB requests the following requirements to provided choice data:

1. It **must** be of class `data.frame`.

2. It **must** be in "wide" format, i.e. each row represents one choice occasion.

3. It **must** contain a column with unique identifier for each decision maker. 

4. It **can** contain a column with identifier for each choice situation of each decider. If this information is missing, these identifier are generated automatically by the appearance of the choices in the data set. 

5. It **can** contain a column with the observed choices. Such a column is required for model fitting but not for prediction.

6. It **must** contain columns for the values of each alternative specific covariate for each alternative and for each decider specific covariate.


### Example

We can check that the Train data set[^1] fulfills these requirements:

```{r overview Train data}
data("Train", package = "mlogit")
str(Train)
```

It is a data frame, each row represents one choice occasion, the columns `id` and `choiceid` identify the deciders and the choice occasions, respectively. The column `choice` gives the observed choices. We see four alternative-specific covariates, namely `price`, `time`, `change` and `comfort`. There values are given for each alternative.

[^1]: The data set contains 2929 stated choices by 235 Dutch individuals deciding between two train trip options based on the price, the travel time, the level of comfort, and the number of changes. 


## Specify the model formula

We have to tell RprobitB which covariates we want to include in our model via specifying a `formula` object. The structure of such a `formula` object is `choice ~ A | B | C`, where

- `choice` is the dependent variable (the discrete choice we aim to explain),

- `A` are alternative and choice situation specific covariates with a generic coefficient (we call them covariates of type 1),

- `B` are choice situation specific covariates with alternative specific coefficients (we call them covariates of type 2),

- and `C` are alternative and choice situation specific covariates with
alternative specific coefficients (we call them covariates of type 3).


Specifying a `formula` object for RprobitB must be consistent with the following rules:

- By default, alternative specific constants (ASCs) are added to the model[^2]. They can be removed by adding `+ 0` in the second spot, e.g. `choice ~ A | B + 0`.

- To exclude covariates of the backmost categories, use either `0`, e.g. `choice ~ A | B | 0` or just leave this part out and write `choice ~ A | B `. However, to exclude covariates of front categories, we have to use `0`, e.g. `choice ~ 0 | B`. 

- To include more than one covariate of the same category, use `+`, e.g. `choice ~ A1 + A2 | B`.

- If we don't want to include any covariates of the second category but we want to estimate alternative specific constants, add `1` in the second spot, e.g. `choice ~ A | 1`. The expression `choice ~ A | 0` is interpreted as no covariates of the second category and no alternative specific constants.

To impose random effects for specific variables, we need to define a character vector `re` with the corresponding variable names. To have random effects for the alternative specific constants, include `"ASC"` in `re`.

[^2]: ASCs capture the average effect on utility of all factors that are not included in the model. Due to identifiability, we cannot estimate ASCs for all the alternatives. Therefore, they are added for all except for the last alternative.


### Example

We specify a model formula for the Train data set. Say we want to include all the covariates `price`, `time`, `comfort` and `change`, which are all alternative specific, so either of type 1 or type 3. The difference between type 1 and type 3 is that in the former case we would estimate a generic coefficient (i.e. a coefficient that is constant across alternatives), whereas in the latter case we would estimate alternative specific coefficients. 

Deciding between type 1 and type 3 for these covariates belongs into the topic of model selection, for which we provide a separate vignette. For now, we go with type 1 for all covariates and add ASCs: 

```{r train formula}
form <- choice ~ price + time + comfort + change | 1
```

Additionally, we specify random effects for `price` and `time` (because we would typically expect heterogeneity here):

```{r train re}
re <- c("price","time")
```

RprobitB provides the function `check_form()` which can be used to check if `form` and `re` are correctly interpreted:

```{r check form}
check_form(form = form, re = re)
```


## The `prepare_data()` function in RprobitB

Before model estimation with RprobitB, any empirical choice data set `choice_data`  must go through the `prepare_data()` function:

```{r, eval = FALSE}
data <- prepare_data(form = form, choice_data = choice_data)
```

The function performs compatibility checks and data transformations and returns an object of class `RprobitB_data` that can be fed into the estimation routine. The following arguments are optional:

- `re`: The character vector of variable names of `form` with random effects.

- `alternatives`: We may not want to consider all alternatives in `choice_data`. In that case, we can specify a character vector `alternatives` with selected names of alternatives. If not specified, the choice set is defined by the observed choices.

- `id`: A character, the name of the column in `choice_data` that contains a unique identifier for each decision maker. The default is `"id"`. 

- `idc`: A character, the name of the column in `choice_data` that contains a unique identifier for each choice situation given the decision maker. Per default, these identifier are generated by the appearance of the choices in the data set.  

- `standardize`: A character vector of variable names of `form` that get standardized. Covariates of type 1 or 3 have to be addressed by `<covariate>_<alternative>`. If `standardize = "all"`, all covariates get standardized.


### Example

Let's prepare the Train data set for estimation with our previous specification of `form` and `re`:

```{r prepare Train}
data <- prepare_data(form = form, choice_data = Train, re = re)
```

The `summary` function gives an overview over the data:

```{r summary Train}
summary(data)
```


## Simulate choice data in RprobitB using the `simulate_choices()` function

RprobitB provides the function `simulate_choices` that simulates discrete choice data. Say we want to simulate the choices of `N` deciders in `T` choice occasions[^3] among `J` alternatives from a model formulation `form`, we have to call

```{r, eval = FALSE}
data <- simulate(form = form, N = N, T = T, J = J)
```

The function `simulate()` has the following optional arguments:

- `re`: The character vector of variable names of `form` with random effects.

- `alternatives`: A character vector of length `J` with the names of the choice alternatives. If not specified, the alternatives are labeled by the first `J` upper-case letters of the Roman alphabet.

- `covariates`: A named list of covariate values. Each element must be a vector of length equal to the number of choice occasions and named according to a covariate. Covariates for which no values are specified are drawn from a standard normal distribution.

- `standardize`: A character vector of variable names of `form` that get standardized, see [below](#standardize-covariates).

- `seed`: Set a seed for the simulation.

We can specify true parameter values[^4] by adding values for

- `alpha`, the fixed coefficient vector,
  
- `C`, the number (greater or equal 1) of latent classes of decision makers,
  
- `s`, the vector of class weights,
  
- `b`, the matrix of class means as columns,
  
- `Omega`, the matrix of class covariance matrices as columns,
  
- `Sigma`, the differenced error term covariance matrix,
  
- `Sigma_full`, the full error term covariance matrix.

[^3]: `T` can be either a positive number, representing a fixed number of choice occasions for each decision maker, or a vector of length `N` with decision maker specific numbers of choice occasions.

[^4]: See the vignette "Introduction" for more details.


### Example

For illustration, we simulate a data set of `N = 100` deciders choosing at `T = 10` choice occasions among the `J = 2` alternatives bus and car:

```{r data sim meta}
N <- 100
T <- 10
J <- 2
alternatives <- c("car", "bus")
```

We specify `form` and `re` as follows:

```{r data sim form}
form <- choice ~ cost | income | time
re <- c("cost", "time")
```

Finally, we may want to specify true parameter values. Parameter values that are not specified get randomly drawn. RprobitB provides the function `overview_effects()` which can be used to get an overview over the effects for which parameters can be specified. Below, the column `re` contains booleans that indicate whether the effect is connected to a random effect.

```{r data sim overview}
overview_effects(form = form, re = re, alternatives = alternatives)
```

Hence, the coefficient vector `alpha` must be of length 3, where the first coefficient is for `income_bus` and the second for `ASC_bus`. The matrix `b` must be of dimension `C` x 3, where by default `C = 1` and row 1 to 3 correspond to `cost`, `time_bus`, and `time_car`, respectively.

An exemplary call of `simulate_choices()` looks like follows, where `summary()` can again be used to get an overview over the data:

```{r data sim}
data <- simulate_choices(
  form = choice ~ cost | income | time,
  N = 100,
  T = 10,
  J = 2,
  re = c("cost", "time"),
  alternatives = c("car", "bus"),
  seed = 1,
  alpha = c(-1,1),
  C = 2
)
summary(data)
``` 

## Train and test data set

RprobitB provides the function `train_test()` that can be used to split the output of `prepare_data()` or `simulate_choices()` into a train and a test subset. This is useful when evaluating the prediction performace of a fitted model.

For example, the following code puts 70% of deciders from `data` into the train subsample and 30% of deciders in the test subsample:

```{r data split deciders}
train_test(data, test_proportion = 0.3, by = "N")
```

Alternatively, the following code puts 2 randomly chosen choice occasions per decider from `data` into the test subsample, the rest goes into the train subsample:

```{r data split occasions}
train_test(data, test_number = 2, by = "T", random = TRUE, seed = 1)
```
