---
title: "Introduction"
author: "Lennart Oelschläger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to RprobitB!

## For what tasks is this package designed?

With RprobitB[^1] you can 

1. analyze choices made by deciders among a discrete set of alternatives,

2. estimate [(latent class) (mixed) (multinomial) probit models](#levels-of-model-complexity) in a Bayesian framework,

3. model heterogeneity by approximating any underlying mixing distributions through a mixture of normal distributions,

4. identify latent classes of decision makers.

Run `install.packages("RprobitB")` in your R console to install the latest version of RprobitB.

[^1]: The package name RprobitB is a portmanteau, combining **R** (the programming language), **probit** (the model class) and **B** (for Bayes, the estimation method).

Why the notation **(latent class) (mixed) (multinomial) probit model**? Because RprobitB can fit probit models of increasing complexity:

- Most basic, modelling the choice between two alternatives (the **probit model**).

- Considering more than two alternatives leads to the **multinomial probit model**.

- If we incorporate random effects, the model gets the prefix *mixed*.

- The most general model is the **latent class mixed multinomial probit model**, which approximates the mixing distribution through a mixiture of normal distributions.

## A package overview

```{r, echo = FALSE}
library(DiagrammeR)
mermaid("
graph TD
    prepare --> RprobitB_data
    simulate --> RprobitB_data
    RprobitB_data --> mcmc
    mcmc --> RprobitB_fit
")
```

1. [How is the model defined?](#the-model)

2. [How to prepare or simulate choice data?](data_management.html)

3. [How to fit a model?](model_fitting.html)
