---
title: "Get started with RprobitB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with RprobitB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
link-citations: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
  fig.align = "center"
)
```

People make choices all the time: which train to take, which electricity
supplier to sign with, whether to halve the chess clock for the chance of an
extra tournament point. Discrete choice models turn such decisions into
statements about preferences, and **RprobitB** [@Oelschlaeger2026d] estimates
them in a Bayesian probit framework. This vignette fits one model to real
panel data and walks through the summaries, diagnostics, and plots that
every later analysis builds on. The four other vignettes take it from
there, and the last section says which one answers which question.

## The model in brief

Suppose a decider faces $J$ alternatives and picks the one they like best.
The probit model gives "like best" a precise meaning by attaching a latent
utility to every alternative,

$$U_{ntj} = X_{ntj}' \beta_n + \epsilon_{ntj}, \qquad \epsilon_{nt} \sim
\mathcal{N}(0, \Sigma),$$

where $n$ indexes deciders, $t$ their choice occasions, and $X_{ntj}$ holds
the covariates that the model formula assigns to alternative $j$. The
alternative with the largest utility is the one that gets chosen, and the
choice is all we observe. Two things therefore cannot be learned from
choices alone: the level of utility and its scale. **RprobitB** works with
utility differences relative to a base alternative, which settles the level,
and fixes either one variance or one coefficient, which settles the scale.
The coefficient vector $\beta_n$ is shared by all deciders unless you declare
random effects, in which case every decider draws their own coefficients
from a population distribution.
`vignette("v02_model_variants")` spells out the normalization and the prior,
and `vignette("v03_heterogeneity")` the ways of letting coefficients differ.
Bayesian estimation of the multinomial probit model goes back to
@McCulloch1994 and @Imai2005a, and @Oelschlaeger2026c gives a unified
account of the heterogeneity models that **RprobitB** implements.

## Which train would you take?

In 1987, a survey commissioned by the Dutch national railways presented 235
travelers with pairs of hypothetical train trips and asked which one they
would book [@BenAkiva1993]. The two trips of a pair differ in
price, travel time, number of changes, and comfort class, and every traveler
answered about a dozen such questions. The **mlogit** package
[@Croissant2020] ships the 2929 answers as `Train`. The data are in wide
format: one row per question, with the attributes of trip `A` in columns
ending in `_A` and those of trip `B` in columns ending in `_B`. The prices
are recorded in cents of Dutch guilders and the travel times in minutes;
converting them to euro and hours makes the estimates easier to read.
These travelers come back twice: in `vignette("v03_heterogeneity")` they
split into two classes with different values of time, and in
`vignette("v04_prediction")` every one of them gets their own price
coefficient.

```{r data}
library(RprobitB)
set.seed(1)
data("Train", package = "mlogit")
Train$price_A <- Train$price_A / 100 / 2.20371
Train$price_B <- Train$price_B / 100 / 2.20371
Train$time_A <- Train$time_A / 60
Train$time_B <- Train$time_B / 60
head(Train[, 1:7])
```

The formula `choice ~ A | B | C` sorts covariates into three groups:

- `A`: attributes of the alternatives, such as the price of a trip, that
  share one coefficient across alternatives,
- `B`: characteristics of the decider, such as income, that do not vary
  across alternatives and therefore get alternative-specific coefficients,
- `C`: attributes of the alternatives with alternative-specific coefficients.

All four train attributes are of type `A`. The comfort class is stored as
a number, but it is a category with three levels, and `factor(comfort)`
turns it into dummy variables the way it does in `lm()`. The `0` in the
second part drops the alternative-specific constants, which makes sense
here: the labels `A` and `B` were handed out arbitrarily, so nobody prefers
one label over the other. Finally, `column_decider` names the traveler and
`column_occasion` the question, so that `fit()` knows which rows belong to
the same traveler. Their answers then share one set of coefficients and
enter the likelihood together. `vignette("v02_model_variants")` treats the
three covariate types in detail.

```{r fit}
model <- fit(
  choice ~ price + time + change + factor(comfort) | 0,
  data = Train,
  column_decider = "id",
  column_occasion = "choiceid",
  iterations = 2000,
  warmup = 1000,
  chains = 2,
  progress = FALSE
)
model
```

Four chains with 1000 iterations each are the default. Two chains of 2000
iterations draw the same number of times in total, and two chains are enough
for the convergence checks below.
The `set.seed(1)` above makes the run reproducible, also when the chains run
in parallel.

## Reading the posterior

`summary()` reports the posterior mean, mode, and standard deviation of every
population-level parameter together with the convergence diagnostics of the
`posterior` package [@Buerkner2026]. Add `probs` if you also want
quantiles.

```{r summary}
summary(model)
```

The coefficient names follow the formula. Price, time, and changes all pull
utility down, as they should. Comfort is a factor whose level `0` is the
best class, so the negative coefficients of the dummies for levels `1` and
`2` say that travelers dislike the poorer classes. One parameter is missing
from the table: `Sigma[B,B]`, the error variance of the utility difference,
is fixed to one by the default normalization, so there is nothing to
summarize. It remains available in the draws.

Because of this normalization, coefficients are measured in units of the
error scale, so their absolute size means little on its own. The ratio of
two coefficients is free of that scale, and it answers the more interesting
question anyway: what would compensate a traveler for one more hour on the
train? `interpret()` computes such ratios draw by draw and expresses every
effect in the units of a reference effect, here the price:

```{r interpret}
compensation <- interpret(model, reference = "price")
compensation
```

@BenAkiva1993 collected these data to estimate what an hour of travel time
is worth, and found, as here, that price, travel time, changes, and the
poorer comfort classes all make a trip less attractive. Their model let the
value of time differ between travelers, which is what
`vignette("v03_heterogeneity")` does with the same data; the single
coefficient per attribute used here averages over them.

A negative value is a price reduction. Read this way, these travelers would
pay about
`r round(abs(compensation$mean[compensation$effect == "time"]), 1)` euro (at
1987 prices) to shorten a trip by one hour, and about
`r round(abs(compensation$mean[compensation$effect == "factor(comfort)2"]), 1)`
euro to travel in the best rather than the worst comfort class. The intervals
come from the posterior distribution of the ratios, not from the ratio of
two posterior means. `interpret()` can also read the estimates as changes of
the choice probabilities, through `type = "ame"` and `type = "mea"`, which
`vignette("v04_prediction")` covers.

Back to the summary table. Its last two columns judge how well the sampler
did its job and follow the current recommendations [@Vehtari2021]:

- `rhat` compares the chains with each other and should stay below 1.01.
- `ess_bulk` is the effective sample size for the center of the posterior.
  A few hundred draws are plenty for posterior means.

Two arguments change what the table shows. `probs` adds quantiles, and
`statistics` picks the measures, among them the tail effective sample size
that governs quantiles and the Monte Carlo standard error of the mean, which
says how much of the reported digits is sampling noise. `?summary.RprobitB_fit`
explains each measure.

```{r summary-statistics}
summary(
  model,
  statistics = c("mean", "mcse_mean", "ess_bulk", "ess_tail"),
  probs = c(0.05, 0.95)
)
```

The Monte Carlo standard errors are about twenty times smaller than the
posterior standard deviations, so the sampler adds little noise to what the
data say.

`coef()` and `confint()` extract point estimates and equal-tailed credible
intervals.

```{r coef}
coef(model)
coef(model, type = "median")
confint(model, level = 0.9)
```

Mean and median agree here, which is what one expects from posteriors as
symmetric as these. `vcov()` adds the posterior covariances, which say how
the estimates move together:

```{r vcov}
round(vcov(model), 5)
```

The comfort coefficients share the largest covariance, because the two
dummies describe the same factor. The same information as a picture, with
the thick bars covering the middle half of each posterior and the thin lines
its middle 90 percent:

```{r interval}
plot(model, type = "interval")
```

The price coefficient looks like a point only because it is small next to
the others. Every coefficient carries the scale of its own covariate, and
prices run into tens of euro while the comfort dummies are zero or one.

## A look at the chains

`plot()` delegates to `bayesplot` [@Gabry2025]. Trace plots show how the
chains wander through the parameter space, and density overlays compare their
marginal posteriors. Chains that have converged look like stationary noise
without any trend, and their densities lie almost on top of each other.

```{r trace}
plot(model, type = "trace", variables = c("beta[price]", "beta[time]"))
plot(model, type = "density", variables = c("beta[price]", "beta[time]"))
```

Rank plots are the sharper test. They pool the draws of all chains, rank
them, and show each chain's share of the ranks: with converged chains every
histogram is flat, and a chain stuck in its own region shows up as a peak.
Autocorrelation plots say how quickly a chain forgets where it was, which is
what the effective sample size summarizes in one number.

```{r rank}
plot(model, type = "rank", variables = c("beta[price]", "beta[time]"))
plot(model, type = "acf", variables = c("beta[price]", "beta[time]"))
```

Pairs plots show two parameters at once and reveal the posterior
correlations that make a sampler slow.

```{r pairs}
plot(model, type = "pairs", variables = c("beta[price]", "beta[time]"))
```

Any additional argument is passed on to the `bayesplot` function.

## When the diagnostics complain

Diagnostics are only useful if you act on them. The fit below is
deliberately too short: 200 iterations with 100 of them warmup.

```{r short}
short <- update(model, iterations = 200, warmup = 100)
summary(short, statistics = c("mean", "sd", "rhat", "ess_bulk"))
```

The estimates are not far off, but the effective sample sizes stay below a
hundred and `rhat` climbs above the threshold of 1.01: the two chains have
not met yet. The remedy is more iterations, and the same fit
with ten times as many puts every diagnostic in order.

```{r longer}
longer <- update(short, iterations = 2000, warmup = 1000)
summary(longer, statistics = c("mean", "sd", "rhat", "ess_bulk"))
```

If more iterations do not help, the problem is usually the model rather than
the sampler: a coefficient that the data cannot identify, or a mixture whose
classes overlap. `vignette("v03_heterogeneity")` returns to that case.

## Working with the draws

Nothing is locked away inside the fitted object. The retained draws are
stored as a `draws_array` of the `posterior` package with dimensions
iteration, chain, and variable, so every tool of that package applies
directly.

```{r draws}
draws <- posterior::as_draws(model)
dim(draws)
posterior::summarise_draws(draws, "mean", "quantile2")
```

The usual accessors work as well: `formula()` returns the fitted formula,
`model.frame()` the data the model was fitted to, and `nobs()` the number of
independent units in the likelihood, which for panel data is the number of
deciders rather than the number of choices.

```{r accessors}
formula(model)
nobs(model)
head(model.frame(model)[, 1:4])
```

## Running chains in parallel

**RprobitB** never selects a parallel backend on its own. Independent chains run
through the `future` framework [@Bengtsson2021], so a plan set before fitting
is picked up automatically.

```{r parallel, eval=FALSE}
future::plan(future::multisession, workers = 4)
parallel_model <- fit(
  choice ~ price + time + change + factor(comfort) | 0,
  data = Train,
  column_decider = "id",
  column_occasion = "choiceid"
)
future::plan(future::sequential)
```

With `progress = TRUE`, the default in interactive sessions, all chains
report their progress through the `progressr` package [@Bengtsson2026]. The
marginal effects of `interpret()` and the log-likelihood draws behind
`WAIC()` and `loo()` follow the same plan and report progress in the same
way.

## How long does this take?

The fit above draws 2000 iterations for 2929 choices in about a second, and
the cost grows roughly in proportion to iterations times choices. Three
things change the picture:

- **Random coefficients.** Every decider gets their own coefficient vector
  in every iteration, so the cost grows with the number of deciders and with
  the square of the number of random coefficients.
  `vignette("v03_heterogeneity")` fits such models in seconds for a few
  hundred deciders.
- **Alternatives.** A free error covariance between many alternatives mixes
  slowly and needs longer chains, not faster ones.
  `vignette("v02_model_variants")` shows a four-alternative model that needs
  fifteen thousand iterations before its covariance entries settle.
- **Model evaluation.** `WAIC()`, `loo()`, and `bayes_factor()` evaluate the
  likelihood for every retained draw, so they can cost more than the fit
  itself when the model has random coefficients.

When a fit or an evaluation is slow, the two levers are the future plan
above and `thin`, which keeps every `thin`-th draw and shrinks everything
downstream.

## Where to go next

- Is my model specified the way I mean it?
  `vignette("v02_model_variants")` covers the normalization, the prior, the
  covariate types, individual choice sets, and ordered and ranked responses.
- Do all deciders share these preferences?
  `vignette("v03_heterogeneity")` lets coefficients vary continuously over
  the population or differ between latent classes.
- What does the model predict, and what does a coefficient mean in
  probabilities? `vignette("v04_prediction")` covers predictions for the
  population and for single deciders, scenarios, and marginal effects.
- Which of two models should I believe?
  `vignette("v05_model_evaluation")` compares them with WAIC, PSIS-LOO, and
  Bayes factors.

The first two work through each variant twice: once on simulated data, where
the truth is known and can be compared with the estimates, and once on a
real data set.

## References
