---
title: "Model specification and variants"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model specification and variants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
link-citations: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
  fig.align = "center"
)
```

This vignette describes how to specify a probit choice model with `fit()`.
It covers the normalization of the utility scale and the prior distribution,
which every fit involves, and then the model variants: the three types of
covariates and the alternative-specific constants, choice sets that differ
between occasions, and ordered and ranked responses. Preference
heterogeneity between deciders is the subject of the vignette
[Modeling preference heterogeneity][v03], and @Oelschlaeger2026c gives the
methodological background. From the covariate types onward, each
section first fits the variant to simulated data and then to a data set of
the **AER** package [@Kleiber2008], the **MASS** package
[@VenablesRipley2002], or the **mlogit** package [@Croissant2020]. Without
data, `fit()` simulates the requested model before estimating it, and
`summary()` prints the true values in a `dgp` column beside the posterior
summaries, converted to the normalization of the fit.

```{r setup}
library(RprobitB)
set.seed(1)
```

## Normalization

The probit model introduced in the vignette
[Get started with RprobitB][v01] assigns every alternative $j$ at occasion
$t$ of decider $n$ the latent utility
$U_{ntj} = X_{ntj}^\top \beta_n + \epsilon_{ntj}$ with jointly normal errors
$\epsilon_{nt} \sim \mathrm{N}(0, \Sigma)$, and the decider chooses the
alternative with the largest utility. The choice probability of an
alternative is the probability that its utility exceeds the utilities of all
others. These probabilities are invariant to adding a constant to all
utilities and to multiplying all utilities by a positive number, so neither
the level nor the scale of the utilities is identified. **RprobitB**
therefore works with utility differences relative to a base alternative, by
default the first in alphabetical order, and fixes one parameter through
`scale`:

- `scale = NULL` fixes the error variance of the first utility difference to
  one. This is the default. The covariance matrix of the error differences is
  reported as `Sigma[<alternative>,<alternative>]`, with rows and columns
  named by the alternatives other than the base. For three alternatives `A`,
  `B`, and `C` with base `A`, its entries are `Sigma[B,B]`, `Sigma[C,B]`, and
  `Sigma[C,C]`.
- `scale = c(z = -1)` fixes the coefficient of `z` to `-1` instead, so that
  all other coefficients are measured in units of it. With a price
  coefficient fixed in this way, the other coefficients are willingness-to-pay
  values.

The sampler draws all parameters without restriction and rescales every
retained draw afterwards. Variables that the normalization fixes remain in
the draws but are omitted from `summary()`, `coef()`, and `vcov()`.

The following demonstration shows the effect of the normalization on the
reported values. We simulate data with coefficients `1` for `x` and `-0.5`
for `z` and fit them under the default scale. Simulated alternatives are
labeled with capital letters unless `alternatives` names them, here `A` and
`B`, and `A` is the base. `update()` then refits the same simulated data
with the coefficient of `z` fixed to `-1`, so only the normalization differs
between the two fits.

```{r normalization}
scale_default <- fit(
  choice ~ x + z | 0,
  dgp_parameters = list(beta = c(x = 1, z = -0.5)),
  n_deciders = 300,
  chains = 1
)
summary(scale_default)
scale_z <- update(scale_default, scale = c(z = -1))
summary(scale_z)
```

```{r normalization-deviation, include=FALSE}
scale_table <- summary(scale_z)$posterior
scale_deviation <- max(
  abs(scale_table$mean - scale_table$dgp) / scale_table$sd
)
```

Under the default scale, the `dgp` column shows the coefficients as
specified. `Sigma[B,B]`, the error variance of the utility difference
between alternative `B` and the base `A`, is fixed to one and therefore not
listed. Under
the coefficient normalization, `beta[z]` is fixed to `-1` instead of its
true `-0.5`, so all utilities are multiplied by `2`: the true `beta[x]`
becomes `2` and the true error variance `4`. `summary()` converts the true
values to the normalization of the fit, so the `dgp` column remains
comparable.

The convergence diagnostics of the two tables differ because `update()`
runs the sampler again and because the rescaled variables are different
functions of the draws: `beta[x]` is now a ratio of two coefficients, which
mixes differently than a single coefficient.

## Prior distribution

**RprobitB** estimates every model with a Gibbs sampler, which draws each
block of parameters in turn from its conditional posterior distribution given
the other parameters and the latent utilities. The prior is conjugate for
every block, so each conditional posterior belongs to the same family as the
prior. Fixed coefficients and class means have normal priors, covariance
matrices have inverse Wishart priors, class weights have a Dirichlet prior,
and the log-increments between ordered thresholds have a normal prior. The
section "Prior distribution" of `?fit` lists all components with their
defaults, which are weakly informative for coefficients and covariances.
Single components are overridden through a named list, and the complete
prior of a fit is stored in its `prior` component. The latent-utility data
augmentation of the sampler goes back to @Albert1993, and @Imai2005a
developed its marginal augmentation for the multinomial probit model.

The following demonstration simulates 100 deciders with a true coefficient
of `-1` under the default prior and then refits the same data under two
priors with mean `1`: a moderate one with variance `0.5` and a tight one
with variance `0.01`.

```{r prior}
default_prior <- fit(
  choice ~ x | 0,
  dgp_parameters = list(beta = c(x = -1)),
  chains = 1
)
default_prior$prior
summary(default_prior)
moderate_prior <- update(
  default_prior, prior = list(fixed_mean = 1, fixed_covariance = matrix(0.5))
)
tight_prior <- update(
  default_prior, prior = list(fixed_mean = 1, fixed_covariance = matrix(0.01))
)
data.frame(
  variable = "beta[x]", dgp = -1, default = coef(default_prior),
  moderate = coef(moderate_prior), tight = coef(tight_prior),
  row.names = NULL
)
```

```{r prior-deviation, include=FALSE}
prior_deviation <- abs(coef(default_prior) + 1) /
  summary(default_prior)$posterior$sd
```

Under the default prior, the posterior mean deviates from the true value by
`r round(prior_deviation, 1)` posterior standard deviations. The moderate
prior shifts the posterior mean by about
`r round(abs(coef(moderate_prior) - coef(default_prior)), 2)` towards the
prior mean. The tight prior has a standard deviation of `0.1` around `1` and
moves the posterior mean to `r round(coef(tight_prior), 2)`. A prior on a
coefficient is unproblematic as long as its variance reflects the actual
uncertainty about the coefficient.

## Covariate types and alternative-specific constants

The formula `choice ~ A | B | C` distinguishes three kinds of covariates:
attributes of the alternatives with one shared coefficient (`A`),
characteristics of the decider with alternative-specific coefficients (`B`),
and attributes of the alternatives with alternative-specific coefficients
(`C`). Alternative-specific constants, one intercept per alternative other
than the base that captures the utility of the alternative beyond its
attributes, are included by default and removed with `0` in the second part.
The following demonstration simulates all three types at once, and
`summary()` shows the posterior summaries beside the normalized true values.

```{r covariate-types-sim}
covariate_types <- fit(
  choice ~ x | z | w,
  n_deciders = 500,
  dgp_parameters = list(beta = c(
    x = 0.5, z_B = -0.5, ASC_B = 0.25, w_A = -0.5, w_B = 0.5
  )),
  chains = 1
)
summary(covariate_types)
```

```{r covariate-types-deviation, include=FALSE}
types_table <- summary(covariate_types)$posterior
types_deviation <- max(
  abs(types_table$mean - types_table$dgp) / types_table$sd
)
types_signs <- all(sign(types_table$mean) == sign(types_table$dgp))
```

The posterior means of the five coefficients, including the two
alternative-specific coefficients of `w` and the constant of alternative
`B`, deviate from the true values by at most `r round(types_deviation, 1)`
posterior standard deviations, and
`r if (types_signs) "all five" else "not all"` have the sign of the true
value. `base` selects the alternative against which the alternative-specific
coefficients are measured, by default the first in alphabetical order, here
`A`. Switching the base to `B` changes the parameterization, not the model:

```{r base}
base_b <- update(covariate_types, base = "B")
coef(base_b)[c("beta[x]", "beta[z_A]", "beta[ASC_A]")]
```

```{r base-shift, include=FALSE}
base_shift <- abs(
  coef(base_b)[["beta[x]"]] - coef(covariate_types)[["beta[x]"]]
)
```

The shared coefficient of `x` differs from the previous fit by only
`r sprintf("%.4f", base_shift)`, the Monte Carlo error of two separate
runs, while the coefficient of `z` and the constant now describe alternative
`A` relative to `B` and therefore change sign.

In 1987, 210 travelers between Sydney and Melbourne reported which of four
modes they had taken: air, train, bus, or car. The `TravelMode` data of the
**AER** package [@Kleiber2008] are in long format with one row per mode and
a `"yes"`/`"no"` indicator of the chosen mode, which `fit()` expects as a
logical or `0`/`1` variable. Terminal waiting time (`wait`), in-vehicle
cost (`vcost`), and travel time (`travel`) vary across modes and enter as
type `A`. Household income and the size of the traveling party describe the
traveler and enter as type `B`, with the alternative-specific constants
included by default. Cost and income are converted from Australian dollars
to euro.

```{r travel-formula}
travel_formula <- choice ~ wait + vcost + travel | income + size
```

The first alternative in alphabetical order, `air`, is the base, so the
constants and the coefficients of income and party size describe the other
modes relative to flying. A free error covariance between four alternatives
is weakly identified when every traveler is observed once, and the sampler
mixes slowly for its entries, so the chains run 20000 iterations and retain
every twentieth draw.

```{r travel}
data("TravelMode", package = "AER")
TravelMode$choice <- TravelMode$choice == "yes"
TravelMode$vcost <- TravelMode$vcost / 1.6196
TravelMode$income <- TravelMode$income / 1.6196
travel <- fit(
  formula = travel_formula,
  data = TravelMode,
  format = "long",
  column_decider = "individual",
  column_alternative = "mode",
  iterations = 20000,
  warmup = 10000,
  thin = 20,
  chains = 2,
  progress = FALSE
)
summary(travel)
```

```{r travel-diagnostics, include=FALSE}
travel_table <- summary(travel)$posterior
travel_sigma <- startsWith(travel_table$variable, "Sigma")
travel_rhat <- c(
  beta = max(travel_table$rhat[!travel_sigma]),
  Sigma = max(travel_table$rhat[travel_sigma])
)
travel_slowest <- paste0(
  "`", travel_table$variable[which.min(travel_table$ess_bulk)], "`"
)
travel_cost <- travel_table[travel_table$variable == "beta[vcost]", ]
travel_cost_ratio <- abs(travel_cost$mean) / travel_cost$sd
```

At this length, `rhat` is at most `r sprintf("%.3f", travel_rhat[["beta"]])`
for the coefficients and `r sprintf("%.3f", travel_rhat[["Sigma"]])` for the
covariance entries, and the smallest bulk effective sample size,
`r round(min(travel_table$ess_bulk))`, belongs to `r travel_slowest`. The
fit took `r round(travel$sampler$elapsed[["total"]])` seconds. The three
attribute coefficients are negative: waiting, cost, and travel time reduce
the utility of a mode. The cost coefficient lies only
`r round(travel_cost_ratio, 1)` posterior standard deviations from zero, so
these data contain little information about a value of time, and a
compensation with the cost as reference would have an uninformative
credible interval. The income and party size coefficients are
alternative-specific and have no direct interpretation on the utility scale,
so `interpret(type = "mea")` computes marginal effects, the derivatives of
the choice probabilities with respect to a covariate, for a traveler with
average covariates:

```{r travel-income}
mode_effects <- interpret(travel, type = "mea")
mode_effects[mode_effects$covariate == "income", ]
```

```{r travel-numbers, include=FALSE}
income_effect <- function(mode) {
  row <- mode_effects$covariate == "income" & mode_effects$alternative == mode
  100 * mode_effects$mean[row]
}
direction <- function(x) if (x < 0) "lowers" else "raises"
```

An additional thousand euro of household income
`r direction(income_effect("train"))` the probability of the train by about
`r sprintf("%.1f", abs(income_effect("train")))` percentage points and
`r direction(income_effect("air"))` that of the plane by about
`r sprintf("%.1f", abs(income_effect("air")))` points. The four effects sum
to zero because the four probabilities sum to one.

## Individual choice sets

Not every alternative is available at every occasion: a traveler without a
car cannot drive, and a route without a rail link has no train option.
Unordered choices can therefore be made from occasion-specific subsets of
the alternatives. In long format, an occasion lists only the rows of its
available alternatives, and no further argument is needed. The sampler
imputes the latent utilities of unavailable alternatives without
restriction, so they do not affect the choice, and predictions assign them
probability zero. Ordered and ranked models require complete choice sets.

Between Montreal and Toronto, travelers can fly, drive, or take the train,
but not all modes are available on every trip. The `ModeCanada` data of the
**mlogit** package cover 4324 trips in this corridor [@Bhat1995].

```{r canada-load}
data("ModeCanada", package = "mlogit")
head(ModeCanada)
```

Cost
(`cost`), in-vehicle time (`ivt`), out-of-vehicle time (`ovt`), and service
frequency (`freq`) vary across modes; household income and the number of
urban trip endpoints are trip-specific. The data also contain the bus, which
was chosen on only `r sum(ModeCanada$alt == "bus" & ModeCanada$choice == 1)`
trips, too few to identify its coefficients and error covariances, so the
bus rows and the trips on which it was chosen are removed. Cost and income
are converted from Canadian dollars to euro, and trips with a single
remaining alternative are dropped.

```{r canada-data}
ModeCanada$cost <- ModeCanada$cost / 1.6151
ModeCanada$income <- ModeCanada$income / 1.6151
bus_trips <- ModeCanada$case[ModeCanada$alt == "bus" & ModeCanada$choice == 1]
canada_data <- ModeCanada[
  ModeCanada$alt != "bus" & !(ModeCanada$case %in% bus_trips),
]
set_size <- table(canada_data$case)
canada_data <- canada_data[set_size[as.character(canada_data$case)] > 1, ]
table(table(canada_data$case))
```

The choice sets are read from the rows of each trip. Every trip is one
decider, and the model has the same structure as the travel mode model
above: three attributes of type `A`, two trip characteristics of type `B`,
and the constants, all relative to `air`.

```{r canada}
canada <- fit(
  choice ~ cost + ivt + ovt + freq | income + urban,
  data = canada_data,
  format = "long",
  column_decider = "case",
  column_alternative = "alt",
  iterations = 20000,
  warmup = 10000,
  thin = 10,
  chains = 2,
  progress = FALSE
)
summary(canada)
```

```{r canada-numbers, include=FALSE}
canada_table <- summary(canada)$posterior
canada_slowest <- paste0(
  "`", canada_table$variable[which.min(canada_table$ess_bulk)], "`"
)
time_ratio <- interpret(canada, reference = "ivt", effects = "ovt")
canada_income <- coef(canada)[c("beta[income_car]", "beta[income_train]")]
canada_income_sign <- if (all(canada_income < 0)) {
  "negative"
} else {
  "not both negative"
}
```

The largest `rhat` is `r sprintf("%.3f", max(canada_table$rhat))`, and the
smallest bulk effective sample size, `r round(min(canada_table$ess_bulk))`,
belongs to `r canada_slowest`. A minute out of the vehicle reduces the
utility about `r round(abs(time_ratio$mean), 1)` times as much as a minute
in it, with a 95% credible interval from `r round(abs(time_ratio$upper), 1)`
to `r round(abs(time_ratio$lower), 1)`, and the income coefficients of car
and train are `r canada_income_sign`: a higher income increases the
probability of flying.

## Ordered responses

Some responses are ordered levels, not choices between alternatives: never,
occasionally, regularly, heavily. An ordered model has one latent utility
per occasion and compares it with increasing thresholds `gamma`; the level
is the interval into which the utility falls. `alternatives` gives the
response levels in increasing order. Latent-variable data augmentation
provides a direct Bayesian treatment of ordered probit models
[@Albert1993]. The following demonstration estimates a simulated
three-category model and reports the coefficient and the free threshold
beside their true values.

```{r ordered-sim}
ordered_sim <- fit(
  choice ~ x | 0,
  alternatives = c("low", "middle", "high"),
  choice_type = "ordered",
  n_deciders = 500,
  dgp_parameters = list(beta = c(x = 1), gamma = c(0, 1)),
  chains = 1
)
summary(ordered_sim)
```

```{r ordered-deviation, include=FALSE}
ordered_table <- summary(ordered_sim)$posterior
ordered_deviation <- max(abs(ordered_table$mean - ordered_table$dgp))
```

The posterior means of the coefficient and of the threshold between the
middle and the high category deviate from their true values by at most
`r signif(ordered_deviation, 1)`.

The `survey` data of the **MASS** package [@VenablesRipley2002] come from
237 statistics students at the University of Adelaide who reported how often
they smoke, together with their age and how much they exercise. The smoking
level `Smoke` is stored as a factor whose levels are in alphabetical order;
`alternatives` puts them in their natural order from never to heavy. The
other survey questions are not used; `fit()` ignores their missing values.

```{r ordered}
data("survey", package = "MASS")
levels(survey$Smoke)
smoking_levels <- c("Never", "Occas", "Regul", "Heavy")
smoking <- fit(
  Smoke ~ Age + Exer | 0,
  data = survey,
  alternatives = smoking_levels,
  choice_type = "ordered",
  column_decider = NULL,
  chains = 1
)
summary(smoking)
```

The first threshold is fixed to zero and the error variance to one; the
remaining thresholds `gamma[2]` and `gamma[3]` are estimated. Each student
has one latent utility, normally distributed around its systematic part,
and the thresholds partition it into the four levels. The area under the
density between two thresholds is the probability of that level. The figure
shows the density of a student whose systematic utility is zero.

```{r ordered-figure}
thresholds <- coef(smoking)[c("gamma[2]", "gamma[3]")]
cuts <- c(-Inf, 0, thresholds, Inf)
shades <- grey(seq(0.45, 0.9, length.out = length(smoking_levels)))
utility <- seq(-3.5, 3.5, length.out = 400)
plot(
  utility, dnorm(utility),
  type = "n", axes = FALSE, ylab = "",
  xlab = "latent utility of a student"
)
for (k in seq_along(smoking_levels)) {
  inside <- utility >= cuts[k] & utility <= cuts[k + 1]
  polygon(
    c(max(cuts[k], -3.5), utility[inside], min(cuts[k + 1], 3.5)),
    c(0, dnorm(utility[inside]), 0),
    col = shades[k], border = NA
  )
}
lines(utility, dnorm(utility), lwd = 2)
axis(1, at = c(-3, 0, 3))
legend(
  "topright", legend = smoking_levels, fill = shades, border = NA, bty = "n"
)
```

```{r smoking-numbers, include=FALSE}
smoking_draws <- posterior::as_draws_matrix(smoking)
smoking_negative <- sprintf("%.2f", mean(smoking_draws[, "beta[Age]"] < 0))
smokers <- sum(survey$Smoke != "Never", na.rm = TRUE)
```

The occasional and the regular level together occupy an interval of width
`r round(coef(smoking)[["gamma[3]"]], 2)` to the right of the first
threshold, and the two outer levels the unbounded intervals beyond. A
covariate shifts the density along the utility axis, so one coefficient per
covariate describes its effect on all four levels. The factor `Exer` enters
through its dummy variables relative to the students who exercise
frequently. The posterior probability that the age coefficient is negative
is `r smoking_negative`: older students report smoking less. With
`r smokers` smokers among the `r nobs(smoking)` students who answered the
question, the data contain little information about the exercise
contrasts. What does the age coefficient mean for the probabilities of the
levels? `interpret(type = "ame")` differentiates the probability of each
level with respect to age and averages the derivatives over the students;
the vignette [Posterior prediction][v04] explains marginal effects in more
detail:

```{r ordered-interpret}
age_effects <- interpret(smoking, type = "ame")
age_effects
```

One more year of age raises the probability of never smoking by about
`r round(100 * age_effects$mean[age_effects$alternative == "Never"], 1)`
percentage points and lowers the probabilities of the other three levels.

## Ranked responses

Ranked data record the complete order of the alternatives. In wide format,
the response columns combine the response name with each alternative, for
example `rank_Xbox`. The model is the same probit model as for unordered
choices, but the likelihood uses the full ordering of the utilities. The
following demonstration simulates rankings of three alternatives and
compares the coefficient and the free covariance parameters with the true
values.

```{r ranked-sim}
ranked_sim <- fit(
  rank ~ x | 0,
  choice_type = "ranked",
  n_deciders = 300,
  dgp_parameters = list(
    beta = c(x = 1),
    Sigma = rbind(c(0, 0, 0), c(0, 1, 0.2), c(0, 0.2, 1))
  ),
  chains = 1
)
ranked_summary <- summary(
  ranked_sim, variables = c("beta[x]", "Sigma[C,B]", "Sigma[C,C]")
)
ranked_summary
```

```{r ranked-deviation, include=FALSE}
ranked_table <- ranked_summary$posterior
ranked_deviation <- round(
  abs(ranked_table$mean - ranked_table$dgp) / ranked_table$sd, 1
)
```

A full ranking contains more information than a single choice. The
posterior means of the coefficient, the covariance, and the variance deviate
from their true values by `r ranked_deviation[1]`, `r ranked_deviation[2]`,
and `r ranked_deviation[3]` posterior standard deviations.

The `Game` data of the **mlogit** package contain complete rankings of six
gaming platforms by 91 Dutch respondents, together with whether they own
each platform (`own`), their age, and their weekly gaming hours; the source
study develops a rank-ordered choice model for these data [@Fok2012]. The
ranks are stored in the columns `ch.Xbox`, `ch.PlayStation`, and so on, so
the response in the formula is `ch` and `delimiter = "."` separates it from
the alternative.

```{r ranked}
data("Game", package = "mlogit")
gaming <- fit(
  ch ~ own | age + hours,
  data = Game,
  alternatives = c(
    "Xbox", "PlayStation", "PSPortable", "GameCube", "GameBoy", "PC"
  ),
  choice_type = "ranked",
  delimiter = ".",
  column_decider = NULL,
  iterations = 2000,
  warmup = 1000,
  chains = 2,
  progress = FALSE
)
coef(gaming)[1:6]
```

The coefficient of `own` is positive: owning a platform raises its rank. The
alternative-specific constants and the coefficients of `age` and `hours` are
relative to the base alternative `GameBoy`, the first platform in
alphabetical order, regardless of the order in `alternatives`. Which
platform gains from additional gaming hours? For a ranked model, the
marginal effects of `interpret()` refer to the probability of being ranked
first, here for a respondent of average age who plays the average number of
hours:

```{r ranked-interpret}
platform_effects <- interpret(gaming, type = "mea")
platform_effects[platform_effects$covariate == "hours", ]
```

```{r ranked-numbers, include=FALSE}
pc_hours <- platform_effects$mean[
  platform_effects$covariate == "hours" & platform_effects$alternative == "PC"
]
```

Heavy gamers prefer the PC: every additional weekly hour raises the
probability of ranking it first by about `r round(100 * pc_hours, 1)`
percentage points and lowers the probabilities of the other platforms by the
same amount in total.

## Further reading

The vignette [Modeling preference heterogeneity][v03] covers coefficients
that differ between deciders. The vignette [Posterior prediction][v04]
computes predictions and marginal effects from a fitted model, and the
vignette [Bayesian model evaluation][v05] compares competing specifications.

[v01]: https://loelschlaeger.de/RprobitB/articles/v01_get_started.html
[v03]: https://loelschlaeger.de/RprobitB/articles/v03_heterogeneity.html
[v04]: https://loelschlaeger.de/RprobitB/articles/v04_prediction.html
[v05]: https://loelschlaeger.de/RprobitB/articles/v05_model_evaluation.html

## References
