---
title: "Model specification and variants"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model specification and variants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
link-citations: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
  fig.align = "center"
)
```

This vignette describes how to specify a probit choice model with `fit()`.
It covers the normalization of the utility scale and the prior distribution,
which every fit involves, and then the model variants: the three types of
covariates and the alternative-specific constants, choice sets that differ
between occasions, and ordered and ranked responses. Preference
heterogeneity between deciders is the subject of the vignette
[Modeling preference heterogeneity][v03], and @Oelschlaeger2026c gives the
methodological background. From the covariate types onward, each
section first fits the variant to simulated data and then to a data set of
the **AER** package [@Kleiber2008], the **MASS** package
[@VenablesRipley2002], or the **mlogit** package [@Croissant2020]. Without
data, `fit()` simulates the requested model before estimating it, and
`summary()` prints the true values in a `dgp` column beside the posterior
summaries, converted to the normalization of the fit.

```{r setup}
library(RprobitB)
set.seed(1)
```

## Normalization

The probit model introduced in the vignette
[Get started with RprobitB][v01] assigns every alternative $j$ at occasion
$t$ of decider $n$ the latent utility
$U_{ntj} = X_{ntj}^\top \beta_n + \epsilon_{ntj}$ with jointly normal errors
$\epsilon_{nt} \sim \mathrm{N}(0, \Sigma)$, and the decider chooses the
alternative with the largest utility. The choice probability of an
alternative is the probability that its utility exceeds the utilities of all
others. These probabilities are invariant to adding a constant to all
utilities and to multiplying all utilities by a positive number, so neither
the level nor the scale of the utilities is identified. **RprobitB**
therefore works with utility differences relative to a base alternative, by
default the first in alphabetical order, and fixes one parameter through
`scale`:

- `scale = NULL` fixes the error variance of the first utility difference to
  one. This is the default. The covariance matrix of the error differences is
  reported as `Sigma[<alternative>,<alternative>]`, with rows and columns
  named by the alternatives other than the base. For three alternatives `A`,
  `B`, and `C` with base `A`, its entries are `Sigma[B,B]`, `Sigma[C,B]`, and
  `Sigma[C,C]`.
- `scale = c(z = -1)` fixes the coefficient of `z` to `-1` instead, so that
  all other coefficients are measured in units of it. With a price
  coefficient fixed in this way, the other coefficients are willingness-to-pay
  values.

The sampler draws all parameters without restriction and rescales every
retained draw afterwards. Variables that the normalization fixes remain in
the draws but are omitted from `summary()`, `coef()`, and `vcov()`.

The following demonstration shows the effect of the normalization on the
reported values. We simulate data with coefficients `1` for `x` and `-0.5`
for `z` and fit them under the default scale. Simulated alternatives are
labeled with capital letters unless `alternatives` names them, here `A` and
`B`, and `A` is the base. `update()` then refits the same simulated data
with the coefficient of `z` fixed to `-1`, so only the normalization differs
between the two fits.

```{r normalization}
scale_default <- fit(
  choice ~ x + z | 0,
  dgp_parameters = list(beta = c(x = 1, z = -0.5)),
  n_deciders = 300,
  chains = 1
)
summary(scale_default)
scale_z <- update(scale_default, scale = c(z = -1))
summary(scale_z)
```

Under the default scale, the `dgp` column shows the coefficients as
specified. `Sigma[B,B]`, the error variance of the utility difference
between alternative `B` and the base `A`, is fixed to one and therefore not
listed. Under the coefficient normalization, `beta[z]` is fixed to `-1`
instead of its true `-0.5`, so all utilities are multiplied by `2`: the true
`beta[x]` becomes `2` and the true error variance `4`. `summary()` converts
the true values to the normalization of the fit, so the `dgp` column remains
comparable.

The convergence diagnostics of the two tables differ because `update()`
runs the sampler again and because the rescaled variables are different
functions of the draws: `beta[x]` is now a ratio of two coefficients, which
mixes differently than a single coefficient.

## Prior distribution

**RprobitB** estimates every model with a Gibbs sampler, which draws each
block of parameters in turn from its conditional posterior distribution given
the other parameters and the latent utilities. The prior is conjugate for
every block, so each conditional posterior belongs to the same family as the
prior. Fixed coefficients and class means have normal priors, covariance
matrices have inverse Wishart priors, class weights have a Dirichlet prior,
and the log-increments between ordered thresholds have a normal prior. The
section "Prior distribution" of `?fit` lists all components with their
defaults, which are weakly informative for coefficients and covariances.
Single components are overridden through a named list, and the complete
prior of a fit is stored in its `prior` component.

The following demonstration simulates 100 deciders with a true coefficient
of `-1` under the default prior and then refits the same data under two
priors with mean `1`: a moderate one with variance `0.5` and a tight one
with variance `0.01`.

```{r prior}
default_prior <- fit(
  choice ~ x | 0,
  dgp_parameters = list(beta = c(x = -1)),
  chains = 1
)
default_prior$prior
summary(default_prior)
moderate_prior <- update(
  default_prior, prior = list(fixed_mean = 1, fixed_covariance = matrix(0.5))
)
tight_prior <- update(
  default_prior, prior = list(fixed_mean = 1, fixed_covariance = matrix(0.01))
)
data.frame(
  variable = "beta[x]", dgp = -1, default = coef(default_prior),
  moderate = coef(moderate_prior), tight = coef(tight_prior),
  row.names = NULL
)
```

The table compares the posterior means under the three priors with the true
value. The moderate prior shifts the posterior mean only slightly towards the
prior mean. The tight prior has a standard deviation of `0.1` around `1`,
outweighs the data, and pulls the posterior mean far away from the true
value. A prior on a coefficient is unproblematic as long as its variance
reflects the actual uncertainty about the coefficient.

## Covariate types and alternative-specific constants

The formula `choice ~ A | B | C` distinguishes three kinds of covariates:
attributes of the alternatives with one shared coefficient (`A`),
characteristics of the decider with alternative-specific coefficients (`B`),
and attributes of the alternatives with alternative-specific coefficients
(`C`). Alternative-specific constants are included by default and removed with 
`0` in the second part. The following demonstration simulates all three types at 
once, and `summary()` shows the posterior summaries beside the true values.

```{r covariate-types-sim}
covariate_types <- fit(
  choice ~ x | z | w,
  n_deciders = 500,
  dgp_parameters = list(beta = c(
    x = 0.5, z_B = -0.5, ASC_B = 0.25, w_A = -0.5, w_B = 0.5
  )),
  chains = 1
)
summary(covariate_types)
```

The summary lists the shared coefficient of `x`, the coefficient of `z` and
the constant of alternative `B`, and the two alternative-specific
coefficients of `w`, each beside its true value. `base` selects the
alternative against which the alternative-specific coefficients are
measured, by default the first in alphabetical order, here `A`. Switching
the base to `B` changes only the parameterization, not the model:

```{r base}
base_b <- update(covariate_types, base = "B")
coef(base_b)[c("beta[x]", "beta[z_A]", "beta[ASC_A]")]
```

The shared coefficient of `x` is unchanged up to Monte Carlo error, while
the coefficient of `z` and the constant now describe alternative `A`
relative to `B` and therefore change sign.

In 1987, 210 travelers between Sydney and Melbourne reported which of four
modes they had taken: air, train, bus, or car. The `TravelMode` data of the
**AER** package [@Kleiber2008] are in long format with one row per mode and
a `"yes"`/`"no"` indicator of the chosen mode, which `fit()` expects as a
logical or `0`/`1` variable. Terminal waiting time (`wait`), in-vehicle
cost (`vcost`), and travel time (`travel`) vary across modes and enter as
type `A`. Household income and the size of the traveling party describe the
traveler and enter as type `B`, with the alternative-specific constants
included by default. Cost and income are converted from Australian dollars
to euro.

```{r travel-formula}
travel_formula <- choice ~ wait + vcost + travel | income + size
```

The first alternative in alphabetical order, `air`, is the base, so the
constants and the coefficients of income and party size describe the other
modes relative to flying.

```{r travel}
data("TravelMode", package = "AER")
TravelMode$choice <- TravelMode$choice == "yes"
TravelMode$vcost <- TravelMode$vcost / 1.6196
TravelMode$income <- TravelMode$income / 1.6196
travel <- fit(
  formula = travel_formula,
  data = TravelMode,
  format = "long",
  column_decider = "individual",
  column_alternative = "mode",
  iterations = 6000,
  warmup = 3000,
  thin = 30,
  chains = 2,
  progress = FALSE
)
summary(travel)
```

The three attribute coefficients are negative: waiting, cost, and travel
time reduce the utility of a mode. The income and party size coefficients
are alternative-specific and have no direct interpretation on the utility
scale, so `interpret(type = "mea")` computes marginal effects, the
derivatives of the choice probabilities with respect to a covariate, for a
traveler with average covariates:

```{r travel-income}
mode_effects <- interpret(travel, type = "mea")
mode_effects[mode_effects$covariate == "income", ]
```

A higher household income lowers the probability of the train and raises
the probabilities of the plane and the car.

## Individual choice sets

Not every alternative is available at every occasion: a traveler without a
car cannot drive, and a route without a rail link has no train option.
Unordered choices can therefore be made from occasion-specific subsets of
the alternatives. In long format, an occasion lists only the rows of its
available alternatives, and no further argument is needed. The sampler
imputes the latent utilities of unavailable alternatives without
restriction, so they do not affect the choice, and predictions assign them
probability zero.

Between Montreal and Toronto, travelers can fly, drive, or take the train,
but not all modes are available on every trip. The `ModeCanada` data of the
**mlogit** package cover 4324 trips in this corridor [@Bhat1995].

```{r canada-load}
data("ModeCanada", package = "mlogit")
head(ModeCanada)
```

Cost (`cost`), in-vehicle time (`ivt`), out-of-vehicle time (`ovt`), and
service frequency (`freq`) vary across modes; household income and the
number of urban trip endpoints are trip-specific. The data also contain the
bus, which
was chosen on only `r sum(ModeCanada$alt == "bus" & ModeCanada$choice == 1)`
trips, too few to identify its coefficients and error covariances, so the
bus rows and the trips on which it was chosen are removed. Cost and income
are converted from Canadian dollars to euro, and trips with a single
remaining alternative are dropped. The first 1000 of the remaining trips
keep the computation short.

```{r canada-data}
ModeCanada$cost <- ModeCanada$cost / 1.6151
ModeCanada$income <- ModeCanada$income / 1.6151
bus_trips <- ModeCanada$case[ModeCanada$alt == "bus" & ModeCanada$choice == 1]
canada_data <- ModeCanada[
  ModeCanada$alt != "bus" & !(ModeCanada$case %in% bus_trips),
]
set_size <- table(canada_data$case)
canada_data <- canada_data[set_size[as.character(canada_data$case)] > 1, ]
canada_data <- canada_data[
  canada_data$case %in% unique(canada_data$case)[1:1000],
]
table(table(canada_data$case))
```

The choice sets are read from the rows of each trip. Every trip is one
decider, and the model has the same structure as the travel mode model
above: three attributes of type `A`, two trip characteristics of type `B`,
and the constants, all relative to `air`.

```{r canada}
canada <- fit(
  choice ~ cost + ivt + ovt + freq | income + urban,
  data = canada_data,
  format = "long",
  column_decider = "case",
  column_alternative = "alt",
  iterations = 6000,
  warmup = 3000,
  thin = 15,
  chains = 2,
  progress = FALSE
)
summary(canada)
```

The income coefficients of car and train are negative: a higher income
increases the probability of flying. 

## Ordered responses

Some responses are ordered levels. An ordered model has one latent utility
per occasion and compares it with increasing thresholds `gamma`; the level
is the interval into which the utility falls. `alternatives` gives the
response levels in increasing order. The utility includes an intercept
`ASC` by default, as unordered models include alternative-specific
constants, and `| 0` removes it. Latent-variable data augmentation
provides a direct Bayesian treatment of ordered probit models
[@Albert1993]. The following demonstration estimates a simulated
three-category model and reports the intercept, the coefficient, and the
free threshold beside their true values.

```{r ordered-sim}
ordered_sim <- fit(
  choice ~ x,
  alternatives = c("low", "middle", "high"),
  choice_type = "ordered",
  n_deciders = 500,
  dgp_parameters = list(beta = c(ASC = 0.5, x = 1), gamma = c(0, 1)),
  chains = 1
)
summary(ordered_sim)
```

The summary lists the intercept, the coefficient, and the free threshold
`gamma[2]` beside their true values. The first threshold is fixed to zero
and the error variance to one, which identifies the scale of the utility
and the level of the thresholds.

The `survey` data of the **MASS** package [@VenablesRipley2002] come from
237 statistics students at the University of Adelaide who reported how often
they smoke, together with their age and how much they exercise. The smoking
level `Smoke` is stored as a factor whose levels are in alphabetical order;
`alternatives` puts them in their natural order from never to heavy. The
other survey questions are not used; `fit()` ignores their missing values.

```{r ordered}
data("survey", package = "MASS")
levels(survey$Smoke)
smoking_levels <- c("Never", "Occas", "Regul", "Heavy")
smoking <- fit(
  Smoke ~ Age + Exer,
  data = survey,
  alternatives = smoking_levels,
  choice_type = "ordered",
  column_decider = NULL,
  chains = 1
)
summary(smoking)
```

With four levels, the thresholds `gamma[2]` and `gamma[3]` are estimated.
Each student has one latent utility, normally distributed around its
systematic part,
and the thresholds partition it into the four levels. The area under the
density between two thresholds is the probability of that level. The figure
shows the density of a student whose systematic utility is zero.

```{r ordered-figure}
thresholds <- coef(smoking)[c("gamma[2]", "gamma[3]")]
cuts <- c(-Inf, 0, thresholds, Inf)
shades <- grey(seq(0.45, 0.9, length.out = length(smoking_levels)))
utility <- seq(-3.5, 3.5, length.out = 400)
plot(
  utility, dnorm(utility),
  type = "n", axes = FALSE, ylab = "",
  xlab = "latent utility of a student"
)
for (k in seq_along(smoking_levels)) {
  inside <- utility >= cuts[k] & utility <= cuts[k + 1]
  polygon(
    c(max(cuts[k], -3.5), utility[inside], min(cuts[k + 1], 3.5)),
    c(0, dnorm(utility[inside]), 0),
    col = shades[k], border = NA
  )
}
lines(utility, dnorm(utility), lwd = 2)
axis(1, at = c(-3, 0, 3))
legend(
  "topright", legend = smoking_levels, fill = shades, border = NA, bty = "n"
)
```

A covariate shifts the density along the utility axis, so one coefficient
per covariate describes its effect on all four levels. The age coefficient
is small and its credible interval includes zero: the age of a student does
not affect how much they smoke. The factor `Exer` enters through its dummy
variables relative to the students who exercise frequently; both
coefficients are negative but uncertain. `interpret(type = "ame")`
translates the age coefficient into probabilities: it differentiates the
probability of each level with respect to age and averages the derivatives
over the students. The vignette [Posterior prediction][v04] explains
marginal effects in more detail.

```{r ordered-interpret}
age_effects <- interpret(smoking, type = "ame")
age_effects
```

One more year of age changes the probability of never smoking by less than
one percentage point, and the credible intervals of all four effects
include zero.

## Ranked responses

Ranked data record the complete order of the alternatives. In wide format,
the response columns combine the response name with each alternative, for
example `rank_Xbox`. The model is the same probit model as for unordered
choices, but the likelihood uses the full ordering of the utilities. The
following demonstration simulates rankings of three alternatives and
compares the coefficient and the free covariance parameters with the true
values.

```{r ranked-sim}
ranked_sim <- fit(
  rank ~ x | 0,
  choice_type = "ranked",
  n_deciders = 300,
  dgp_parameters = list(
    beta = c(x = 1),
    Sigma = rbind(c(0, 0, 0), c(0, 1, 0.2), c(0, 0.2, 1))
  ),
  chains = 1
)
summary(ranked_sim, variables = c("beta[x]", "Sigma[C,B]", "Sigma[C,C]"))
```

The `Game` data of the **mlogit** package contain complete rankings of six
gaming platforms by 91 Dutch respondents, together with whether they own
each platform (`own`), their age, and their weekly gaming hours; the source
study develops a rank-ordered choice model for these data [@Fok2012]. The
ranks are stored in the columns `ch.Xbox`, `ch.PlayStation`, and so on, so
the response in the formula is `ch` and `delimiter = "."` separates it from
the alternative.

```{r ranked}
data("Game", package = "mlogit")
gaming <- fit(
  ch ~ own | age + hours,
  data = Game,
  alternatives = c(
    "Xbox", "PlayStation", "PSPortable", "GameCube", "GameBoy", "PC"
  ),
  choice_type = "ranked",
  delimiter = ".",
  column_decider = NULL,
  iterations = 1000,
  warmup = 500,
  thin = 20,
  chains = 2,
  progress = FALSE
)
coef(gaming)[1:6]
```

The coefficient of `own` is positive: owning a platform raises its rank. The
alternative-specific constants and the coefficients of `age` and `hours` are
relative to the base alternative `GameBoy`. Which
platform gains from additional gaming hours? For a ranked model, the
marginal effects of `interpret()` refer to the probability of being ranked
first, here for a respondent of average age who plays the average number of
hours:

```{r ranked-interpret}
platform_effects <- interpret(gaming, type = "mea")
platform_effects[platform_effects$covariate == "hours", ]
```

```{r ranked-numbers, include=FALSE}
pc_hours <- platform_effects$mean[
  platform_effects$covariate == "hours" & platform_effects$alternative == "PC"
]
```

Heavy gamers prefer the PC: every additional weekly hour raises the
probability of ranking it first by about `r round(100 * pc_hours, 1)`
percentage points.

## Further reading

The vignette [Modeling preference heterogeneity][v03] covers coefficients
that differ between deciders. The vignette [Posterior prediction][v04]
computes predictions and marginal effects from a fitted model, and the
vignette [Bayesian model evaluation][v05] compares competing specifications.

[v01]: https://loelschlaeger.de/RprobitB/articles/v01_get_started.html
[v03]: https://loelschlaeger.de/RprobitB/articles/v03_heterogeneity.html
[v04]: https://loelschlaeger.de/RprobitB/articles/v04_prediction.html
[v05]: https://loelschlaeger.de/RprobitB/articles/v05_model_evaluation.html

## References
