---
title: "Model specification and variants"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model specification and variants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
link-citations: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
  fig.align = "center"
)
```

Every model variant of **RprobitB** [@Oelschlaeger2026d] is requested through
arguments of `fit()`: which covariates get which kind of coefficient, whether
the choice sets differ between occasions, and whether the response is a
choice, an ordered category, or a ranking. This vignette starts with the two
things every probit fit rests on, the normalization and the prior, and then
goes through the variants one by one. Letting preferences differ between
deciders has a vignette of its own, and @Oelschlaeger2026c lays out the
methodological background.

From the covariate types onward, each section first fits the variant to
simulated data and then to a data set of the **mlogit** package
[@Croissant2020] or the **MASS** package [@VenablesRipley2002]. With
`data = NULL`, `fit()` simulates the requested model before estimating it,
and `summary()` prints the true values in a `dgp` column beside the posterior
summaries, converted to the normalization of the fit; where the data were
simulated by hand, a small table does the same. The simulated fits use one
chain, the empirical ones two, so that their diagnostics can be read.

```{r setup}
library(RprobitB)
set.seed(1)
```

## Normalization

Utilities are identified only up to level and scale: adding a constant to
all of them or multiplying them by a positive number changes nothing about
which alternative wins. **RprobitB** therefore works with differences relative
to a base alternative, chosen by `base` and defaulting to the first
alternative in alphabetical order, and fixes one parameter through `scale`.
How this is done is a central design question of Bayesian multinomial probit
models [@McCulloch1994; @Imai2005a]:

- `scale = NULL` fixes the error variance of the first utility difference to
  one. This is the default and yields the posterior variable names
  `Sigma[B,B]`, `Sigma[C,B]`, and so on.
- `scale = c(price = -1)` fixes a coefficient instead. All other coefficients
  are then measured in units of that coefficient, which is convenient when
  willingness-to-pay measures are the goal.
- `scale = c("Sigma_B,B" = 1)` names the fixed variance explicitly.

The sampler draws all parameters without restriction and rescales every
retained draw afterwards. Variables that the normalization fixes stay in the
draws but are left out of `summary()`, `coef()`, and `vcov()`.

The proof of concept shows what the normalization does to the numbers. Data
generated with coefficients `1` for `x` and `-0.5` for `z` are fitted once
under the default scale and once with the coefficient of `z` fixed to `-1`.
Seeding both calls identically makes `fit()` simulate the same data, so only
the normalization differs.

```{r normalization}
set.seed(1)
scale_default <- fit(
  choice ~ x + z | 0,
  dgp_parameters = list(beta = c(x = 1, z = -0.5)),
  n_deciders = 300,
  chains = 1
)
summary(scale_default)
set.seed(1)
scale_z <- fit(
  choice ~ x + z | 0,
  dgp_parameters = list(beta = c(x = 1, z = -0.5)),
  scale = c(z = -1),
  n_deciders = 300,
  chains = 1
)
summary(scale_z)
```

Under the default scale, the `dgp` column shows the coefficients as
specified, and `Sigma[B,B]` is fixed to one and therefore not listed. Under
the coefficient normalization, `beta[z]` is the fixed variable, and
everything else is expressed in units of it: the true `beta[x]` becomes `2`,
the true error variance `4`, and the estimates follow within their posterior
uncertainty. `summary()` always converts the truth to the normalization of
the fit, so the columns are comparable in both cases.

## Prior distribution

The prior is conjugate for every Gibbs update; the section "Prior
distribution" of `?fit` lists all of its components. Fixed coefficients and
class means have normal priors, covariance matrices have inverse Wishart
priors, class weights have a Dirichlet prior, and the log-increments between
ordered thresholds have a normal prior. The defaults for coefficients and
covariances are weakly informative. The defaults for the mixture
concentrations are a different matter: they encode assumptions about how
many classes are occupied and deserve a sensitivity analysis, as discussed
below. Single components are overridden through a named list, and the
complete prior of a fit is stored in its `prior` component. The
latent-utility data augmentation behind the sampler goes back to
@Albert1993; @Imai2005a developed its marginal augmentation for the
multinomial probit model.

The proof of concept simulates 100 deciders with a true coefficient of `-1`
under the default prior and then refits the same data under two priors that
expect the coefficient near `1`: a moderate one with variance `0.5` and a
tight one with variance `0.01`. `update()` takes care of the refits: it
reuses the simulated data of the first fit and replaces only the arguments
that are named.

```{r prior}
default_prior <- fit(
  choice ~ x | 0,
  dgp_parameters = list(beta = c(x = -1)),
  chains = 1
)
default_prior$prior
summary(default_prior)
moderate_prior <- update(
  default_prior, prior = list(fixed_mean = 1, fixed_covariance = matrix(0.5))
)
tight_prior <- update(
  default_prior, prior = list(fixed_mean = 1, fixed_covariance = matrix(0.01))
)
data.frame(
  variable = "beta[x]", dgp = -1, default = coef(default_prior),
  moderate = coef(moderate_prior), tight = coef(tight_prior),
  row.names = NULL
)
```

The default prior recovers the truth within the posterior uncertainty. The
moderate prior pulls the estimate about
`r round(abs(coef(moderate_prior) - coef(default_prior)), 2)` towards its
mean, a visible
but harmless nudge that a few hundred more observations would erase. The
tight prior is a different story: it insists on a value near `1` with a
standard deviation of `0.1`, and it drags the estimate all the way to zero.
Priors on coefficients are harmless as long as they are honest about their
uncertainty. The priors on mixture concentrations further below deserve
attention even at their defaults.

## Covariate types and alternative-specific constants

The formula `choice ~ A | B | C` distinguishes three kinds of covariates:
attributes of the alternatives with one shared coefficient (`A`),
characteristics of the decider with alternative-specific coefficients (`B`),
and attributes of the alternatives with alternative-specific coefficients
(`C`). Alternative-specific constants are included by default and removed
with `0` in the second part.

The proof of concept simulates all three types at once. Its summary lines up
the posterior estimates with the normalized true values.

```{r covariate-types-recovery}
covariate_types <- fit(
  choice ~ x | z | w,
  n_deciders = 500,
  dgp_parameters = list(beta = c(
    x = 0.5, z_B = -0.5, ASC_B = 0.25, w_A = -0.5, w_B = 0.5
  )),
  chains = 1
)
summary(covariate_types)
```

All five coefficients are recovered with the right signs and sizes,
including the two alternative-specific effects of `w` and the constant of
alternative `B`. The remaining deviations amount to at most two and a half
posterior standard deviations, which is the sampling variation to expect
from one simulated data set.

`base` chooses which alternative the alternative-specific coefficients are
measured against. The default is the first in alphabetical order, here `A`;
switching to `B` mirrors the constant and the decider effect without
changing the model.

```{r base}
base_b <- update(covariate_types, base = "B")
coef(base_b)[c("beta[x]", "beta[z_A]", "beta[ASC_A]")]
```

The shared coefficient of `x` is untouched, while `z` and the constant now
describe alternative `A` relative to `B` and therefore change sign.

Now for a real choice with several alternatives. Where do you go fishing:
the beach, a pier, a private boat, or a charter boat? The `Fishing` data of
the **mlogit** package [@Croissant2020] contain this decision for 1182 US
respondents together with the cost of a trip and the expected catch rate of
each mode, and their monthly income [@Herriges1999]. Cost and catch rate
vary across modes and enter as type `A`; income is a property of the
respondent and enters as type `B`. The columns are named `price.beach`,
`catch.pier`, and so on, which `delimiter = "."` tells `fit()`. Every row is
one respondent, and `column_decider = NULL` numbers them. The first
alternative in alphabetical order, `beach`, is the base, so the constants
and the income coefficients describe the other modes relative to fishing
from the beach.

```{r fishing}
data("Fishing", package = "mlogit")
fishing <- fit(
  mode ~ price + catch | income,
  data = Fishing,
  delimiter = ".",
  column_decider = NULL,
  iterations = 15000,
  warmup = 7500,
  thin = 5,
  chains = 2,
  progress = FALSE
)
summary(fishing)
```

Why fifteen thousand iterations for a model with eight coefficients? Because
of the five entries of `Sigma` below them. A free error covariance between
four alternatives is weakly identified when every respondent is observed
only once, and such parameters move slowly through the posterior. At this
length the coefficients are comfortable and the covariance entries are
usable; the default thousand iterations left `rhat` near 1.5 for them. The
fit takes a few seconds, which is what a cross-section of this size costs.

Higher prices make a mode less and higher catch rates make it more
attractive, the two signs that @Herriges1999 also report; they fitted nested
logit models to these data, so the error structure differs while the story
does not. How much is a better catch worth? `interpret()` expresses the
catch coefficient in units of the price, so the number is the price increase
that just offsets a one unit higher catch rate:

```{r fishing-interpret}
catch_value <- interpret(fishing, reference = "price", effects = "catch")
catch_value
```

Anglers would accept a trip that costs about `r round(catch_value$mean)`
dollars more for a catch rate that is one unit higher, and the interval
comes from the posterior of the ratio, not from the ratio of two posterior
means.

And what does income do? Its three coefficients are alternative-specific
and hard to read on the utility scale, so `interpret(type = "mea")`
translates them into probabilities at the average angler:

```{r fishing-income}
income_effects <- interpret(fishing, type = "mea")
income_effects[income_effects$covariate == "income", ]
```

```{r fishing-numbers, include=FALSE}
income_row <- function(mode) {
  row <- income_effects$covariate == "income" &
    income_effects$alternative == mode
  100 * 1000 * income_effects$mean[row]
}
```

Income is recorded in dollars, so the effects are per dollar and look tiny.
Per thousand dollars of monthly income, anglers move towards the private
boat by about `r round(income_row("boat"), 1)` percentage points and away
from the charter boat by about `r round(abs(income_row("charter")), 1)`
points. The four effects sum to zero, because an angler has to fish
somewhere.

## Heterogeneous preferences

Coefficients need not be the same for every decider. `random_effects` lets
them vary continuously over the population, and `classes` sorts deciders
into groups with their own coefficients. Both are covered in
`vignette("v03_heterogeneity")`, together with the six available mixing
distributions and the four ways of updating latent classes.

## Individual choice sets

Not every alternative is on the menu at every occasion: a traveler without a
car cannot drive, and a route without a rail link has no train option.
Unordered choices may therefore be made from occasion-specific subsets of
the alternatives. In long format, an occasion simply lists the rows of its
available alternatives. The sampler imputes the latent utilities of
unavailable alternatives without any restriction, so they do not affect the
choice, and predictions assign them probability zero.

`fit()` simulates complete choice sets only, so the proof of concept
simulates its data by hand, reproducibly through the `set.seed()` call at
the top. Alternatives A and B are always available and C is available to
every second decider. Utilities use a coefficient of one; with A as the
base, the two estimable error differences have unit variances and zero
covariance. Availability is fixed before the choices are simulated.

```{r choice-sets-recovery}
choice_set_data <- expand.grid(
  deciderID = 1:500, alternative = LETTERS[1:3]
)
choice_set_data <- subset(
  choice_set_data, alternative != "C" | deciderID %% 2 == 0
)
choice_set_data$x <- rnorm(nrow(choice_set_data))
utility <- choice_set_data$x + ifelse(
  choice_set_data$alternative == "A", 0, rnorm(nrow(choice_set_data))
)
choice_set_data$choice <- ave(
  utility, choice_set_data$deciderID,
  FUN = function(x) as.integer(x == max(x))
)
choice_sets <- fit(
  choice ~ x | 0,
  data = choice_set_data,
  format = "long",
  column_alternative = "alternative",
  chains = 1
)
choice_set_truth <- c("beta[x]" = 1, "Sigma[C,B]" = 0, "Sigma[C,C]" = 1)
choice_set_summary <- summary(
  choice_sets, variables = names(choice_set_truth)
)$posterior
data.frame(
  variable = choice_set_summary$variable,
  dgp = unname(choice_set_truth[choice_set_summary$variable]),
  mean = choice_set_summary$mean,
  mode = choice_set_summary$mode,
  sd = choice_set_summary$sd
)
```

The coefficient is recovered closely, and the free error parameters within
their posterior uncertainty, from choice sets of mixed size.

Between Montreal and Toronto, travelers can fly, take the bus, drive, or
take the train, but not every trip has all four on offer. The `ModeCanada`
data of the **mlogit** package [@Croissant2020] cover 4324 trips in this
corridor with two to four available modes each [@Bhat1995]. Cost (`cost`),
in-vehicle time (`ivt`), out-of-vehicle time (`ovt`), and service frequency
(`freq`) vary across modes; household income and the number of urban trip
endpoints are trip-specific. The data are in long format with one row per
available mode, and the cost and the income are converted from Canadian
dollars to euro.

```{r canada}
data("ModeCanada", package = "mlogit")
ModeCanada$cost <- ModeCanada$cost / 1.6151
ModeCanada$income <- ModeCanada$income / 1.6151
table(table(ModeCanada$case))
canada <- fit(
  choice ~ cost + ivt + ovt + freq | income + urban,
  data = ModeCanada,
  format = "long",
  column_decider = "case",
  column_alternative = "alt",
  chains = 1
)
coef(canada)[1:4]
```

The choice sets are read from the rows of each trip; no further argument is
needed. The estimates carry a well-known lesson of transport research, and
the one @Bhat1995 drew from these very data: a minute spent waiting or
walking weighs several times as much as a minute spent in the vehicle.
`interpret()` computes this ratio with its posterior uncertainty, and with
the cost as reference it turns both time coefficients into values of travel
time.

```{r canada-interpret}
time_ratio <- interpret(canada, reference = "ivt", effects = "ovt")
time_ratio
time_values <- interpret(canada, reference = "cost", effects = c("ivt", "ovt"))
time_values
```

A minute out of the vehicle counts about `r round(abs(time_ratio$mean), 1)`
times as much as a minute in it, with a credible interval from
`r round(abs(time_ratio$upper), 1)` to `r round(abs(time_ratio$lower), 1)`.
That is the one number in this vignette that can be held against the
literature rather than only against a sign: mode choice studies of this
corridor and of others place the ratio at roughly two to three, and the
interval covers that range. In money, one more minute in the vehicle must be
offset by a fare that is `r round(abs(time_values$mean[1]), 2)` euro lower,
an hour therefore by about `r round(60 * abs(time_values$mean[1]))` euro.

The two traveler characteristics enter with one coefficient per mode,
relative to the air travelers who form the base:

```{r canada-decider}
coef(canada)[grepl("income|urban", names(coef(canada)))]
```

Income argues for flying: bus, car, and train all carry a negative income
coefficient relative to the plane. Urban trip endpoints work differently.
They speak for the bus and the train, whose stations sit in the city
centers, and against the car, which is where the parking problem is.

Ordered and ranked models require complete choice sets.

## Ordered responses

Some responses are levels rather than picks: never, occasionally, regularly,
heavily. Ordered models estimate one utility per occasion and compare it
with increasing thresholds `gamma`; the response levels are given through
`alternatives` in increasing order. Latent-variable data augmentation
provides a direct Bayesian treatment of ordered probit models
[@Albert1993]. The proof of concept estimates a simulated
three-category model and reports the coefficient and the free threshold
beside their true values.

```{r ordered-recovery}
ordered_recovery <- fit(
  choice ~ x | 0,
  alternatives = c("low", "middle", "high"),
  choice_type = "ordered",
  n_deciders = 500,
  dgp_parameters = list(beta = c(x = 1), gamma = c(0, 1)),
  chains = 1
)
summary(ordered_recovery)
```

Both the coefficient and the threshold between the middle and the high
category are recovered within a tenth of their true values.

The `survey` data of the **MASS** package [@VenablesRipley2002] come from
237 statistics students at the University of Adelaide who reported how often
they smoke, together with their age and how much they exercise. The survey
is a teaching data set rather than a choice experiment, so there is no
earlier choice model to compare the estimates with. The smoking level
`Smoke` is stored as a factor whose levels are in alphabetical order;
`alternatives` puts them in their natural order from never to heavy. The
other survey questions are not used and may contain missing values, which
`fit()` ignores.

```{r ordered}
data("survey", package = "MASS")
levels(survey$Smoke)
smoking_levels <- c("Never", "Occas", "Regul", "Heavy")
smoking <- fit(
  Smoke ~ Age + Exer | 0,
  data = survey,
  alternatives = smoking_levels,
  choice_type = "ordered",
  column_decider = NULL,
  chains = 1
)
summary(smoking)
```

The first threshold is fixed to zero and the error variance to one; the
remaining thresholds `gamma[2]` and `gamma[3]` are estimated. The picture
behind these numbers: every student has one latent utility, normally
distributed around their systematic part, and the thresholds cut it into the
four levels. The area under the curve between two thresholds is the
probability of that level. The curve below belongs to a student whose
systematic utility is zero.

```{r ordered-figure}
thresholds <- coef(smoking)[c("gamma[2]", "gamma[3]")]
cuts <- c(-Inf, 0, thresholds, Inf)
shades <- grey(seq(0.45, 0.9, length.out = length(smoking_levels)))
utility <- seq(-3.5, 3.5, length.out = 400)
plot(
  utility, dnorm(utility),
  type = "n", axes = FALSE, ylab = "",
  xlab = "latent utility of a student"
)
for (k in seq_along(smoking_levels)) {
  inside <- utility >= cuts[k] & utility <= cuts[k + 1]
  polygon(
    c(max(cuts[k], -3.5), utility[inside], min(cuts[k + 1], 3.5)),
    c(0, dnorm(utility[inside]), 0),
    col = shades[k], border = NA
  )
}
lines(utility, dnorm(utility), lwd = 2)
axis(1, at = c(-3, 0, 3))
legend(
  "topright", legend = smoking_levels, fill = shades, border = NA, bty = "n"
)
```

The three smoking levels sit close together in the narrow region to the
right of the first threshold: a student whose utility moves one unit to the
right passes through all of them. A coefficient shifts the whole curve
sideways, which is why one number per covariate describes its effect on all
four levels at once. The factor
`Exer` enters through its dummy variables relative to the students who
exercise frequently. Older students report smoking less. Students who
exercise only sometimes appear to smoke less than the frequent exercisers as
well, which is not what one might expect; with fewer than 50 smokers among
the 237 students, such contrasts rest on thin evidence.

What does a coefficient of an ordered model mean for the probabilities of
the levels? `interpret(type = "ame")` differentiates the probability of
every level with respect to age and averages the derivatives over the
students; `vignette("v04_prediction")` explains these marginal effects and
their variant `type = "mea"` in more detail:

```{r ordered-interpret}
age_effects <- interpret(smoking, type = "ame")
age_effects
```

One more year of age raises the probability of never smoking by about
`r round(100 * age_effects$mean[age_effects$alternative == "Never"], 1)`
percentage points and lowers the probabilities of all other levels. The four
effects sum to zero, because the levels are exhaustive.

## Rankings

Ranked data record the complete order of the alternatives. In wide format,
the response columns combine the response name with each alternative, for
example `rank_Xbox` or, with `delimiter = "."`, `ch.Xbox`. The model is the
same probit model as for unordered choices, but the likelihood uses the
full ordering of the utilities. The proof of concept simulates rankings of
three alternatives and compares the
coefficient and the free covariance parameters with the truth.

```{r ranked-recovery}
ranked_recovery <- fit(
  rank ~ x | 0,
  choice_type = "ranked",
  n_deciders = 300,
  dgp_parameters = list(
    beta = c(x = 1),
    Sigma = rbind(c(0, 0, 0), c(0, 1, 0.2), c(0, 0.2, 1))
  ),
  chains = 1
)
summary(ranked_recovery, variables = c("beta[x]", "Sigma[C,B]", "Sigma[C,C]"))
```

The full ordering carries more information than a single choice: the
coefficient is recovered closely, and the covariance and the variance of the
error differences within their posterior uncertainty.

Which gaming platform would you rank first? The `Game` data of the
**mlogit** package [@Croissant2020] contain complete rankings of six
platforms by 91 Dutch respondents, together with whether they own each
platform (`own`), their age, and their weekly gaming hours. The source study
develops a rank-ordered choice model for these data [@Fok2012]. The ranks
are stored in the columns `ch.Xbox`, `ch.PlayStation`, and so on, so the
response in the formula is `ch`.

```{r ranked}
data("Game", package = "mlogit")
gaming <- fit(
  ch ~ own | age + hours,
  data = Game,
  alternatives = c(
    "Xbox", "PlayStation", "PSPortable", "GameCube", "GameBoy", "PC"
  ),
  choice_type = "ranked",
  delimiter = ".",
  column_decider = NULL,
  chains = 1
)
coef(gaming)[1:6]
```

Owning a platform raises its rank, as @Fok2012 also find. The
alternative-specific constants and
the coefficients of `age` and `hours` are relative to the base alternative,
which is `GameBoy`, the first platform in alphabetical order, regardless of
the order in `alternatives`. Which platform gains from a gamer's experience?
For a ranked model, the marginal effects of `interpret()` refer to the
probability of being ranked first. `type = "mea"` evaluates them for a
respondent of average age who plays the average number of hours:

```{r ranked-interpret}
platform_effects <- interpret(gaming, type = "mea")
platform_effects
```

```{r ranked-numbers, include=FALSE}
pc_hours <- platform_effects$mean[
  platform_effects$covariate == "hours" & platform_effects$alternative == "PC"
]
```

The PC is the platform of the heavy gamers: every additional weekly hour
raises the probability of ranking it first by about
`r round(100 * pc_hours, 1)` percentage points, taken from the other
platforms. The age effects are small and uncertain; 91 respondents cannot
say much about them.

## RprobitB among its neighbors

Several R packages estimate related models. What sets **RprobitB** apart is the
combination of Bayesian probit errors with panel random coefficients and
finite or nonparametric mixtures, together with individual choice sets,
posterior prediction, and decider-level model evaluation. The comparison below concerns the documented estimation routines
of each package, not their data management or reporting features.

| Package | Main estimator and response family | Heterogeneity relevant here |
|:--|:--|:--|
| **Rchoice** [@Sarrias2016] | Simulated maximum likelihood for binary, ordinal, and count responses | Continuous random parameters in cross-sectional or panel data |
| **mlogit** [@Croissant2020] | Maximum likelihood for multinomial logit and extensions | Random-parameter logit, but no Bayesian probit mixture |
| **logitr** [@Helveston2023] | Maximum likelihood for multinomial and mixed logit, in preference or willingness-to-pay space | Normal and log-normal random parameters, but no discrete mixture |
| **MNP** [@Imai2005b; @Imai2025] | Bayesian multinomial probit, including varying choice sets and rankings | No mixture on panel-level random coefficients |
| **bayesm** [@Rossi2026] | Separate Bayesian multinomial probit and hierarchical multinomial logit routines | Finite and Dirichlet-process mixing for hierarchical logit, not hierarchical probit |
| **gmnl** [@Sarrias2017] | Maximum or simulated maximum likelihood for multinomial logit variants | Continuous, finite latent-class, and mixed-mixed heterogeneity |
| **RprobitB** [@Oelschlaeger2026d] | Bayesian binary, multinomial, ordered, and ranked probit | Panel random coefficients with a fixed, sparse finite, or Dirichlet-process mixture, and a weight-based update of the class count |

## Where to go next

The specification is only half of the model. Whether the coefficients are
allowed to differ between deciders is the other half, and
`vignette("v03_heterogeneity")` covers it. Once a model is specified and
fitted, `vignette("v04_prediction")` turns it into predictions and marginal
effects, and `vignette("v05_model_evaluation")` decides between competing
specifications.

## References
