---
title: "Modeling preference heterogeneity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling preference heterogeneity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
link-citations: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
  fig.align = "center"
)
```

A model with fixed coefficients says that everyone weighs price, time, and
comfort in the same way. That is rarely true. Some travelers watch the fare,
others the clock; some households would pay a premium for a local electricity
supplier, others would not. This vignette covers the two ways **RprobitB**
[@Oelschlaeger2026d] lets preferences differ between deciders: random
coefficients, which spread them continuously over the population, and latent
classes, which sort deciders into groups. Both are requested through an
argument of `fit()`, as are the specification choices of
`vignette("v02_model_variants")`, and both build on the fixed-coefficient
model of `vignette("v01_get_started")`. @Oelschlaeger2026c treats the
methodological background.

Each section first estimates the variant on simulated data and compares the
estimates with the parameters that generated them, then applies it to
empirical data from the **mlogit** package [@Croissant2020] and the
**choicedata** package [@Oelschlaeger2026a]. Panel data are essential
here: the population distribution of a coefficient is learned from deciders
observed repeatedly, so all simulations use several choices per decider.

```{r setup}
library(RprobitB)
set.seed(1)
```

## Random coefficients

Terms named in `random_effects` receive a coefficient of their own for every
decider, drawn from a population distribution that the sampler estimates
along with everything else. Such hierarchical models capture continuous
preference heterogeneity and allow inference about the coefficients of
individual deciders [@Allenby1998].

An unnamed vector requests correlated normal effects. A named vector selects
the mixing distribution of each term, written as
`"<covariate>" = "<distribution>"`, where `"ASC"` names the
alternative-specific constants:

| Value | Distribution | Coefficient |
|:--|:--|:--|
| `"cn"` | correlated normal | any sign |
| `"n"` | uncorrelated normal | any sign |
| `"cln"` | correlated log-normal | positive |
| `"ln"` | uncorrelated log-normal | positive |
| `"cln-"` | correlated log-normal | negative |
| `"ln-"` | uncorrelated log-normal | negative |

Two choices are behind these six values. The first is the shape of the
distribution. A normal coefficient may take any value, which is the natural
choice for an attribute that some deciders like and others dislike. A
log-normal coefficient is the exponential of a normal variable and is
therefore always positive, or, with the trailing minus, always negative.
Use it when theory fixes the sign: a higher price should not raise the
utility of any decider, however price-insensitive they are. Log-normal
effects are estimated on the latent normal scale, so `mu`, `Omega`, and the
individual draws refer to the normal variable whose exponential enters the
utility.

The second choice is whether an effect is correlated with the other
correlated effects. Deciders who care about travel time may also care about
comfort, and the `c` prefix lets the model learn that. Uncorrelated effects
keep their own variance but no covariance with anything else, which shows up
as zeros in the covariance matrix `Omega`.

The proof of concept combines both choices. The price coefficient is
negative log-normal and therefore uncorrelated, while travel time and
comfort receive correlated normal effects with a positive covariance.

```{r mixing}
mixing <- fit(
  choice ~ price + time + comfort | 0,
  random_effects = c(price = "ln-", time = "cn", comfort = "cn"),
  n_deciders = 400,
  n_occasions = 12,
  dgp_parameters = list(
    beta = c(price = -1, time = -0.8, comfort = 0.5),
    Omega = rbind(c(0.25, 0, 0), c(0, 0.4, 0.2), c(0, 0.2, 0.3))
  ),
  iterations = 2000,
  warmup = 1000,
  chains = 2,
  save_individual_draws = TRUE
)
summary(mixing)
```

The `dgp` column shows what the three specifications produce. All three
means are recovered, and for the price coefficient `mu[price]` is the mean
of the latent normal variable, so the coefficient entering the utility is
minus its exponential and stays negative for every decider. The covariance
matrix mirrors the specification: `time` and `comfort` share the estimated
covariance `Omega[time,comfort]`, while `price` has no covariance entry
with either of them, because an uncorrelated effect has none to estimate.

`interpret()` turns such coefficients into trade-offs. For a random effect
it uses the coefficient of the median decider, for the log-normal price
therefore minus the exponential of `mu[price]`:

```{r mixing-interpret}
interpret(mixing, reference = "price")
```

The `dgp_parameters` put the latent mean of the price effect at `-1`, so
the median price coefficient is `-exp(-1)`, about `-0.37`. One unit of time
is therefore worth `r round(0.8 / -exp(-1), 2)` units of price and one unit
of comfort `r round(-0.5 / -exp(-1), 2)`, which the intervals cover.

Individual coefficients are what the panel buys. `coef(level =
"individual")` returns the posterior mean coefficient of every decider, on
the latent normal scale for log-normal effects.

```{r individual}
head(coef(mixing, level = "individual"))
```

Their spread is the estimated heterogeneity. Turning the latent price values
into coefficients shows that every decider dislikes a higher price, and
that the tenth percentile reacts almost twice as strongly as the
ninetieth.

```{r individual-spread}
individual <- coef(mixing, level = "individual")
round(quantile(-exp(individual[, "price"]), c(0.1, 0.5, 0.9)), 2)
```

A picture says it faster. The histogram shows the price coefficient of each
of the 400 deciders, with the three quantiles above marked. Every decider
sits on the negative side, which is what the log-normal specification
enforces, and the tail to the left holds those who react most strongly to a
higher price.

```{r individual-figure}
price <- -exp(individual[, "price"])
hist(
  price,
  breaks = 30, col = "grey85", border = "white",
  main = "", xlab = "price coefficient of a decider"
)
abline(v = quantile(price, c(0.1, 0.5, 0.9)), lwd = 2)
```

### How much is a local electricity supplier worth?

The `Electricity` data of the **mlogit** package [@Croissant2020] come
from a stated choice experiment in which 361 US households chose 8 to 12
times among four hypothetical suppliers that differ in price (`pf`),
contract length (`cl`), whether the supplier is local (`loc`) or well-known
(`wk`), and whether it offers time-of-day (`tod`) or seasonal (`seas`) rates
[@Huber2001]. @Huber2001 used these data to check whether hierarchical
Bayesian and classical estimates of individual preferences agree, and found
that they largely do. The attribute columns end in the supplier number
without a delimiter, `pf1` to `pf4`, so an underscore is inserted first, and
the choice occasions of a household are numbered in the order of the rows.

Contract length and locality receive correlated normal random coefficients,
the other three attributes one coefficient for everyone. Keeping the number
of random coefficients small is deliberate: each one adds a row and a column
to every covariance draw, and the chains below have to be long enough that
the result can be believed rather than merely displayed. Fixing the price
coefficient to `-1` identifies the scale and puts every other coefficient in
cents per kWh, that is, expresses it directly as a willingness to pay. All
six attributes have to enter the model: suppliers with time-of-day or
seasonal rates carry a price of zero in these data, so leaving their dummies
out would hand their missing price level to the price coefficient and turn
it positive.

```{r electricity}
data("Electricity", package = "mlogit")
names(Electricity) <- sub("([a-z]+)([1-4])$", "\\1_\\2", names(Electricity))
Electricity$occasion <- ave(Electricity$id, Electricity$id, FUN = seq_along)
electricity <- fit(
  choice ~ pf + cl + loc + wk + tod + seas | 0,
  data = Electricity,
  random_effects = c("cl", "loc"),
  column_decider = "id",
  column_occasion = "occasion",
  scale = c(pf = -1),
  iterations = 3000,
  warmup = 1500,
  thin = 2,
  chains = 2,
  save_individual_draws = TRUE,
  progress = FALSE
)
summary(electricity)
```

The four supplier attributes other than price and contract length are
dummy variables. Households dislike long contracts and prefer local and
well-known suppliers, the three signs that the original analysis reports as
well. Because the price coefficient is fixed, `interpret()` reads the means
directly as willingness to pay in cents per kWh and attaches credible
intervals:

```{r electricity-interpret}
interpret(electricity, effects = c("cl", "loc", "wk"))
```

On average, households would accept a price about
`r round(coef(electricity)["mu[loc]"], 1)` cents per kWh higher for
a local supplier, and every additional year of contract length would have
to be compensated by a price about
`r round(abs(coef(electricity)["mu[cl]"]), 2)` cents lower. The
`Omega[...]` variables describe how widely these valuations differ between
households: the standard deviation of the locality premium is
`r round(sqrt(coef(electricity)["Omega[loc,loc]"]), 1)` cents,
so the households disagree about it by more than its average size.

A word on cost. This fit takes a few seconds because it draws one
coefficient pair per household in every iteration. Doubling the households
roughly doubles the time, and each further random coefficient adds a row and
a column to every covariance draw. The five-attribute model on all 361
households that the original study used is a matter of minutes rather than
seconds, and it needs far more than four thousand iterations before its
covariance entries settle down.

## Latent classes

A single normal distribution is a strong assumption about how preferences
are spread across a population. Perhaps there are commuters and leisure
travelers, or households that care about price and households that care
about service. `latent_class_effects` names the effects that differ between
`classes` latent classes. For a random effect this replaces its normal
distribution by a finite mixture of normals, which yields the latent-class
mixed multinomial probit model of @Oelschlaeger2021; @Oelschlaeger2026c
places it among the broader challenges of modeling choice-behavior
heterogeneity. Random effects that are not named keep one distribution for
all deciders, and a named coefficient that is not random gets one value per
class without varying within it, which a later subsection shows. Three
different questions call for three different specifications:

- `class_update = "fixed"` when exactly $K$ substantively meaningful classes
  are assumed;
- `class_update = "sparse"` when `classes` is a deliberately generous upper
  bound and redundant classes should empty out;
- `class_update = "dirichlet_process"` for a Dirichlet-process mixture whose
  number of occupied classes is itself random.

These models answer different questions, so pick one when you design the
study rather than after seeing the class-specific results. A fourth
option, `class_update = "weight_based"`, reproduces the split/remove/merge
heuristic of earlier **RprobitB** versions; it is a search procedure, not
another Bayesian model.

All four updates are demonstrated on one simulated data set with two
classes. The fixed-class fit simulates it; the other three are refits of it
through `update()`, which reuses its simulated data and replaces only the
arguments that are named, so that all updates are judged on identical
observations. The empirical application is shown for the fixed update.

### A fixed number of classes

For `class_update = "fixed"`, `classes = K` fixes the number of classes
and every class stays occupied, which is the direct specification for a
filled $K$-class model chosen a priori. If classes should instead be
allowed to empty, that is the question the sparse finite mixture below
answers.

The fixed model uses
$(w_1,\ldots,w_K)\sim\operatorname{Dirichlet}(\delta,\ldots,\delta)$. Its
default `class_concentration = 1` is uniform on the weight simplex;
$\delta < 1$ favors unequal weights near its boundary and $\delta > 1$
shrinks weights toward $1/K$. This prior remains influential when the
likelihood separates the classes only weakly. It can be fixed to another
positive scalar or given a gamma hyperprior with, for example,
`prior = list(class_concentration = c(shape = 2, rate = 2))`. Report the
choice and check sensitivity whenever class sizes drive the interpretation.

Which class is called the first one is arbitrary: swapping the labels leaves
the likelihood unchanged, so the sampler may swap them mid-run and a
posterior mean would average over both meanings. Ordering the classes by one
parameter, say by decreasing weight, does not reliably repair this
[@Stephens2000]. **RprobitB** instead leaves the sampler free and relabels the
retained draws afterwards. It reads off a representative grouping of the
deciders from the posterior co-clustering matrix [@Dahl2006], matches every
draw's allocation to it [@Papastamoulis2010], and permutes weights, means,
covariances, and allocations accordingly. Sorting the relabeled weights at
the end is a naming convention, not the thing that identifies the classes.

The proof of concept uses eight occasions per decider and two
well-separated classes: a majority whose coefficient is centered at `-1` and
a minority centered at `2`. These conditions are deliberately favorable,
because class-specific distributions are hard to learn from short panels.
Two chains of four thousand iterations are long enough here that the
diagnostics in the summary can be taken at face value.

```{r classes}
mixture <- fit(
  choice ~ x | 0,
  random_effects = "x",
  latent_class_effects = "x",
  classes = 2,
  n_deciders = 80,
  n_occasions = 8,
  save_individual_draws = TRUE,
  dgp_parameters = list(
    beta = list(c(x = -1), c(x = 2)),
    Omega = list(matrix(0.2), matrix(0.2)),
    weights = c(0.6, 0.4)
  ),
  iterations = 4000,
  warmup = 2000,
  chains = 2,
  progress = FALSE
)
summary(mixture, variables = c(
  "weight[1]", "weight[2]", "mu[x,1]", "mu[x,2]",
  "Omega[x,x,1]", "Omega[x,x,2]"
))
```

The relabeled weights and means can be read class by class. The class near
`-1` holds about half of these 80 deciders, a little less than its
population weight of `0.6`, and the class near `2` the rest. The class
variances are the least precise entries, but `rhat` stays at the threshold
and the effective sample sizes are in the hundreds, so the table describes
the posterior rather than one chain's wanderings. That check is worth
repeating for every mixture fit, because relabeling cannot repair poor
mixing or weak class separation.

The three updates below relabel their draws in the same way, so their
summaries carry class-specific weights and means as well. Where the number
of classes varies, a class exists only in part of the draws, and the
`occupied` column reports that share. Classes that no decider ever
occupies are left out, which is why the class indices in the tables below
may skip a number. The refits on shared data have no
`dgp` column, so a small helper compares them with the stored truth: the
most probable number of occupied classes, and, after assigning every decider
to the true class whose mean is closer to their posterior mean coefficient,
the share of deciders in the second class and the average coefficient in
each class. The true weight is the population weight; the realized share
among 80 simulated deciders differs from it by sampling variation.

```{r class-recovery}
truth <- mixture$simulation$dgp_parameters
class_means <- c(truth$beta[[1]][["x"]], truth$beta[[2]][["x"]])
recover_classes <- function(x) {
  occupancy <- latent_class_diagnostics(x)$occupancy
  individual <- coef(x, level = "individual")[, "x"]
  upper <- individual > mean(class_means)
  data.frame(
    variable = c("n_classes", "weight[2]", "mu[x,1]", "mu[x,2]"),
    dgp = c(length(truth$weights), truth$weights[2], class_means),
    estimate = c(
      occupancy$n_classes[which.max(occupancy$probability)],
      mean(upper), mean(individual[!upper]), mean(individual[upper])
    )
  )
}
recover_classes(mixture)
```

The two classes are visible in the individual coefficients. Each decider
contributes one value, and the two estimated class means mark where the
mixture places its classes.

```{r class-figure}
individual_x <- coef(mixture, level = "individual")[, "x"]
hist(
  individual_x,
  breaks = 30, col = "grey85", border = "white",
  main = "", xlab = "x coefficient of a decider"
)
abline(v = coef(mixture)[c("mu[x,1]", "mu[x,2]")], lwd = 2)
```

Two clearly separated groups of deciders, one on either side of zero, and
almost nobody in between: that is the pattern a mixture is meant to find. A
single normal distribution would have to put most of its mass where no
decider actually is.

The label-invariant co-clustering matrix and the allocation uncertainty are
available regardless of class names:

```{r class-diagnostics}
class_diagnostics <- latent_class_diagnostics(mixture)
class_diagnostics$occupancy
class_diagnostics$membership[1:6, ]
class_diagnostics$co_clustering[1:6, 1:6]
```

Asking for two filled classes is not evidence that two meaningful
populations exist. When $K$ is not fixed by the research design, fit each
candidate value to the same data and compare the decider-level predictive
accuracy. The `loo` package [@Vehtari2026] supplies the comparison and
`WAIC()` is the less diagnostic alternative;
`vignette("v05_model_evaluation")` covers both. Standard `AIC()` and `BIC()`
also work, because `logLik()` returns a proper `logLik` object, but their
asymptotic justification is fragile for mixtures, so PSIS-LOO is preferred.

```{r fixed-k-selection}
models_by_K <- lapply(1:3, function(K) {
  class_effects <- if (K > 1) "x" else character()
  update(
    mixture, classes = K, latent_class_effects = class_effects,
    iterations = 2000, warmup = 1000, thin = 20
  )
})
loo::loo_compare(
  lapply(models_by_K, loo, ghk_draws = 50, progress = FALSE)
)
```

The two-class model wins over the one-class model by several standard
errors, and the third class buys nothing: its expected predictive accuracy
is within noise of the second. That is the answer one hopes for from data
that were generated with two classes. The three fits are thinned and use
fewer GHK draws than the default, because each `loo()` call evaluates the
panel likelihood of every decider under every retained draw.

An empirical application of the fixed update follows in the next subsection,
where the train travelers of `vignette("v01_get_started")` split into two
classes with different values of time.

### Coefficients that differ only between classes

The mixture above lets the coefficient vary within a class as well: every
decider has their own `x` coefficient, drawn from the normal distribution of
their class. The classical latent class model is stricter. All members of a
class share the same coefficient, so the heterogeneity lives entirely in the
class membership [@Kamakura1989; @Greene2003], which market research knows
as segmentation.

**RprobitB** fits this model when a covariate in `latent_class_effects` has no
random effect: its fixed coefficient then takes one value per class,
reported as
`beta[<effect>,<class>]`. Random and class-specific effects share the class
allocation, so a model may have classes that differ in their price
coefficient while the time coefficient varies normally within each class,
either with the same distribution in every class or, if `time` is named in
both arguments, with class-specific distributions.

The proof of concept simulates two classes of travelers, a majority with a
price coefficient of `-2` and a minority with `-0.5`, who value time
equally.

```{r latent-class-effects}
class_specific <- fit(
  choice ~ price + time | 0,
  latent_class_effects = "price",
  classes = 2,
  n_deciders = 300,
  n_occasions = 10,
  dgp_parameters = list(
    beta = list(c(price = -2, time = 1), c(price = -0.5, time = 1)),
    weights = c(0.6, 0.4)
  ),
  chains = 1
)
summary(class_specific)
```

The estimated class weights of
`r round(coef(class_specific)[["weight[1]"]], 2)` and
`r round(coef(class_specific)[["weight[2]"]], 2)` and the class-specific price
coefficients of `r round(coef(class_specific)[["beta[price,1]"]], 2)` and
`r round(coef(class_specific)[["beta[price,2]"]], 2)` are close to the values
used for the simulation, and the common time coefficient is estimated from
all deciders at once. Since every class has its own price coefficient,
`interpret()` reports one willingness to pay per class:

```{r latent-class-interpret}
interpret(class_specific, reference = "price")
```

Back to the train travelers of `vignette("v01_get_started")`: do all of
them trade time against money at the same rate, or are there classes with
different values of time? The fit below uses the first 100 travelers, keeps the
price normalization, and gives the time coefficient two class-specific
values, so that each class has its own value of an hour of travel time,
while the remaining coefficients are common.

```{r train-classes}
data("Train", package = "mlogit")
Train$price_A <- Train$price_A / 100 / 2.20371
Train$price_B <- Train$price_B / 100 / 2.20371
Train$time_A <- Train$time_A / 60
Train$time_B <- Train$time_B / 60
train_small <- Train[Train$id %in% unique(Train$id)[1:100], ]
train_classes <- fit(
  choice ~ price + time + change + factor(comfort) | 0,
  data = train_small,
  latent_class_effects = "time",
  classes = 2,
  column_decider = "id",
  column_occasion = "choiceid",
  scale = c(price = -1),
  iterations = 2500,
  warmup = 1250,
  chains = 2,
  progress = FALSE
)
summary(train_classes, variables = c(
  "weight[1]", "weight[2]", "beta[time,1]", "beta[time,2]"
))
```

With the price coefficient fixed at `-1`, the class-specific time
coefficients are values of travel time in euro per hour, which `interpret()`
reports class by class:

```{r train-classes-interpret}
time_by_class <- interpret(train_classes, effects = "time")
time_by_class
```

The larger class of
`r round(100 * coef(train_classes)[["weight[1]"]])` percent of the
travelers values an hour at
`r round(abs(time_by_class$mean[1]), 1)` euro, the smaller class at
`r round(abs(time_by_class$mean[2]), 1)` euro. The small class takes
the faster trip almost regardless of its price, a pattern that a single
normal distribution of the time coefficient would blur into a heavy tail.
The diagnostics of both weights and both time coefficients are sound, so
this is a result rather than a demonstration.

### Weight-based class updates

The weight-based update of @Oelschlaeger2021 is kept for backward
compatibility and for reproducing earlier **RprobitB** analyses;
@Oelschlaeger2026c discusses its broader context. Every `buffer` warmup
iterations, the heuristic attempts at most one operation, in this order:

1. remove the smallest class if its weight is below `epsmin`;
2. split the largest class if its weight is above `epsmax`;
3. merge the closest pair if the Euclidean distance of their means is below
   `deltamin`.

After a split, the new means are displaced by `deltashift` times the leading
within-class standard deviation. The original defaults are `buffer = 50`,
`epsmin = 0.01`, `epsmax = 0.7`, `deltamin = 0.1`, and `deltashift = 0.5`;
override them with `weight_based_control`. `max_classes` bounds the
splitting.

These dimension changes have no Metropolis-Hastings acceptance step and
correspond to no prior on the class count. Updates stop after warmup, so the
retained draws condition on whatever dimension each chain selected. The
reported `n_classes` is accordingly a heuristic result, not a posterior
distribution. There is no universally valid calibration of the five
constants: any use should report them and repeat the analysis over
substantively plausible settings. For inferential claims, prefer the fixed,
sparse finite, or Dirichlet-process specifications.

The proof of concept keeps the update defaults, and a second run tightens
two of the five constants through `weight_based_control`: it checks more
often and refuses to split a class before its weight passes 0.8.

```{r weight-based}
weight_based <- update(mixture, class_update = "weight_based")
recover_classes(weight_based)
tuned <- update(
  mixture,
  class_update = "weight_based",
  weight_based_control = list(buffer = 25, epsmax = 0.8)
)
recover_classes(tuned)
```

The heuristic keeps both classes, and the share and the class means come out
as in the table of the fixed fit. That is what it was designed for; the
caveats above concern what its result means, not whether it can find two
well-separated classes.

### Sparse finite mixtures

A sparse finite mixture fixes a generous upper bound $K$, permits empty
classes, and places a symmetric Dirichlet prior

\[
(w_1,\ldots,w_K) \mid e_0 \sim \operatorname{Dirichlet}(e_0,\ldots,e_0)
\]

on the weights. A small $e_0$ favors emptying redundant classes. Under
regularity conditions, overfitted classes empty asymptotically when the
Dirichlet parameter is below half the dimension of a class-specific
parameter [@Rousseau2011]. This result motivates sparsity but does not make
the prior harmless: both $K$ and $e_0$ affect the posterior of the occupied
count [@FruehwirthSchnatter2019].

Before looking at the data, the expected number of occupied classes
after $N$ deciders under a fixed $e_0$ is

\[
K\left[1-
\frac{\Gamma(Ke_0)\,\Gamma((K-1)e_0+N)}
     {\Gamma((K-1)e_0)\,\Gamma(Ke_0+N)}\right].
\]

The expression makes $e_0$ readable as a statement about class counts
instead of an arbitrary small number. Evaluated at the prior mean
$e_0 = 0.005$ of
the default hyperprior below, with $K = 6$ and the 80 deciders of the
simulated data, it gives about 1.1 occupied classes: the prior leans
firmly towards simplicity. For a gamma hyperprior, draw plausible values of
$e_0$, evaluate the expression, and inspect the implied distribution before
fitting.

Unlike the threshold-based updater, `class_update = "sparse"` uses the
finite-mixture posterior itself to empty classes. The default
`class_concentration = c(shape = 1, rate = 200)` puts the gamma hyperprior
$e_0\sim\operatorname{Gamma}(1,200)$ on the shape-rate scale used by R. The
concentration is updated from its full conditional with a fixed
log-random-walk Metropolis-Hastings step. Its proposal scale affects the
sampling efficiency, not the posterior target; `rhat` and the effective
sample size diagnose that step like any other scalar posterior variable.

The proof of concept offers six classes to the two-class data.

```{r sparse}
sparse <- update(mixture, classes = 6, class_update = "sparse")
summary(sparse)
latent_class_diagnostics(sparse)$occupancy
recover_classes(sparse)
```

Despite the prior's preference for a single class, the occupied count
concentrates on two, and the class structure is recovered as by the fixed
model. The concentration parameter is the slowest-mixing quantity, as its
effective sample size shows. A sensitivity analysis should be run with
longer chains until both `n_classes` and `class_concentration` mix
satisfactorily.

For a sensitivity analysis, vary the upper bound and the concentration
prior, then compare occupancy, co-clustering, predictive accuracy, and the
substantive conclusions. A useful less-sparse alternative is
$e_0\sim\operatorname{Gamma}(2,4K)$, whose expected total mass $K e_0$
matches the default DP precision below
[@FruehwirthSchnatter2019].

```{r sparse-sensitivity}
sparse_less <- update(
  mixture,
  classes = 6,
  class_update = "sparse",
  prior = list(class_concentration = c(shape = 2, rate = 24))
)
latent_class_diagnostics(sparse_less)$occupancy
recover_classes(sparse_less)
```

Six classes on offer and a prior that resists emptying them: the occupied
count now spreads from two to five instead of concentrating on two. The
class means are still recovered, because the extra classes hold few
deciders, but the answer to the question how many classes there are has
changed with the prior. That is exactly what a sensitivity analysis is for,
and it is the reason to report the concentration prior alongside the
occupancy table rather than the table alone.

### Dirichlet-process mixtures

For `class_update = "dirichlet_process"`, the precision $\alpha$ controls
the prior tendency to open new classes. The default
$\alpha\sim\operatorname{Gamma}(2,4)$ has mean 0.5; a larger $\alpha$
generally favors more occupied classes. Conditional on a fixed $\alpha$,
the Chinese-restaurant-process prior expects
$\sum_{i=1}^{N}\alpha/(\alpha+i-1)$ occupied classes after $N$ deciders,
which for $\alpha = 0.5$ and 80 deciders is about 3.2. Use this quantity to
translate beliefs about plausible class counts into candidate values or
gamma hyperpriors. **RprobitB** updates $\alpha$ with the beta-gamma
augmentation of @Escobar1995 and the allocations with the
auxiliary-parameter algorithm of @Neal2000, using ten auxiliary classes
per reassignment; this choice affects computation and mixing, not the
posterior target. `max_classes` is a computational cap, not the inferred
count: if posterior mass reaches the cap, refit with a larger value.

The proof of concept lets the process decide how many classes the
two-class data need.

```{r dirichlet}
dynamic <- update(
  mixture, class_update = "dirichlet_process", iterations = 2500,
  warmup = 1250
)
summary(dynamic)
latent_class_diagnostics(dynamic)$occupancy
recover_classes(dynamic)
```

`max_classes` caps how many classes the sampler may open. The cap is a
computational limit, not an estimate, and **RprobitB** says so when the draws
reach it:

```{r max-classes}
capped <- update(dynamic, max_classes = 3)
latent_class_diagnostics(capped)$occupancy
```

The warning is the signal to refit with a larger cap before reading the
occupancy table.

The occupancy distribution is wider than under the sparse finite prior and
keeps a substantial share of its mass on three or more classes. That is
the prior at work: for a data set of this size it expects about three
occupied classes, and 80 deciders do not overrule it completely.
The label-invariant comparison nevertheless recovers the two true groups,
because any superfluous class holds only a few deciders. Whether the
occupancy table can be trusted is decided by the `rhat` and the effective
sample size of the occupied count. Both are acceptable here, though the
effective sample size of `n_classes` is the smallest in the table: enough
for the broad shape of the distribution, not for its third decimal. The run
stays below the cap, so the remaining step is a sensitivity analysis over
the prior on $\alpha$. Material mass at the cap would be a different matter:
increase `max_classes` before interpreting the result.

The two priors can be compared directly by the posterior of the occupied
count. The sparse finite prior concentrates on the two classes that the data
contain, the Dirichlet process spreads its mass wider and keeps classes
that hold only a few deciders.

```{r occupancy-figure}
occupancy_shares <- function(x, bound) {
  occupancy <- latent_class_diagnostics(x)$occupancy
  shares <- stats::setNames(numeric(bound), seq_len(bound))
  shares[occupancy$n_classes] <- occupancy$probability
  shares
}
bound <- max(
  latent_class_diagnostics(sparse)$occupancy$n_classes,
  latent_class_diagnostics(dynamic)$occupancy$n_classes
)
barplot(
  rbind(
    "sparse finite" = occupancy_shares(sparse, bound),
    "Dirichlet process" = occupancy_shares(dynamic, bound)
  ),
  beside = TRUE, col = c("grey60", "grey85"), border = "white",
  xlab = "occupied classes", ylab = "posterior probability", legend.text = TRUE,
  args.legend = list(x = "topright", bty = "n")
)
```

The occupied class count, the concentration, and the co-clustering
probabilities do not refer to class labels at all and are therefore the
safest summaries here. The relabeled class-specific parameters come with
their `occupied` share: a class that exists in only a small fraction of the
draws describes a handful of deciders, not a group in the population. Read
the two
together, and treat classes with a small share as sampling noise rather
than as findings.

The prior on $\alpha$ belongs in the reported analysis. For example,
$\alpha\sim\operatorname{Gamma}(1,200/K)$ has the same prior mean as $K e_0$
under the default sparse finite prior. Comparing this matched DP fit with a
sparse finite fit helps separate the effect of the mixing model from that of
the concentration prior [@FruehwirthSchnatter2019]. In every
case, report the prior, the posterior of `class_concentration`, the
occupancy distribution, whether the cap was reached, and the sensitivity to
defensible alternatives.

## Ordered and ranked responses

Both mechanisms work for ordered and ranked responses as well. What they
need is a panel: a population distribution is learned from deciders observed
repeatedly, and the ordered and ranked data sets of the previous vignettes
record one response per decider. The demonstration therefore uses a
simulated panel of rankings, 100 deciders who order three alternatives five
times each, whose coefficient varies normally around `-1`.

```{r ranked-random}
ranked_random <- fit(
  rank ~ x | 0,
  choice_type = "ranked",
  random_effects = "x",
  n_deciders = 100,
  n_occasions = 5,
  n_alternatives = 3,
  dgp_parameters = list(beta = c(x = -1), Omega = matrix(0.3)),
  iterations = 4000,
  warmup = 2000,
  chains = 2,
  progress = FALSE
)
summary(ranked_random)
```

The population mean and its variance are recovered next to their true
values, and a full ranking of three alternatives per occasion carries enough
information to do so from a hundred deciders. Latent classes are requested
the same way, with `latent_class_effects` and `classes`, and everything this
vignette says about relabeling, occupancy, and the concentration priors
applies unchanged.

Ordered responses accept the same arguments. Here a hundred deciders rate
something on a three-point scale eight times each, again with a coefficient
that varies normally around `-1`.

```{r ordered-random}
ordered_random <- fit(
  choice ~ x | 0,
  alternatives = c("low", "middle", "high"),
  choice_type = "ordered",
  random_effects = "x",
  n_deciders = 100,
  n_occasions = 8,
  dgp_parameters = list(
    beta = c(x = -1), Omega = matrix(0.3), gamma = c(0, 1)
  ),
  iterations = 4000,
  warmup = 2000,
  chains = 2,
  progress = FALSE
)
summary(ordered_random)
```

The threshold `gamma[2]` is estimated alongside the population mean and its
variance, and its diagnostics are worth a look: the thresholds are the one
block of an ordered model that no Gibbs step can draw in closed form, so they
are updated by a random walk whose step size is tuned during warmup. That
tuning is what keeps the `gamma` rows mixing once random coefficients enter,
and a long panel makes the thresholds easier still to pin down. Read their
`rhat` and effective sample size the way `vignette("v01_get_started")` reads
any diagnostic.

## Where to go next

Heterogeneity pays off in prediction: with individual coefficients in hand,
`predict(type = "conditional")` in `vignette("v04_prediction")` tailors the
choice probabilities to each decider that took part in the fit. Deciding how
many classes the data support, or whether random coefficients are worth
their price at all, is the business of
`vignette("v05_model_evaluation")`.

## References
