---
title: "Modeling preference heterogeneity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling preference heterogeneity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
link-citations: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
  fig.align = "center"
)
```

The probit model assigns alternative $j$ at occasion $t$ of
decider $n$ the latent utility
$U_{ntj} = X_{ntj}^\top \beta_n + \epsilon_{ntj}$ with the coefficient
vector $\beta_n$. A model with fixed coefficients sets $\beta_n = \beta$
for all deciders and thereby assumes that all deciders weigh price, time,
and comfort in the same way. This is rarely true: some travelers react mainly to
the fare, others to travel time, and some households would pay a premium for
a local electricity supplier while others would not. **RprobitB** lets
coefficients differ between deciders in two ways, which can be combined:
random coefficients follow a continuous distribution over the population,
and latent classes divide the deciders into groups. Both are requested
through arguments of `fit()`. @Oelschlaeger2026c treats the methodological
background. Each variant is first estimated on simulated data, where the
estimates can be compared with the parameters that generated them, and then
applied to data of the **mlogit** package [@Croissant2020]. Panel data are
essential here: the population distribution of a coefficient is estimated
from deciders observed repeatedly, so all simulations use several choices
per decider.

```{r setup}
library(RprobitB)
set.seed(1)
```

## Random coefficients

Each term named in `random_effects` receives a separate coefficient for every
decider, drawn from a population distribution whose parameters the sampler
estimates together with the other parameters. Such hierarchical models
capture continuous preference heterogeneity and allow inference about the
coefficients of individual deciders [@Allenby1998]. The population
distribution is normal on a latent scale: the random coefficients of decider
$n$ are $\beta_n \sim \mathrm{N}(\mu, \Omega)$ with the mean vector $\mu$ and
the covariance matrix $\Omega$, reported as `mu[<effect>]` and
`Omega[<effect>,<effect>]`. The mixing distribution of an effect decides how
its latent normal variable enters the utility. An unnamed vector requests
correlated normal effects. A named vector selects the mixing distribution of
each term as `"<covariate>" = "<distribution>"`, where `"ASC"` names the
alternative-specific constants:

| Value | Distribution | Coefficient |
|:--|:--|:--|
| `"cn"` | correlated normal | any sign |
| `"n"` | uncorrelated normal | any sign |
| `"cln"` | correlated log-normal | positive |
| `"ln"` | uncorrelated log-normal | positive |
| `"cln-"` | correlated log-normal | negative |
| `"ln-"` | uncorrelated log-normal | negative |

The six values combine two choices. The first is the shape of the
distribution. A normal coefficient can take any value, which suits an
attribute that some deciders like and others dislike. A log-normal
coefficient is the exponential of a normal variable and therefore always
positive, or, with the trailing minus, always negative. It suits an attribute
whose sign is fixed by theory: a higher price does not raise the utility of
any decider. Log-normal effects are estimated on the latent normal scale, so
`mu`, `Omega`, and the individual draws refer to the normal variable whose
exponential enters the utility. The second choice is whether an effect is
correlated with the other correlated effects. Deciders who value travel time
may also value comfort, and the `c` prefix adds the covariance between such
effects to the model. An uncorrelated effect has its own variance but no
covariance with any other effect, which appears as zeros in `Omega`.

The following demonstration combines both choices. The price coefficient is
negative log-normal and therefore uncorrelated. Travel time and comfort
receive correlated normal effects with a positive covariance.

```{r mixing}
mixing <- fit(
  choice ~ price + time + comfort | 0,
  random_effects = c(price = "ln-", time = "cn", comfort = "cn"),
  n_deciders = 400,
  n_occasions = 12,
  dgp_parameters = list(
    beta = c(price = -1, time = -0.8, comfort = 0.5),
    Omega = rbind(c(0.25, 0, 0), c(0, 0.4, 0.2), c(0, 0.2, 0.3))
  ),
  iterations = 2000,
  warmup = 1000,
  chains = 2,
  save_individual_draws = TRUE
)
summary(mixing)
```

```{r mixing-deviation, include=FALSE}
mixing_table <- summary(mixing)$posterior
mixing_mu <- startsWith(mixing_table$variable, "mu")
mixing_deviation <- max(
  abs(mixing_table$mean - mixing_table$dgp)[mixing_mu] /
    mixing_table$sd[mixing_mu]
)
```

The `dgp` column lists the parameters that generated the data. The posterior
means of the three population means deviate from their true values by at
most `r round(mixing_deviation, 1)` posterior standard deviations. For the
price coefficient, `mu[price]` is the mean of the latent normal variable, so
the coefficient that enters the utility is minus its exponential and
negative for every decider. `time` and `comfort` share the estimated
covariance `Omega[time,comfort]`, while `price` has no covariance entry with
either of them.

`interpret()` converts the coefficients into trade-offs. For a random effect
it uses the coefficient of the median decider, which for the log-normal
price is minus the exponential of `mu[price]`:

```{r mixing-interpret}
mixing_trade <- interpret(mixing, reference = "price")
mixing_trade
```

```{r mixing-inside, include=FALSE}
mixing_true <- c(time = 0.8 / -exp(-1), comfort = -0.5 / -exp(-1))
mixing_rows <- match(names(mixing_true), mixing_trade$effect)
mixing_inside <- all(
  mixing_trade$lower[mixing_rows] < mixing_true &
    mixing_true < mixing_trade$upper[mixing_rows]
)
```

The `dgp_parameters` set the latent mean of the price effect to `-1`, so
the median price coefficient is `-exp(-1)`, about `-0.37`. One unit of time
is therefore worth `r round(0.8 / -exp(-1), 2)` units of price and one unit
of comfort `r round(-0.5 / -exp(-1), 2)` units.
`r if (mixing_inside) "Both values lie" else "Not both values lie"` inside
the 95% credible intervals.

The panel structure makes individual coefficients estimable.
`coef(level = "individual")` returns the posterior mean coefficient of every
decider, on the latent normal scale for log-normal effects. The histogram
shows the price coefficient of each of the 400 deciders, all negative as the
log-normal specification enforces, with the left tail containing the
deciders who react most strongly to a higher price.

```{r individual}
individual <- coef(mixing, level = "individual")
head(individual)
price <- -exp(individual[, "price"])
hist(
  price,
  breaks = 30, col = "grey85", border = "white",
  main = "", xlab = "price coefficient of a decider"
)
```

### Willingness to pay for electricity supplier attributes

The `Electricity` data of the **mlogit** package come from a stated choice
experiment in which 361 US households chose 8 to 12 times among four
hypothetical suppliers. The suppliers differ in price (`pf`), contract
length (`cl`), whether the supplier is local (`loc`) or well-known (`wk`),
and whether it offers time-of-day (`tod`) or seasonal (`seas`) rates
[@Huber2001]. The attribute columns end in the supplier number without a
delimiter, `pf1` to `pf4`, so an underscore is inserted first, and the
choice occasions of a household are numbered in the order of the rows.

Contract length and locality receive correlated normal random coefficients,
the other attributes one coefficient for all households. Fixing the price
coefficient to `-1` identifies the scale and expresses every other
coefficient in cents per kWh, that is, directly as a willingness to pay. All
six attributes must enter the model: suppliers with time-of-day or seasonal
rates have a price of zero in these data, so without their dummies the price
coefficient would absorb the effect of their zero price.

```{r electricity}
data("Electricity", package = "mlogit")
names(Electricity) <- sub("([a-z]+)([1-4])$", "\\1_\\2", names(Electricity))
Electricity$occasion <- ave(Electricity$id, Electricity$id, FUN = seq_along)
electricity <- fit(
  choice ~ pf + cl + loc + wk + tod + seas | 0,
  data = Electricity,
  random_effects = c("cl", "loc"),
  column_decider = "id",
  column_occasion = "occasion",
  scale = c(pf = -1),
  iterations = 3000,
  warmup = 1500,
  thin = 2,
  chains = 2,
  save_individual_draws = TRUE,
  progress = FALSE
)
summary(electricity)
```

```{r electricity-numbers, include=FALSE}
loc_mean <- coef(electricity)[["mu[loc]"]]
loc_spread <- sqrt(coef(electricity)[["Omega[loc,loc]"]])
```

Because the price coefficient is fixed, `interpret()` reports the
coefficients directly as willingness to pay in cents per kWh, with credible
intervals:

```{r electricity-interpret}
interpret(electricity, effects = c("cl", "loc", "wk"))
```

On average, households would accept a price about `r round(loc_mean, 1)`
cents per kWh higher for a local supplier, and would need a price about
`r round(abs(coef(electricity)["mu[cl]"]), 2)` cents lower for each
additional year of contract length. The `Omega[...]` variables describe how
much these valuations vary between households: the standard deviation of
the locality premium across households is `r round(loc_spread, 1)` cents,
which `r if (loc_spread > loc_mean) "exceeds" else "is below"` its mean.

## Latent classes

A single normal distribution is a strong assumption about the distribution
of preferences in a population. There may be commuters and leisure
travelers, or households that focus on price and households that focus on
service. `latent_class_effects` names the effects that differ between
`classes` latent classes. Every decider belongs to exactly one class. The
class weights $w_1, \dots, w_K$ are the probabilities of membership and sum
to one, and the class allocation of every decider is a latent variable that
the sampler draws together with the parameters. A class is occupied if at
least one decider is allocated to it. Naming a random effect in
`latent_class_effects` replaces its normal distribution by a finite mixture
of normals, which yields the latent-class mixed multinomial probit model of
@Oelschlaeger2021. Random effects that are not named keep one distribution
for all deciders, and a named coefficient that is not random takes one value
per class and does not vary within it, as in the classical latent class
model [@Kamakura1989; @Greene2003]. `class_update` selects how the number
of classes is treated:

- `class_update = "fixed"` when exactly $K$ substantively meaningful classes
  are assumed;
- `class_update = "sparse"` when `classes` is a generous upper bound and
  redundant classes should empty;
- `class_update = "dirichlet_process"` for a Dirichlet-process mixture whose
  number of occupied classes is itself random;
- `class_update = "weight_based"` for the split, remove, and merge heuristic
  of earlier **RprobitB** versions, which is a search procedure, not a
  Bayesian model.

All four updates are demonstrated on one simulated data set with two
classes. The fixed-class fit simulates the data; the other three are refits
through `update()`, which reuses the simulated data, so all updates are
compared on identical observations.

### A fixed number of classes

For `class_update = "fixed"`, `classes = K` fixes the number of classes,
and every class stays occupied. The class weights have the prior
$(w_1,\ldots,w_K)\sim\operatorname{Dirichlet}(\delta,\ldots,\delta)$, where
the default `class_concentration = 1` is uniform on the weight simplex. The
class labels are arbitrary: swapping them leaves the likelihood unchanged,
so the sampler may swap them during a run. **RprobitB** leaves the sampler
unconstrained and relabels the retained draws afterwards. It derives a
representative grouping of the deciders from the posterior co-clustering
matrix, whose entries are the posterior probabilities that two deciders
belong to the same class [@Dahl2006], matches the allocation of every draw
to this grouping [@Papastamoulis2010], and permutes weights, means,
covariances, and allocations accordingly. The classes are then numbered by
decreasing weight, which is a naming convention.

The following demonstration uses eight occasions per decider and two
well-separated classes: a majority with coefficients centered at `-1` and a
minority centered at `2`. The class-specific means and variances mix slowly,
so the two chains run twenty thousand iterations each and retain every fifth
draw.

```{r classes}
mixture <- fit(
  choice ~ x | 0,
  random_effects = "x",
  latent_class_effects = "x",
  classes = 2,
  n_deciders = 80,
  n_occasions = 8,
  save_individual_draws = TRUE,
  dgp_parameters = list(
    beta = list(c(x = -1), c(x = 2)),
    Omega = list(matrix(0.2), matrix(0.2)),
    weights = c(0.6, 0.4)
  ),
  iterations = 20000,
  warmup = 10000,
  thin = 5,
  chains = 2,
  progress = FALSE
)
mixture_summary <- summary(mixture, variables = c(
  "weight[1]", "weight[2]", "mu[x,1]", "mu[x,2]",
  "Omega[x,x,1]", "Omega[x,x,2]"
))
mixture_summary
```

```{r mixture-diagnostics, include=FALSE}
mixture_table <- mixture_summary$posterior
mixture_mu <- mixture_table$mean[
  match(c("mu[x,1]", "mu[x,2]"), mixture_table$variable)
]
mixture_lower <- which.min(abs(mixture_mu + 1))
mixture_weight <- mixture_table$mean[
  mixture_table$variable == paste0("weight[", mixture_lower, "]")
]
mixture_rhat <- max(mixture_table$rhat)
mixture_ess <- min(mixture_table$ess_bulk)
mixture_slowest <- paste0(
  "`", mixture_table$variable[which.min(mixture_table$ess_bulk)], "`"
)
```

The relabeled weights and means can be read class by class. The class near
`-1` has a posterior mean weight of `r round(mixture_weight, 2)` against a
population weight of `0.6`, and the class near `2` the rest. The largest
`rhat` in the table is `r sprintf("%.3f", mixture_rhat)`, and the smallest
bulk effective sample size, `r round(mixture_ess)`, belongs to
`r mixture_slowest`. Check both for every mixture fit, because relabeling
cannot compensate for poor mixing or weak class separation.

`latent_class_diagnostics()` returns the posterior distribution of the
number of occupied classes, the membership probabilities of the deciders
after relabeling, and the co-clustering matrix, which does not depend on the
class labels:

```{r class-diagnostics}
class_diagnostics <- latent_class_diagnostics(mixture)
class_diagnostics$occupancy
class_diagnostics$membership[1:6, ]
class_diagnostics$co_clustering[1:6, 1:6]
```

The three refits below relabel their draws in the same way. When the number
of classes varies, a class exists only in part of the draws, and the
`occupied` column of `summary()` reports this share. The refits have no
`dgp` column, so a small helper compares them with the stored true
parameters: the most probable number of occupied classes and, after
assigning every decider to the true class whose mean is closer to their
posterior mean coefficient, the share of deciders in the second class and
the average coefficient in each class.

```{r class-sim}
truth <- mixture$simulation$dgp_parameters
class_means <- c(truth$beta[[1]][["x"]], truth$beta[[2]][["x"]])
recover_classes <- function(x) {
  occupancy <- latent_class_diagnostics(x)$occupancy
  individual <- coef(x, level = "individual")[, "x"]
  upper <- individual > mean(class_means)
  data.frame(
    variable = c("n_classes", "weight[2]", "mu[x,1]", "mu[x,2]"),
    dgp = c(length(truth$weights), truth$weights[2], class_means),
    estimate = c(
      occupancy$n_classes[which.max(occupancy$probability)],
      mean(upper), mean(individual[!upper]), mean(individual[upper])
    )
  )
}
mixture_sim <- recover_classes(mixture)
mixture_sim
```

Fitting two filled classes is not evidence that two meaningful populations
exist. When $K$ is not fixed by the research design, fit each candidate
value to the same data and compare the decider-level predictive accuracy
with the **loo** package [@Vehtari2026], as the vignette
[Bayesian model evaluation][v05] describes. The fits are thinned, because
each `loo()` call evaluates the panel likelihood of every decider under
every retained draw. This likelihood is a multivariate normal probability
that is simulated with the GHK simulator, and `ghk_draws = 50` reduces its
simulation draws from the default 500.

```{r fixed-k-selection}
models_by_K <- lapply(1:3, function(K) {
  class_effects <- if (K > 1) "x" else character()
  update(
    mixture, classes = K, latent_class_effects = class_effects,
    iterations = 2000, warmup = 1000, thin = 20
  )
})
k_comparison <- loo::loo_compare(
  lapply(models_by_K, loo, ghk_draws = 50, progress = FALSE)
)
k_comparison
```

```{r fixed-k-numbers, include=FALSE}
k_row <- function(model) k_comparison[k_comparison$model == model, ]
k_one <- abs(k_row("model1")$elpd_diff)
k_one_se <- k_row("model1")$se_diff
k_pair <- abs(k_row("model3")$elpd_diff - k_row("model2")$elpd_diff)
k_pair_se <- max(k_row("model2")$se_diff, k_row("model3")$se_diff)
```

The one-class model falls short of the best model by `r round(k_one)`,
`r round(k_one / k_one_se)` times the standard error of the difference,
whereas the two- and the three-class model differ by only
`r round(k_pair, 1)` with a standard error of `r round(k_pair_se, 1)`, so
the third class `r if (k_pair < k_pair_se) "adds nothing" else "adds little"`.
This is the expected result for data generated with two classes.

### Class-specific coefficients

Do the train travelers of the vignette [Get started with RprobitB][v01] all
trade time against money at the same rate, or are there classes with
different values of time? The fit below uses the first 100 travelers, fixes
the price coefficient to `-1`, and gives the time coefficient two
class-specific values through `latent_class_effects` without a random
effect, so each class has its own value of travel time and the remaining
coefficients are common to both classes.

```{r train-classes}
data("Train", package = "mlogit")
Train$price_A <- Train$price_A / 100 / 2.20371
Train$price_B <- Train$price_B / 100 / 2.20371
Train$time_A <- Train$time_A / 60
Train$time_B <- Train$time_B / 60
train_small <- Train[Train$id %in% unique(Train$id)[1:100], ]
train_classes <- fit(
  choice ~ price + time + change + factor(comfort) | 0,
  data = train_small,
  latent_class_effects = "time",
  classes = 2,
  column_decider = "id",
  column_occasion = "choiceid",
  scale = c(price = -1),
  iterations = 10000,
  warmup = 5000,
  thin = 2,
  chains = 2,
  progress = FALSE
)
train_classes_summary <- summary(train_classes, variables = c(
  "weight[1]", "weight[2]", "beta[time,1]", "beta[time,2]"
))
train_classes_summary
```

```{r train-classes-diagnostics, include=FALSE}
train_classes_table <- train_classes_summary$posterior
train_classes_rhat <- max(train_classes_table$rhat)
train_classes_ess <- min(train_classes_table$ess_bulk)
```

With the price coefficient fixed at `-1`, the class-specific time
coefficients are values of travel time in euro per hour, which `interpret()`
reports class by class:

```{r train-classes-interpret}
time_by_class <- interpret(train_classes, effects = "time")
time_by_class
```

The larger class, `r round(100 * coef(train_classes)[["weight[1]"]])` percent
of the travelers, values an hour at
`r round(abs(time_by_class$mean[1]), 1)` euro, the smaller class at
`r round(abs(time_by_class$mean[2]), 1)` euro. The smaller class chooses the
faster trip almost regardless of its price, a pattern that a single normal
distribution of the time coefficient would represent as a heavy tail. The
largest `rhat` of the four variables is
`r sprintf("%.3f", train_classes_rhat)` and the smallest bulk effective
sample size `r round(train_classes_ess)`.

### Weight-based class updates

The weight-based update of @Oelschlaeger2021 is kept for reproducing earlier
**RprobitB** analyses. Every `buffer` warmup iterations, it removes the
smallest class if its weight is below `epsmin`, splits the largest class if
its weight is above `epsmax`, or merges the closest pair of classes if the
distance of their means is below `deltamin`, at most one operation in this
order. `weight_based_control` overrides the defaults of these constants,
and `max_classes` bounds the splitting. These dimension changes correspond
to no prior on the number of classes, and they stop after warmup, so the
reported `n_classes` is the outcome of a search, not a posterior
distribution. For inferential claims, prefer the fixed, sparse finite, or
Dirichlet-process specifications.

```{r weight-based}
weight_based <- update(mixture, class_update = "weight_based")
weight_based_sim <- recover_classes(weight_based)
weight_based_sim
```

```{r weight-based-numbers, include=FALSE}
heuristic_shift <- max(abs(
  weight_based_sim$estimate[-1] - mixture_sim$estimate[-1]
))
```

The run ends with `r weight_based_sim$estimate[1]` classes, and the
share and the class means differ from those of the fixed fit by at most
`r round(heuristic_shift, 2)`. This is the case the heuristic was designed
for; the caveats above concern the interpretation of its result.

### Sparse finite mixtures

A sparse finite mixture fixes a generous upper bound $K$, permits empty
classes, and places the symmetric Dirichlet prior with a small concentration
$e_0$ on the weights, which favors emptying redundant classes
[@Rousseau2011; @FruehwirthSchnatter2019]. The default
`class_concentration = c(shape = 1, rate = 200)` is the gamma hyperprior
$e_0\sim\operatorname{Gamma}(1,200)$ with mean `0.005`. Under a fixed
$e_0$, the prior expected number of occupied classes among $N$ deciders is

\[
K\left[1-
\frac{\Gamma(Ke_0)\,\Gamma((K-1)e_0+N)}
     {\Gamma((K-1)e_0)\,\Gamma(Ke_0+N)}\right],
\]

```{r expected-classes, include=FALSE}
expected_classes <- function(K, e0, N) {
  K * (1 - exp(
    lgamma(K * e0) + lgamma((K - 1) * e0 + N) -
      lgamma((K - 1) * e0) - lgamma(K * e0 + N)
  ))
}
```

which translates $e_0$ into a statement about the number of classes. At the
prior mean $e_0 = 0.005$, with $K = 6$ and the 80 deciders of the simulated
data, it gives about `r round(expected_classes(6, 0.005, 80), 1)` occupied
classes: the prior expects close to a single occupied class. The refit sets
the upper bound to six classes for the two-class data.

```{r sparse}
sparse <- update(mixture, classes = 6, class_update = "sparse")
sparse_summary <- summary(sparse)
sparse_summary
sparse_occupancy <- latent_class_diagnostics(sparse)$occupancy
sparse_occupancy
sparse_sim <- recover_classes(sparse)
sparse_sim
```

```{r sparse-numbers, include=FALSE}
occupancy_share <- function(occupancy, classes) {
  sum(occupancy$probability[occupancy$n_classes %in% classes])
}
sparse_two <- occupancy_share(sparse_occupancy, 2)
sparse_table <- sparse_summary$posterior
sparse_ess <- min(sparse_table$ess_bulk, na.rm = TRUE)
sparse_slowest <- paste0(
  "`", sparse_table$variable[which.min(sparse_table$ess_bulk)], "`"
)
```

Although the prior favors a single class, the posterior assigns a
probability of `r round(100 * sparse_two)` percent to exactly two occupied
classes, and the class means are recovered as by the fixed model,
`r round(sparse_sim$estimate[3], 2)` and
`r round(sparse_sim$estimate[4], 2)` against `-1` and `2`. The smallest
bulk effective sample size, `r round(sparse_ess)`, belongs to
`r sparse_slowest`. Both $K$ and $e_0$ affect the posterior of the number of
occupied classes, so a sensitivity analysis should vary them. A less sparse
alternative is $e_0\sim\operatorname{Gamma}(2,4K)$
[@FruehwirthSchnatter2019]:

```{r sparse-sensitivity}
sparse_less <- update(
  mixture,
  classes = 6,
  class_update = "sparse",
  prior = list(class_concentration = c(shape = 2, rate = 24))
)
less_occupancy <- latent_class_diagnostics(sparse_less)$occupancy
less_occupancy
less_sim <- recover_classes(sparse_less)
less_sim
```

With a prior that empties classes less readily, the posterior of the number
of occupied classes spreads from `r min(less_occupancy$n_classes)` to
`r max(less_occupancy$n_classes)` and assigns exactly two classes a
probability of only `r round(100 * occupancy_share(less_occupancy, 2))`
percent. The class means are still recovered,
`r round(less_sim$estimate[3], 2)` and
`r round(less_sim$estimate[4], 2)`, because the extra classes contain
few deciders, but the estimated number of classes has changed with the
prior. Report the concentration prior together with the occupancy table.

### Dirichlet-process mixtures

For `class_update = "dirichlet_process"`, the precision $\alpha$ controls
the prior tendency to open new classes, with the default
$\alpha\sim\operatorname{Gamma}(2,4)$ of mean `0.5`. Conditional on a fixed
$\alpha$, the prior expects $\sum_{i=1}^{N}\alpha/(\alpha+i-1)$ occupied
classes among $N$ deciders, which for $\alpha = 0.5$ and 80 deciders is
about `r round(sum(0.5 / (0.5 + seq_len(80) - 1)), 1)`. **RprobitB**
updates $\alpha$ with the augmentation of @Escobar1995 and the allocations
with the algorithm of @Neal2000. `max_classes` caps the number of classes
the sampler may open, and `fit()` warns when the draws reach the cap. The
number of occupied classes mixes slowly, because a decider opens a new class
only rarely, so the refit runs longer chains than the reference fit, thinned
to keep the fitted object small.

```{r dirichlet}
dynamic <- update(
  mixture, class_update = "dirichlet_process", max_classes = 15,
  iterations = 30000, warmup = 15000, thin = 10
)
dynamic_summary <- summary(dynamic)
dynamic_summary
dynamic_occupancy <- latent_class_diagnostics(dynamic)$occupancy
dynamic_occupancy
dynamic_sim <- recover_classes(dynamic)
dynamic_sim
```

```{r dirichlet-numbers, include=FALSE}
dynamic_table <- dynamic_summary$posterior
dynamic_classes <- dynamic_table[dynamic_table$variable == "n_classes", ]
dynamic_maximum <- max(dynamic_occupancy$n_classes)
dynamic_cap <- dynamic$model$latent_classes$maximum
```

The occupancy distribution is wider than under the sparse finite prior: it
assigns a posterior probability of
`r round(100 * occupancy_share(dynamic_occupancy, 3:15))` percent to three
or more classes, against
`r round(100 * occupancy_share(sparse_occupancy, 3:15))` percent under the
sparse finite prior. This reflects the prior, which expects about three
occupied classes for 80 deciders and remains influential at this sample
size. The label-invariant comparison of the helper nevertheless recovers the
two true groups, `r round(dynamic_sim$estimate[3], 2)` and
`r round(dynamic_sim$estimate[4], 2)`, because the superfluous classes
contain only a few deciders. The `rhat` and the effective sample size of
`n_classes`, here `r sprintf("%.3f", dynamic_classes$rhat)` and
`r round(dynamic_classes$ess_bulk)`, decide whether the occupancy table is
reliable. The largest number of classes in any draw is `r dynamic_maximum`,
`r if (dynamic_maximum < dynamic_cap) "below" else "at"` the cap of
`r dynamic_cap`. The number of occupied classes, the concentration, and the
co-clustering probabilities do not depend on the class labels and are
therefore the safest summaries of such a fit; a class that exists in only a
small share of the draws describes a few deciders, not a group in the
population. Report the prior on $\alpha$ with the analysis, together with
the sensitivity to defensible alternatives.

## Ordered and ranked responses

Both mechanisms also work for ordered and ranked responses, which the
vignette [Model specification and variants][v02] introduces. They require a
panel, because a population distribution is estimated from deciders observed
repeatedly. The following demonstration uses a simulated panel of rankings:
100 deciders order three alternatives five times each, with a coefficient
that varies normally around `-1`.

```{r ranked-random}
ranked_random <- fit(
  rank ~ x | 0,
  choice_type = "ranked",
  random_effects = "x",
  n_deciders = 100,
  n_occasions = 5,
  n_alternatives = 3,
  dgp_parameters = list(beta = c(x = -1), Omega = matrix(0.3)),
  iterations = 4000,
  warmup = 2000,
  chains = 2,
  progress = FALSE
)
summary(ranked_random)
```

```{r ranked-random-deviation, include=FALSE}
ranked_table <- summary(ranked_random)$posterior
ranked_rows <- match(c("mu[x]", "Omega[x,x]"), ranked_table$variable)
ranked_deviation <- round(
  abs(ranked_table$mean - ranked_table$dgp)[ranked_rows] /
    ranked_table$sd[ranked_rows], 1
)
```

The posterior means of the population mean and variance deviate from their
true values by `r ranked_deviation[1]` and `r ranked_deviation[2]` posterior
standard deviations. Ordered responses accept the same arguments, and latent
classes are requested in the same way as above, with `latent_class_effects`
and `classes`.

## Further reading

The vignette [Posterior prediction][v04] uses the individual coefficients to
compute choice probabilities for each decider in the fit. The vignette
[Bayesian model evaluation][v05] shows how to decide whether random
coefficients improve a model.

[v01]: https://loelschlaeger.de/RprobitB/articles/v01_get_started.html
[v02]: https://loelschlaeger.de/RprobitB/articles/v02_model_variants.html
[v04]: https://loelschlaeger.de/RprobitB/articles/v04_prediction.html
[v05]: https://loelschlaeger.de/RprobitB/articles/v05_model_evaluation.html

## References
