---
title: "Posterior prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Posterior prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
link-citations: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A fitted choice model earns its keep by answering questions. What happens to
the support for a wind-power project if the compensation is doubled? Which
travelers are likely to book the faster train? Does the model still know
what it is doing when it meets people it has never seen? `predict()` in
**RprobitB** [@Oelschlaeger2026d] answers such questions as a posterior
predictive method: it evaluates the choice probabilities of every occasion
under each retained posterior draw and averages the result. Parameter
uncertainty is thereby carried through to the predictions, which also makes
them a natural basis for checking model fit [@Gelman1996]. This vignette
works through predictions for the population and for individual deciders,
uncertainty summaries, what-if scenarios in a stated choice experiment,
out-of-sample prediction of revealed choices, and residuals. The parts use
data sets of the **mlogit** package [@Croissant2020], the **choicedata**
package [@Oelschlaeger2026a], and the **MASS** package
[@VenablesRipley2002]. It builds on the fitted model of
`vignette("v01_get_started")` and puts the individual coefficients of
`vignette("v03_heterogeneity")` to work.

## Not everyone minds the price equally

The `Train` data of the **mlogit** package contain about a dozen choices
by each of 235 Dutch travelers between two hypothetical train trips that
differ in price, travel time, number of changes, and comfort class
[@BenAkiva1993]; as in `vignette("v01_get_started")`, the prices are
converted to euro and the travel times to hours.
`vignette("v01_get_started")` fits a model with one price coefficient for
everyone, and `vignette("v03_heterogeneity")` splits the travelers into two
classes. Here the price coefficient gets a normal random effect instead, so
every traveler has their own price sensitivity. Thinning keeps 50 of the
750 post-warmup draws, which is enough for stable posterior means and keeps
the predictions that follow fast.

```{r fit}
library(RprobitB)
set.seed(1)
data("Train", package = "mlogit")
Train$price_A <- Train$price_A / 100 / 2.20371
Train$price_B <- Train$price_B / 100 / 2.20371
Train$time_A <- Train$time_A / 60
Train$time_B <- Train$time_B / 60

model <- fit(
  choice ~ price + time + change + factor(comfort) | 0,
  data = Train,
  random_effects = "price",
  column_decider = "id",
  column_occasion = "choiceid",
  iterations = 1500,
  warmup = 750,
  thin = 15,
  chains = 2,
  save_individual_draws = TRUE,
  progress = FALSE
)
summary(model)
```

`mu[price]` and `Omega[price,price]` describe how the price coefficient
is spread across the population. The square root of its posterior mean is
`r round(sqrt(coef(model)["Omega[price,price]"]), 3)`, not far from the
size of the mean itself: price sensitivity clearly differs
between travelers, and the predictions below can put that to use. The
variance parameter is also the one that mixes worst, as its `rhat` shows.
Two short, heavily thinned chains are not enough for a real analysis of the
heterogeneity, but they are enough to demonstrate prediction.

## Predicting for the population

By default, `predict()` returns one row per choice occasion with the
identifiers, the most probable alternative in `.prediction`, and the
posterior mean probability of every alternative. These population
predictions integrate over the estimated distribution of the random
coefficients, so they hold for any traveler from the population, not just
for those in the data.

```{r population}
population <- predict(model)
head(population)
```

`uncertainty = TRUE` adds the posterior standard deviation and an
equal-tailed credible interval of every probability. These intervals express
uncertainty about the parameters, not the randomness of the choice itself.

```{r uncertainty}
head(predict(model, uncertainty = TRUE, level = 0.9))
```

## Predicting for the people you observed

A mixed model retains the posterior draws of every individual coefficient,
and `coef(level = "individual")` summarizes them. Once a traveler has
answered a dozen questions, we know quite a bit about their personal price
sensitivity, and `predict(type = "conditional")` uses that knowledge instead
of the population distribution. Conditional predictions exist only for
deciders that were part of the fit, and they are usually sharper because
they exploit the decider's own choice history. This partial pooling is the
classic argument for hierarchical Bayesian choice models [@Allenby1998;
@Huber2001].

```{r conditional}
head(coef(model, level = "individual"))
conditional <- predict(model, type = "conditional")
head(conditional)
```

How much does the personal information help? The share of correctly
predicted choices in the fitted data gives a first answer.

```{r accuracy}
observed <- model.frame(model)$choice
c(
  population = mean(population$.prediction == observed, na.rm = TRUE),
  conditional = mean(conditional$.prediction == observed, na.rm = TRUE)
)
```

A traveler's own history lifts the hit rate noticeably. The hit rate judges
the predictions at a single cut, a probability of one half. A ROC curve
compares them at every cut: as the threshold for predicting trip `B` moves
from one down to zero, it traces the share of `B` choices that are
predicted correctly against the share of `A` choices that are wrongly
predicted as `B`, and the area under the curve is the probability that a
randomly chosen `B` occasion receives a higher `B` probability than a
randomly chosen `A` occasion. The `plotROC` package [@Sachs2017] draws the
curves with `ggplot2` [@Wickham2016].

```{r roc, fig.width=6, fig.height=4}
library(ggplot2)
library(plotROC)
roc_data <- rbind(
  data.frame(
    prediction = "population", chose_B = as.integer(observed == "B"),
    probability = population$probability_B
  ),
  data.frame(
    prediction = "conditional", chose_B = as.integer(observed == "B"),
    probability = conditional$probability_B
  )
)
roc_data$prediction <- factor(
  roc_data$prediction, levels = c("population", "conditional")
)
roc_plot <- ggplot(
  roc_data, aes(d = chose_B, m = probability, color = prediction)
) +
  geom_roc(n.cuts = 0) +
  style_roc()
roc_plot
```

```{r roc-auc, include=FALSE}
roc_auc <- calc_auc(roc_plot)$AUC
```

The conditional curve lies above the population curve, and the area under
it rises from `r round(roc_auc[1], 2)` to `r round(roc_auc[2], 2)`.

## What if the compensation were doubled?

`newdata` accepts a data frame in the layout of the fitted data, with or
without the response column. A scenario analysis thus comes down to editing
a data frame, which is how stated choice experiments are typically put to
use.

Near Setskog in Norway, 308 residents were asked six times to choose between
two plans for a proposed wind-power project and the status quo without the
project, alternative `1` [@Dugstad2024]. The plans varied the number and
height of the turbines, the routing of the power line, and an annual
reduction in municipal taxes offered as compensation. The study also
measured each respondent's collective psychological ownership of the
affected area, a score of how strongly they feel that the landscape belongs
to the residents. That score is the point of the original study
[@Dugstad2024], which reports that a stronger feeling of ownership makes
residents less willing to accept a project. The estimates below say the
same. The score does not vary across alternatives and therefore enters with
alternative-specific coefficients.

```{r wind}
data("wind_power_choice", package = "choicedata")
wind <- fit(
  choice ~ turbines + height + powerline + compensation |
    psychological_ownership,
  data = wind_power_choice,
  column_decider = "respondent",
  column_occasion = "occasion",
  iterations = 600,
  warmup = 300,
  chains = 2,
  progress = FALSE
)
coef(wind)[c("beta[compensation]", "beta[psychological_ownership_2]")]
```

Compensation makes a plan more attractive, and residents with a stronger
feeling of ownership are less willing to leave the status quo. How much
money does that feeling stand for? Since the compensation is money, it is
the natural reference for `interpret()`, and the ownership score is
standardized, so the result is the annual tax reduction that offsets one
standard deviation more ownership:

```{r wind-interpret}
ownership_value <- interpret(
  wind,
  reference = "compensation",
  effects = c("psychological_ownership_2", "psychological_ownership_3")
)
ownership_value
```

For either plan, a resident who feels one standard deviation more strongly
that the area belongs to the residents would need about
`r round(mean(ownership_value$mean))` euro more compensation per year to be
equally likely to accept it. That is of the order of the largest reduction
on offer in the experiment, `r round(max(wind_power_choice$compensation_2))`
euro, and shows why money alone goes only so far here.

So what would happen if the municipality doubled the compensation? The
scenario below doubles it for both plans in the first four choice tasks and
compares the probability of the status quo before and after.

```{r scenario}
scenario <- model.frame(wind)[1:4, ]
scenario$choice <- NULL
scenario$compensation_2 <- 2 * scenario$compensation_2
scenario$compensation_3 <- 2 * scenario$compensation_3
cbind(
  before = predict(wind)$probability_1[1:4],
  after = predict(wind, newdata = scenario)$probability_1
)
```

The status quo loses between five and nine percentage points in every task.
Money helps, but it is not the only thing that matters to the neighbors of a
wind farm.

## Do the chess players behave as the model expects?

Predictive performance is best judged on deciders that played no part in
the estimation; this is the principle behind out-of-sample and leave-one-out
model assessment [@Vehtari2017]. The `lichess_berserk_choice`
data offer a large playground for it. In an arena tournament on Lichess, a
player may go Berserk at the start of a game: their clock is halved, and a
win earns one extra tournament point. Players on a winning streak collect
double points, so for them a loss is particularly costly. The data record
this decision for 5852 players in the Lichess Yearly Rapid Arena of April
2026, game by game, together with the playing color, the player's rating,
the rating difference to the opponent, the remaining tournament time, and
whether the player was on a streak. Unlike the other data sets here, these
are revealed choices from a live tournament and have not been analyzed
before, so there is nothing to compare the estimates with except the
hold-out games below. All of these are game-specific and
constant across the two alternatives, so they enter as type `B` covariates
with a coefficient for the alternative `TRUE`, for example
`beta[rating_TRUE]`; logical covariates appear as dummy variables such as
`streakTRUE`. The fit below keeps the first 300 players and lets
`train_test()` split their games. The function is re-exported from
**choicedata** and splits by decider, not by row: `test_number = 60` puts the games of 60 players into the test set and
everything else into the training set, so no player contributes to both.
`test_proportion` takes a share instead of a count, and `by_occasion = TRUE`
splits within deciders, which is what you want when the question is
forecasting later choices of known deciders rather than choices of new
ones.

```{r lichess}
data("lichess_berserk_choice", package = "choicedata")
players <- unique(lichess_berserk_choice$deciderID)
first_players <- lichess_berserk_choice$deciderID %in% players[1:300]
split <- train_test(lichess_berserk_choice[first_players, ], test_number = 60)
berserk <- fit(
  berserk ~ 0 | white + rating + ratingDifference + minutesRemaining + streak,
  data = split$train,
  column_occasion = "occasionID",
  iterations = 1000,
  warmup = 500,
  chains = 2,
  progress = FALSE
)
coef(berserk)
```

Stronger players go Berserk more often, players on a streak less often, and
everyone more often towards the end of the tournament, when little time
remains to make up for a lost game. How large are these effects in
probabilities? `interpret(type = "mea")` differentiates the Berserk
probability at the average game:

```{r berserk-interpret}
berserk_effects <- interpret(berserk, type = "mea")
berserk_effects
```

```{r berserk-numbers, include=FALSE}
berserk_effect <- function(covariate) {
  row <- berserk_effects$covariate == covariate &
    berserk_effects$alternative == "TRUE"
  berserk_effects$mean[row]
}
```

A player rated 100 points higher is about
`r round(100 * 100 * berserk_effect("rating"), 1)` percentage points more
likely to go Berserk, and every hour that remains in the tournament lowers
the probability by about
`r round(-100 * 60 * berserk_effect("minutesRemaining"), 1)` percentage
points. How well does this carry over to the 60 unseen players? Accuracy
alone is a blunt instrument here, because always predicting `FALSE` already
scores whatever share of the games went without Berserk.

```{r holdout}
holdout_prediction <- predict(berserk, newdata = split$test)
holdout_choice <- as.character(split$test$berserk)
c(
  accuracy = mean(holdout_prediction$.prediction == holdout_choice),
  share_berserk = mean(split$test$berserk)
)
```

The value of the model lies in its probabilities rather than in its hit
rate. A calibration table sorts the hold-out games by their predicted
Berserk probability and compares it with the observed Berserk rate in each
group; groups with fewer than 50 games are too small to judge and dropped.

```{r calibration}
predicted <- holdout_prediction$probability_TRUE
bins <- cut(predicted, breaks = seq(0, 1, by = 0.1))
calibration <- data.frame(
  games = as.vector(table(bins)),
  predicted = as.vector(tapply(predicted, bins, mean)),
  observed = as.vector(tapply(split$test$berserk, bins, mean)),
  row.names = levels(bins)
)
round(calibration[calibration$games >= 50, ], 2)
```

A calibration plot says the same thing at a glance. Points on the diagonal
mean that the predicted probability is the observed rate; points below it
mean the model predicts more Berserk than happens.

```{r calibration-plot, fig.width=5, fig.height=5}
large <- calibration[calibration$games >= 50, ]
plot(
  large$predicted, large$observed,
  xlim = c(0, 1), ylim = c(0, 1), pch = 19,
  cex = 0.5 + 2 * large$games / max(large$games),
  xlab = "predicted Berserk probability",
  ylab = "observed Berserk rate"
)
abline(0, 1, lwd = 2, col = "grey50")
```

The point sizes are proportional to the number of games in a group. The
observed rate rises with the predicted probability, which is what a useful
model must deliver: where it sees a low risk appetite, few players go
Berserk, and where it predicts something close to a coin flip, more than
half of them do. The points sit close to the diagonal in the middle groups
and above it at the top, so the model ranks the games well and is, if
anything, a little too cautious about the games it already considers risky.

## Residuals

`residuals()` returns the observed choice indicators minus the posterior
mean probabilities, one row per choice occasion and one column per
alternative. The rows of observed occasions sum to zero, and occasions with
a missing response yield `NA`. For ranked data, the indicator marks the
first-ranked alternative.

```{r residuals}
model_residuals <- residuals(model)
head(model_residuals)
colMeans(model_residuals, na.rm = TRUE)
```

Average residuals near zero say that the predicted choice shares match the
observed ones. The more revealing use is to group them. Averaging by decider
shows whose choices the model reproduces and whose it does not:

```{r residuals-decider}
by_decider <- tapply(
  model_residuals[, "A"], model.frame(model)$id, mean, na.rm = TRUE
)
round(quantile(by_decider, c(0, 0.25, 0.5, 0.75, 1)), 3)
```

Half of the travelers have an average residual within about a tenth of
zero, while the extremes reach a third: for those travelers the model
predicts trip `A` too often or too rarely across all their questions.
Grouping by a covariate is the second useful cut, here by the price of trip
`A` in four groups of equal size:

```{r residuals-covariate}
price_group <- cut(
  model.frame(model)$price_A,
  breaks = quantile(model.frame(model)$price_A, seq(0, 1, 0.25)),
  include.lowest = TRUE
)
round(tapply(model_residuals[, "A"], price_group, mean, na.rm = TRUE), 3)
```

No group stands out, so the linear price term is doing its job. A systematic
pattern here, for example positive residuals at both ends, would argue for a
nonlinear term or an unmodeled preference class.

## Marginal effects

How much does one euro more change the probability of booking a trip?
Coefficients do not answer that directly, because the probit link is not
linear. `interpret()` differentiates the predicted probabilities instead.
`type = "mea"` evaluates the derivative for one occasion whose covariates
equal the observed averages, which the column `at` reports; `type = "ame"`
evaluates it for every observed occasion and averages, which weighs the
occasions as they occur in the data and is usually the more relevant
summary. Both use every posterior draw, so they come with credible
intervals.

```{r marginal-effects}
interpret(model, type = "mea")
average_effects <- interpret(model, type = "ame")
average_effects
```

The `at` argument replaces the average of selected covariates. Naming the
alternative matters: a price that is common to both trips cancels in the
utility difference and leaves the effects unchanged, while a much cheaper
alternative B pushes the choice towards certainty, where one euro more
hardly moves the probability at all.

```{r marginal-effects-at}
interpret(model, type = "mea", at = c(price_A = 30, price_B = 10))
```

With two alternatives, a change in the attributes of one trip shifts
probability to the other, so the two rows of each covariate mirror each
other. Averaged over the observed trips, one euro more lowers the
probability of a trip by about
`r round(-100 * average_effects$mean[average_effects$covariate == "price"][1])`
percentage points and one hour more by about
`r round(-100 * average_effects$mean[average_effects$covariate == "time"][1])`
points, in line with the compensations in `vignette("v01_get_started")`.

## Ordered responses predict a distribution over levels

For an ordered model, `predict()` returns the probability of every level
rather than of every alternative. The smoking model of
`vignette("v02_model_variants")` predicts how often a student smokes from
their age and exercise habits.

```{r ordered-prediction}
data("survey", package = "MASS")
smoking <- fit(
  Smoke ~ Age + Exer | 0,
  data = survey,
  alternatives = c("Never", "Occas", "Regul", "Heavy"),
  choice_type = "ordered",
  column_decider = NULL,
  iterations = 1500,
  warmup = 750,
  chains = 2,
  progress = FALSE
)
head(predict(smoking))
```

Every row carries four probabilities that sum to one, and `.prediction`
names the most probable level. Because four out of five students never
smoke, that level wins almost everywhere, and the model earns its keep
through the probabilities rather than through the predicted level. The
residuals have one column per level and are read the same way as above.

```{r ordered-residuals}
colMeans(residuals(smoking), na.rm = TRUE)
```

A scenario works as before: `newdata` takes a data frame in the layout of
the fitted data. Ten years of age shift the levels as the coefficients
promised.

```{r ordered-scenario}
students <- model.frame(smoking)[1:3, ]
students$Smoke <- NULL
older <- students
older$Age <- older$Age + 10
cbind(
  before = predict(smoking, newdata = students)$probability_Never,
  after = predict(smoking, newdata = older)$probability_Never
)
```

## Where to go next

Predictions say what a model expects; they do not say whether the model
deserves to be believed over another one. That comparison, on the same
decider-level scale that the hold-out check uses here, is the subject of
`vignette("v05_model_evaluation")`.

## References
