---
title: "Posterior prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Posterior prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
link-citations: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
  fig.align = "center"
)
```

A fitted choice model answers questions about choice probabilities: which of
two train trips a traveler is likely to book, how the support for a
wind-power project changes with the compensation, and how well the model
predicts the choices of deciders that were not used for estimation. In the
probit model introduced in the vignette [Get started with RprobitB][v01],
the choice probability of an alternative is the probability that its latent
utility exceeds the utilities of the other alternatives, and it depends on
the covariates of the occasion and on the parameters. `predict()` evaluates
these probabilities under every retained posterior draw and averages them,
so that the predictions account for the posterior uncertainty about the
parameters [@Gelman1996]. This vignette covers predictions for the
population and for individual deciders, scenarios, out-of-sample
prediction, residuals, and marginal effects. The examples use data sets of
the **mlogit** package [@Croissant2020], the **choicedata** package
[@Oelschlaeger2026a], and the **MASS** package [@VenablesRipley2002].

## A random price coefficient

The `Train` data of the **mlogit** package contain about a dozen choices by
each of 235 Dutch travelers between two hypothetical train trips that differ
in price, travel time, number of changes, and comfort class [@BenAkiva1993].
As in the vignette [Get started with RprobitB][v01], the prices are converted
to euro and the travel times to hours. That vignette fits one price
coefficient for all travelers. Here the price coefficient is a normal random
effect, as introduced in the vignette
[Modeling preference heterogeneity][v03]: every traveler has an individual
price coefficient, drawn from a normal population distribution whose mean
and variance are estimated. The individual draws are saved because the
conditional predictions below use them. Thinning keeps 50 of the 750
post-warmup draws of each chain, 100 draws in total, which is enough for
stable posterior means and keeps the predictions below fast.

```{r fit}
library(RprobitB)
set.seed(1)
data("Train", package = "mlogit")
Train$price_A <- Train$price_A / 100 / 2.20371
Train$price_B <- Train$price_B / 100 / 2.20371
Train$time_A <- Train$time_A / 60
Train$time_B <- Train$time_B / 60
model <- fit(
  choice ~ price + time + change + factor(comfort) | 0,
  data = Train,
  random_effects = "price",
  column_decider = "id",
  column_occasion = "choiceid",
  iterations = 1500,
  warmup = 750,
  thin = 15,
  chains = 2,
  save_individual_draws = TRUE,
  progress = FALSE
)
summary(model)
```

`mu[price]` and `Omega[price,price]` are the mean and the variance of the
price coefficient across the population. The square root of the posterior
mean of the variance is
`r round(sqrt(coef(model)["Omega[price,price]"]), 3)`, against a posterior
mean of `r round(coef(model)["mu[price]"], 3)` for `mu[price]`: price
sensitivity differs between travelers, which the predictions below take into
account.

## Population predictions

By default, `predict()` returns one row per choice occasion with the
identifiers, the most probable alternative in `.prediction`, and the
posterior mean probability of every alternative. These population
predictions integrate over the estimated population distribution of the
random coefficient, so they apply to any traveler from the population, not
only to those in the data.

```{r population}
population <- predict(model)
head(population)
```

`uncertainty = TRUE` adds the posterior standard deviation and an
equal-tailed credible interval of every probability, here at the 90% level.
The intervals reflect the posterior uncertainty about the parameters, not
the randomness of the choice.

```{r uncertainty}
head(predict(model, uncertainty = TRUE, level = 0.9))
```

## Conditional predictions for fitted deciders

The dozen choices of a traveler are informative about their individual
price coefficient. `coef(level = "individual")` returns the posterior mean
of every traveler's coefficient, and `predict(type = "conditional")` uses
the individual draws instead of the population distribution. Conditional
predictions exist only for the deciders in the fitted data, and they are
usually sharper because they use the decider's own choices. This partial
pooling is the main argument for hierarchical Bayesian choice models
[@Allenby1998; @Huber2001].

```{r conditional}
head(coef(model, level = "individual"))
conditional <- predict(model, type = "conditional")
head(conditional)
```

The hit rate, the share of correctly predicted choices in the fitted data,
is one measure of the gain:

```{r accuracy}
observed <- model.frame(model)$choice
hit_rate <- c(
  population = mean(population$.prediction == observed, na.rm = TRUE),
  conditional = mean(conditional$.prediction == observed, na.rm = TRUE)
)
hit_rate
```

The traveler's own choices `r if (diff(hit_rate) > 0) "raise" else "lower"`
the hit rate from `r round(hit_rate[["population"]], 2)` to
`r round(hit_rate[["conditional"]], 2)`. The hit rate evaluates the
predictions at a single threshold, a probability of one half. A ROC curve
compares them at every threshold: as the threshold for predicting trip `B`
decreases from one to zero, the curve plots the share of `B` choices
predicted correctly against the share of `A` choices wrongly predicted as
`B`. The area under the curve is the probability that a randomly chosen `B`
occasion receives a higher `B` probability than a randomly chosen `A`
occasion. The **plotROC** package [@Sachs2017] draws the curves with
**ggplot2** [@Wickham2016].

```{r roc, fig.width=6, fig.height=4}
library(ggplot2)
library(plotROC)
roc_data <- rbind(
  data.frame(
    prediction = "population", chose_B = as.integer(observed == "B"),
    probability = population$probability_B
  ),
  data.frame(
    prediction = "conditional", chose_B = as.integer(observed == "B"),
    probability = conditional$probability_B
  )
)
roc_data$prediction <- factor(
  roc_data$prediction, levels = c("population", "conditional")
)
roc_plot <- ggplot(
  roc_data, aes(d = chose_B, m = probability, color = prediction)
) +
  geom_roc(n.cuts = 0) +
  style_roc()
roc_plot
```

```{r roc-auc, include=FALSE}
roc_auc <- calc_auc(roc_plot)$AUC
```

The conditional curve lies above the population curve, and the area under
the curve rises from `r round(roc_auc[1], 2)` to `r round(roc_auc[2], 2)`.

## Scenario analysis in a stated choice experiment

A scenario predicts the choice probabilities for modified attributes, the
typical use of a stated choice experiment. Near Setskog in Norway, 308
residents were asked six times to choose between two plans for a proposed
wind-power project and the status quo without the project, alternative `1`
[@Dugstad2024]. The plans varied the number and height of the turbines, the
routing of the power line, and an annual reduction in municipal taxes
offered as compensation. The study also measured each respondent's
collective psychological ownership of the affected area, a standardized
score of how strongly they feel that the landscape belongs to the residents.
The score does not vary across alternatives and therefore enters the second
part of the formula, which gives it one coefficient per plan relative to the
status quo:

```{r wind-formula}
wind_formula <- choice ~ turbines + height + powerline + compensation |
  psychological_ownership
```

```{r wind}
data("wind_power_choice", package = "choicedata")
wind <- fit(
  formula = wind_formula,
  data = wind_power_choice,
  column_decider = "respondent",
  column_occasion = "occasion",
  iterations = 10000,
  warmup = 5000,
  thin = 5,
  chains = 2,
  progress = FALSE
)
coef(wind)[c("beta[compensation]", "beta[psychological_ownership_2]")]
```

```{r wind-signs, include=FALSE}
wind_draws <- posterior::as_draws_matrix(wind)
wind_positive <- sprintf(
  "%.2f", mean(wind_draws[, "beta[compensation]"] > 0)
)
wind_negative <- sprintf(
  "%.2f", mean(wind_draws[, "beta[psychological_ownership_2]"] < 0)
)
```

The posterior probability that the compensation coefficient is positive is
`r wind_positive`, and that the ownership coefficient of the first plan is
negative `r wind_negative`: compensation makes a plan more attractive, and
residents with a stronger feeling of ownership are less willing to leave the
status quo. What would happen if the municipality doubled the compensation?
`newdata` accepts a data frame in the layout of the fitted data, with or
without the response column. The scenario below doubles the compensation of
both plans in the first four choice tasks and compares the probability of
the status quo before and after.

```{r scenario}
scenario <- model.frame(wind)[1:4, ]
scenario$choice <- NULL
scenario$compensation_2 <- 2 * scenario$compensation_2
scenario$compensation_3 <- 2 * scenario$compensation_3
status_quo <- cbind(
  before = predict(wind)$probability_1[1:4],
  after = predict(wind, newdata = scenario)$probability_1
)
status_quo
```

```{r scenario-numbers, include=FALSE}
status_quo_loss <- 100 * (status_quo[, "before"] - status_quo[, "after"])
```

The status quo loses between `r round(min(status_quo_loss))` and
`r round(max(status_quo_loss))` percentage points in the four tasks.
Compensation matters, but it is not the only factor in the residents'
acceptance.

## Out-of-sample prediction and calibration

Predictive performance is best judged on deciders that were not used for
estimation [@Vehtari2017]. In an arena tournament on the chess server
Lichess, a player may go Berserk at the start of a game: the clock is
halved, and a win earns one extra tournament point. Players on a winning
streak collect double points, so a loss is more costly for them. The
`lichess_berserk_choice` data of the **choicedata** package record this
decision for 5852 players in the Lichess Yearly Rapid Arena of April 2026,
game by game, together with the playing color, the player's rating, the
rating difference to the opponent, the remaining tournament time, and
whether the player was on a streak. All covariates describe the game and
are constant across the two alternatives, so they enter the second part of
the formula, which gives each of them one coefficient for the alternative
`TRUE` relative to `FALSE`, for example `beta[rating_TRUE]`. Logical
covariates appear as dummy variables such as `streakTRUE`.

```{r berserk-formula}
berserk_formula <- berserk ~ 0 | white + rating + ratingDifference +
  minutesRemaining + streak
```

The fit uses the games of the first 300 players. `train_test()`, which
**RprobitB** re-exports from **choicedata**, splits them by decider, not by
row: `test_number = 60` puts all games of 60 players into the test set and
the games of the other players into the training set, so no player
contributes to both.

```{r lichess}
data("lichess_berserk_choice", package = "choicedata")
players <- unique(lichess_berserk_choice$deciderID)
first_players <- lichess_berserk_choice$deciderID %in% players[1:300]
split <- train_test(lichess_berserk_choice[first_players, ], test_number = 60)
berserk <- fit(
  formula = berserk_formula,
  data = split$train,
  column_occasion = "occasionID",
  iterations = 1000,
  warmup = 500,
  chains = 2,
  progress = FALSE
)
coef(berserk)
```

```{r berserk-signs, include=FALSE}
berserk_draws <- posterior::as_draws_matrix(berserk)
berserk_sign <- c(
  rating = mean(berserk_draws[, "beta[rating_TRUE]"] > 0),
  streak = mean(berserk_draws[, "beta[streakTRUE_TRUE]"] < 0),
  minutes = mean(berserk_draws[, "beta[minutesRemaining_TRUE]"] < 0)
)
berserk_sign <- setNames(sprintf("%.2f", berserk_sign), names(berserk_sign))
```

The posterior probability that the rating coefficient is positive is
`r berserk_sign[["rating"]]`, that the streak coefficient is negative
`r berserk_sign[["streak"]]`, and that the coefficient of the remaining time
is negative `r berserk_sign[["minutes"]]`: stronger players go Berserk more
often, players on a streak less often, and all players more often towards
the end of the tournament. How well does the model predict the games of the
60 players in the test set? `newdata` takes the test set as it is, and the
predicted alternative is compared with the observed one.

```{r holdout}
holdout_prediction <- predict(berserk, newdata = split$test)
holdout_choice <- as.character(split$test$berserk)
holdout_accuracy <- c(
  accuracy = mean(holdout_prediction$.prediction == holdout_choice),
  share_berserk = mean(split$test$berserk)
)
holdout_accuracy
```

The hit rate is `r round(100 * holdout_accuracy[["accuracy"]])` percent,
but the rule that never predicts Berserk would already reach
`r round(100 * (1 - holdout_accuracy[["share_berserk"]]))` percent, so the
hit rate alone is a weak criterion for an unbalanced binary response. A
calibration analysis assesses the predicted probabilities instead. It
groups the hold-out games by predicted Berserk probability in intervals of
width 0.1 and compares the mean predicted probability with the observed
Berserk rate in each group; groups with fewer than 50 games are dropped.

```{r calibration}
predicted <- holdout_prediction$probability_TRUE
bins <- cut(predicted, breaks = seq(0, 1, by = 0.1))
calibration <- data.frame(
  games = as.vector(table(bins)),
  predicted = as.vector(tapply(predicted, bins, mean)),
  observed = as.vector(tapply(split$test$berserk, bins, mean)),
  row.names = levels(bins)
)
large <- calibration[calibration$games >= 50, ]
round(large, 2)
```

The calibration plot shows the same table. Points on the diagonal mean that
the predicted probability equals the observed rate, and the point sizes are
proportional to the number of games in a group.

```{r calibration-plot, fig.width=5, fig.height=5}
plot(
  large$predicted, large$observed,
  xlim = c(0, 1), ylim = c(0, 1), pch = 19,
  cex = 0.5 + 2 * large$games / max(large$games),
  xlab = "predicted Berserk probability",
  ylab = "observed Berserk rate"
)
abline(0, 1, lwd = 2, col = "grey50")
```

```{r calibration-numbers, include=FALSE}
percent <- function(x) round(100 * x)
calibration_gap <- large$observed - large$predicted
calibration_worst <- which.max(abs(calibration_gap))
calibration_direction <- if (calibration_gap[calibration_worst] > 0) {
  "underpredicts"
} else {
  "overpredicts"
}
calibration_rise <- if (all(diff(large$observed) > 0)) {
  "from every group to the next"
} else {
  "between the extreme groups, but not from every group to the next"
}
```

From the group with the lowest predictions to the group with the highest,
the mean predicted probability rises from `r percent(large$predicted[1])`
to `r percent(large$predicted[nrow(large)])` percent and the observed rate
from `r percent(large$observed[1])` to
`r percent(large$observed[nrow(large)])` percent. The observed rate rises
`r calibration_rise`, so the model orders the games by their Berserk rate.
The largest gap between the observed rate and the predicted probability,
`r percent(max(abs(calibration_gap)))` percentage points, occurs in the
group `r rownames(large)[calibration_worst]`, where the model
`r calibration_direction` Berserk.

## Residuals

`residuals()` returns the observed choice indicators minus the posterior
mean probabilities, one row per choice occasion and one column per
alternative. The rows of observed occasions sum to zero, and occasions with
a missing response yield `NA`. A single residual is uninformative, because
the indicator is zero or one while the probability lies in between, so the
residuals are averaged over groups of occasions. The average residual per
traveler shows whose choices the model reproduces:

```{r residuals}
model_residuals <- residuals(model)
head(model_residuals)
by_decider <- tapply(
  model_residuals[, "A"], model.frame(model)$id, mean, na.rm = TRUE
)
decider_quantiles <- quantile(by_decider, c(0, 0.25, 0.5, 0.75, 1))
round(decider_quantiles, 3)
```

```{r residuals-numbers, include=FALSE}
decider_text <- sprintf("%.2f", decider_quantiles)
```

Half of the travelers have an average residual between
`r decider_text[2]` and `r decider_text[4]`, and the extremes are
`r decider_text[1]` and `r decider_text[5]`: for these travelers the model
predicts trip `A` too often or too rarely across all their questions.
Grouping by a covariate checks the functional form instead, here by the
price of trip `A` in four groups of equal size:

```{r residuals-covariate}
price_group <- cut(
  model.frame(model)$price_A,
  breaks = quantile(model.frame(model)$price_A, seq(0, 1, 0.25)),
  include.lowest = TRUE
)
price_residuals <- tapply(
  model_residuals[, "A"], price_group, mean, na.rm = TRUE
)
round(price_residuals, 3)
```

The group averages lie between `r sprintf("%.3f", min(price_residuals))`
and `r sprintf("%.3f", max(price_residuals))`. A systematic pattern, for
example positive residuals at both ends, would indicate a nonlinear price
effect or an unmodeled preference class.

## Marginal effects

How much does one euro more change the probability of booking a trip? The
coefficients do not answer this directly, because the probit link is
nonlinear: the same change of a covariate moves the probability most where
the alternatives are close in utility and little where one alternative
dominates. `interpret()` therefore differentiates the predicted
probabilities numerically. `type = "mea"` evaluates the derivative for one
occasion whose covariates equal the observed averages, reported in the
column `at`. `type = "ame"` evaluates the derivative for every observed
occasion and averages, which weights the occasions as they occur in the data
and is usually the more relevant summary. Both use every posterior draw and
therefore come with credible intervals.

```{r marginal-effects}
interpret(model, type = "mea")
average_effects <- interpret(model, type = "ame")
average_effects
```

With two alternatives, a change in the attributes of one trip shifts
probability to the other, so the two rows of each covariate have equal size
and opposite signs. Averaged over the observed occasions, one euro more
lowers the probability of a trip by about
`r round(-100 * average_effects$mean[average_effects$covariate == "price"][1])`
percentage points and one hour more by about
`r round(-100 * average_effects$mean[average_effects$covariate == "time"][1])`
percentage points. The `at` argument replaces the averages of selected
covariates. A much cheaper trip `B` moves its probability close to one,
where one euro more changes it only marginally:

```{r marginal-effects-at}
interpret(model, type = "mea", at = c(price_A = 30, price_B = 10))
```

## Predictions for ordered responses

An ordered model compares one latent utility per occasion with increasing
thresholds that partition it into the levels of the response, as the
vignette [Model specification and variants][v02] describes. For such a
model, `predict()` returns the probability of every level. The smoking model
of that vignette predicts how often a student smokes from their age and
exercise habits.

```{r ordered-prediction}
data("survey", package = "MASS")
smoking <- fit(
  Smoke ~ Age + Exer | 0,
  data = survey,
  alternatives = c("Never", "Occas", "Regul", "Heavy"),
  choice_type = "ordered",
  column_decider = NULL,
  iterations = 1500,
  warmup = 750,
  chains = 2,
  progress = FALSE
)
smoking_prediction <- predict(smoking)
head(smoking_prediction)
```

```{r ordered-prediction-numbers, include=FALSE}
never_predicted <- sum(smoking_prediction$.prediction == "Never")
never_text <- if (never_predicted == nrow(smoking_prediction)) {
  sprintf("every one of the %d students", never_predicted)
} else {
  sprintf("%d of the %d students", never_predicted, nrow(smoking_prediction))
}
```

Every row has four probabilities that sum to one, and `.prediction` names
the most probable level. Because
`r round(100 * mean(survey$Smoke == "Never", na.rm = TRUE))` percent of the
students never smoke, this level is the most probable for `r never_text`,
and the probabilities are more informative than the predicted level. A
scenario works as before. Ten more years of age shift the probability of
never smoking of the first three students:

```{r ordered-scenario}
students <- model.frame(smoking)[1:3, ]
students$Smoke <- NULL
older <- students
older$Age <- older$Age + 10
age_scenario <- cbind(
  before = predict(smoking, newdata = students)$probability_Never,
  after = predict(smoking, newdata = older)$probability_Never
)
age_scenario
```

```{r ordered-scenario-numbers, include=FALSE}
age_shift <- 100 * (age_scenario[, "after"] - age_scenario[, "before"])
```

The probability `r if (all(age_shift > 0)) "rises" else "changes"` by
`r round(min(abs(age_shift)))` to `r round(max(abs(age_shift)))` percentage
points, in the direction implied by the negative age coefficient.

## Further reading

Predictions describe what a model expects, not whether it is better than
another model. That comparison, on the same decider-level scale as the
hold-out check here, is the subject of the vignette
[Bayesian model evaluation][v05].

[v01]: https://loelschlaeger.de/RprobitB/articles/v01_get_started.html
[v02]: https://loelschlaeger.de/RprobitB/articles/v02_model_variants.html
[v03]: https://loelschlaeger.de/RprobitB/articles/v03_heterogeneity.html
[v05]: https://loelschlaeger.de/RprobitB/articles/v05_model_evaluation.html

## References
