---
title: "Bayesian model evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian model evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
link-citations: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
  fig.align = "center"
)
```

Which of two models should you believe? **RprobitB** [@Oelschlaeger2026d] offers
three complementary answers: the widely applicable information criterion (WAIC),
Pareto-smoothed importance sampling leave-one-out cross-validation
(PSIS-LOO), and Bayes factors. The first two estimate how well a model would
predict new data, using the posterior log-likelihood draws. The third
compares the marginal likelihoods of two models, that is, how well each model
predicted the data before seeing them [@Watanabe2010;
@Vehtari2017; @Kass1995]. In panel data, the
log-likelihood is evaluated per decider, so the repeated choices of one
person count as one independent unit. The examples use data sets of the
**AER** package [@Kleiber2008], the **choicedata** package
[@Oelschlaeger2026a], the **MASS** package [@VenablesRipley2002], and the
**mlogit** package [@Croissant2020]. Fitting and reading a model is
the subject of `vignette("v01_get_started")`, the specifications compared
here come from `vignette("v02_model_variants")` and
`vignette("v03_heterogeneity")`, and what they predict is the subject of
`vignette("v04_prediction")`.

## Does income change how people travel?

In 1987, 210 travelers between Sydney and Melbourne reported which of four
modes they had taken: air, train, bus, or car. The `TravelMode` data of the
**AER** package [@Kleiber2008] are in long format with one row per mode.
They are a standard benchmark for discrete choice models, used among others
in Greene's econometrics textbook [@Greene2003b], where higher income is
likewise found to favor the faster and more expensive modes. Terminal
waiting time (`wait`), in-vehicle cost (`vcost`), and travel time vary
across modes; household income and the size of the traveling party belong
to the traveler. The choice indicator is coded as `"yes"` and `"no"`, which
`fit()` expects as a logical or `0`/`1` indicator, and the cost and the
income are converted from Australian dollars to euro. Cost and time surely
matter, but does income also tilt travelers towards particular modes? The
full model lets income and party size shift the utilities of the modes
relative to the base alternative, the reduced model omits them. Thinning
keeps 250 of the 2000 post-warmup draws, plenty for the information criteria
and short enough for the bridge sampling below.

```{r fit}
library(RprobitB)
set.seed(1)
data("TravelMode", package = "AER")
TravelMode$choice <- TravelMode$choice == "yes"
TravelMode$vcost <- TravelMode$vcost / 1.6196
TravelMode$income <- TravelMode$income / 1.6196

full_model <- fit(
  choice ~ wait + vcost + travel | income + size,
  data = TravelMode,
  format = "long",
  column_decider = "individual",
  column_alternative = "mode",
  iterations = 2000,
  warmup = 1000,
  thin = 8,
  chains = 2,
  progress = FALSE
)
reduced_model <- fit(
  choice ~ wait + vcost + travel | 0,
  data = TravelMode,
  format = "long",
  column_decider = "individual",
  column_alternative = "mode",
  iterations = 2000,
  warmup = 1000,
  thin = 8,
  chains = 2,
  progress = FALSE
)
```

`logLik()` evaluates the log-likelihood at the posterior mean parameters. Its
`df` attribute counts the free parameters after normalization, and `nobs`
counts the independent likelihood units, here the 210 travelers.

```{r loglik}
logLik(full_model)
logLik(reduced_model)
```

**RprobitB** defines no `AIC()` or `BIC()` methods of its own. Because
`logLik()` returns a standard `logLik` object, the generics of R's `stats`
package [@RCoreTeam2026] compute both criteria directly.

```{r classical-criteria}
AIC(full_model, reduced_model)
BIC(full_model, reduced_model)
```

These values are plug-in summaries at the posterior mean. They are handy for
continuity with an existing analysis, but WAIC and PSIS-LOO use the full
posterior and are the primary tools here. The distinction matters most for
finite mixtures, which are singular models: the large-sample argument behind
AIC and BIC can fail for them [@Watanabe2010]. For a substantively fixed
class count, fit filled finite models with `classes = 1, 2, ...` to the same
observations and compare them with the workflow below. Sparse finite and
Dirichlet-process mixtures report a posterior over the occupied count
instead; assess them with label-invariant predictive criteria and check
their sensitivity to the concentration prior.

The weight-based update is a different case again. It selects a dimension
heuristically during warmup, so its `n_classes` draws do not represent
posterior uncertainty. WAIC or PSIS-LOO can still assess the resulting
fitted model, but they do not turn the preceding search into Bayesian
inference about the number of classes.

## WAIC and PSIS-LOO

`WAIC()` and `loo()` return objects of the `loo` package [@Vehtari2026]. Both
report the expected log predictive density `elpd`, an effective number of
parameters, and the criterion on the deviance scale. Lower `waic` and
`looic`, or equivalently higher `elpd`, mean better predictive accuracy.
@Watanabe2010 introduced WAIC and showed that it approximates Bayesian
cross-validation asymptotically; @Vehtari2017 and
@Vehtari2024 developed the practical PSIS-LOO approximation and its
diagnostics.

```{r waic}
WAIC(full_model)
WAIC(reduced_model)
```

`WAIC()` warns about large `p_waic` values: some travelers are influential
enough to make the WAIC approximation shaky, which is exactly the situation
in which PSIS-LOO is preferable. It comes with Pareto-k diagnostics. Values
below the printed threshold mean that the importance sampling for that
traveler is reliable, and a few larger values usually point to travelers
whose choices are surprising under the model [@Vehtari2024].

```{r loo}
loo_full <- loo(full_model)
loo_reduced <- loo(reduced_model)
loo_full
```

The `loo` package plots these diagnostics per traveler. Points below the
dashed line are unproblematic; the few above it are the travelers whose
choices the model finds hardest to reproduce when they are left out.

```{r loo-plot}
plot(loo_full)
```

`loo::loo_compare()` ranks the models by `elpd` and reports the difference to
the best model with its standard error. A difference of several standard
errors, as here, is decisive: knowing a traveler's income and party size
clearly improves the prediction of their mode.

```{r compare}
loo::loo_compare(loo_full, loo_reduced)
```

What is it that income changes? `interpret(type = "mea")` differentiates
the probability of every mode at the average traveler, so the rows for
`income` say where an additional thousand euro of household income shifts
the choice:

```{r income-effects}
mode_effects <- interpret(full_model, type = "mea")
mode_effects
```

```{r income-numbers, include=FALSE}
income_effect <- function(mode) {
  row <- mode_effects$covariate == "income" & mode_effects$alternative == mode
  round(100 * mode_effects$mean[row], 1)
}
```

A higher income moves travelers away from the train, by about
`r abs(income_effect("train"))` percentage points per thousand euro, and
towards the plane and the car, by `r income_effect("air")` and
`r income_effect("car")` points. A larger traveling party has the opposite
effect on the plane and the train, which fits the economics of the trip: the
fare is paid per person, the car is paid per trip.

## Bayes factors

`bayes_factor()` estimates the marginal likelihood of each model with the
`bridgesampling` package [@Gronau2020] and returns their
ratio; values above one favor the first model. Bayes factors weigh the prior
predictive performance of the models, so they are sensitive to the prior,
and the priors of both models should be chosen deliberately when a Bayes
factor informs a substantive decision [@Kass1995]. Bridge sampling
estimates the required ratio of normalizing constants [@Meng1996];
@Gronau2017 give a practical tutorial. **RprobitB** deliberately avoids the
harmonic-mean estimator: it is easy to compute from posterior likelihood
draws, but it can have infinite variance and converge far too slowly for
practical use [@Wolpert2012]. The current implementation requires
fixed single-class models; their normalization does not matter, because all
draws are brought to the same one before bridge sampling. Repeated bridge
runs should agree before a result is reported.

```{r bf}
set.seed(1)
log_bf <- bayes_factor(full_model, reduced_model, log = TRUE)
log_bf
```

A log Bayes factor of about `r round(log_bf$bf)` leaves no doubt either:
the data are $e^{`r round(log_bf$bf)`}$ times more probable under the full
model, even though a Bayes factor weighs how well each model predicted the
data before seeing them and the nine extra parameters of the full model have
to earn their keep against the prior. Information criteria and Bayes factors
answer different questions and need not agree; here they do.

Bridge sampling is a Monte Carlo estimate, so the number should be checked
before it is reported. `repetitions` reruns the bridge
estimate several times and reports the spread, which shows how much the
number depends on any single run:

```{r bf-repetitions}
set.seed(2)
bayes_factor(full_model, reduced_model, log = TRUE, repetitions = 3)
```

The second estimate agrees with the first to within a fraction of a log
unit, so the conclusion does not rest on the accident of one bridge run.

## When the likelihood is an integral

For models with random coefficients, the likelihood of a decider integrates
over the coefficient distribution jointly for all their occasions. The
result is a multivariate normal probability with one dimension per occasion
and alternative difference, which for a panel of moderate length quickly
means dozens of dimensions. `oeli::pmvnorm()` evaluates
probabilities of up to three dimensions exactly and higher dimensions with
the GHK simulator on a fixed sequence of `ghk_draws` quasi-random points, so
results are reproducible and smooth in the parameters. More draws reduce the
simulation error at a proportional cost in time. The GHK simulator is
reviewed and evaluated by @Hajivassiliou1996; @Genz2002 compare the
alternative methods for multivariate normal probabilities.

In Mejia, Ecuador, 98 dairy farmers took part in a choice experiment on
water conservation [@Ortiz2023]. Each farmer faced four choice tasks
with two conservation plans and the status quo. The plans differ in
irrigation technology, manure and waste management, whether training is
offered, and a cost-share payment to the farmer; farm size, milk production,
and cattle density describe the farmer. @Ortiz2023 collected these data to
measure what farmers require in return for conservation practices and report
that the cost-share payment raises the acceptance of a plan, which the
estimates below reproduce. Do farmers differ in how much the payment sways
them? `update()` refits the first model with a normal random
effect on the payment coefficient and keeps everything else, including the
data. The panel likelihood of a farmer then integrates over the coefficient,
which makes the evaluation of the criteria noticeably slower than for the
fixed model.

```{r water}
data("water_conservation_choice", package = "choicedata")
water_effects <- choice ~ irrigation + manure + waste + training + payment |
  farm_size + milk_production + cattle_density
fixed_water <- fit(
  water_effects,
  data = water_conservation_choice,
  format = "long",
  column_decider = "farmer",
  column_occasion = "occasion",
  column_alternative = "alternative",
  iterations = 1000,
  warmup = 500,
  chains = 2,
  progress = FALSE
)
mixed_water <- update(fixed_water, random_effects = "payment")
loo_fixed <- loo(fixed_water)
loo_mixed <- loo(mixed_water)
loo::loo_compare(loo_fixed, loo_mixed)
```

Both models are compared on the same farmer-level scale, and the answer is
sobering: letting the payment coefficient vary does not improve the expected
predictive accuracy, it lowers it. The `N < 100` flag and the large standard
error of the difference caution that a comparison based on 98 farmers is
rough, and the warnings about high Pareto-k values concern a few farmers
whose importance weights are unstable. `loo::loo_moment_match()` would
tighten the comparison. Still, the
message makes sense: four choices per farmer carry little information about
individual payment sensitivities, and the extra flexibility costs more than
it returns.

What do the farmers actually want? With the model chosen, `interpret()`
reads the coefficients as the payment that offsets each practice:

```{r water-interpret}
interpret(fixed_water, reference = "payment", effects = "trainingTRUE")
```

Training is worth a few hundred dollars of cost share to these farmers. The
other practices come with intervals several thousand dollars wide, which is
worth a warning about ratios in general: `interpret()` divides by the
reference coefficient draw by draw, and when that coefficient is not safely
away from zero, as the payment coefficient is not here, a few draws blow the
ratio up and the interval becomes useless. Check the reference coefficient
before reading a compensation.

## Ordered and ranked responses

Nothing about the criteria is specific to unordered choices. The ordered
model of `vignette("v02_model_variants")` asked whether older students smoke
less; the comparison below asks whether the exercise variable earns its
three dummy coefficients.

```{r ordered-comparison}
data("survey", package = "MASS")
smoking_full <- fit(
  Smoke ~ Age + Exer | 0,
  data = survey,
  alternatives = c("Never", "Occas", "Regul", "Heavy"),
  choice_type = "ordered",
  column_decider = NULL,
  iterations = 2000,
  warmup = 1000,
  chains = 2,
  progress = FALSE
)
smoking_age <- update(smoking_full, . ~ Age | 0)
loo::loo_compare(
  loo(smoking_full, progress = FALSE), loo(smoking_age, progress = FALSE)
)
```

The difference is smaller than its standard error, so the exercise dummies
do not improve the prediction of how much a student smokes. With fewer than
fifty smokers among the 237 students that is the expected outcome, and it
agrees with the wide credible intervals that the coefficients carried in
`vignette("v02_model_variants")`.

The same works for rankings. Because a ranking of six platforms carries more
information than a single choice, the likelihood units are the 91
respondents, and `logLik()` counts them as such.

```{r ranked-loglik}
data("Game", package = "mlogit")
gaming <- fit(
  ch ~ own | age + hours,
  data = Game,
  alternatives = c(
    "Xbox", "PlayStation", "PSPortable", "GameCube", "GameBoy", "PC"
  ),
  choice_type = "ranked",
  delimiter = ".",
  column_decider = NULL,
  iterations = 2000,
  warmup = 1000,
  chains = 2,
  progress = FALSE
)
logLik(gaming)
nobs(gaming)
```

## Where to go next

A model that survives this comparison is ready to be read for its content.
`vignette("v04_prediction")` turns it into choice probabilities and marginal
effects, and `vignette("v03_heterogeneity")` offers the richer
specifications that a comparison like the one above can then judge.

## References
