---
title: "Bayesian model evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian model evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
link-citations: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
  fig.align = "center"
)
```

This vignette describes how to compare competing model specifications.
**RprobitB** provides three criteria for this purpose: the widely
applicable information criterion (WAIC), Pareto-smoothed importance sampling
leave-one-out cross-validation (PSIS-LOO), and Bayes factors. WAIC and
PSIS-LOO estimate the expected predictive accuracy of a model for new data
from the pointwise log-likelihood of the posterior draws
[@Watanabe2010; @Vehtari2017]. A Bayes factor is the ratio of the marginal
likelihoods of two models, that is, of the likelihood averaged over the
prior of each model [@Kass1995]. In panel data, the pointwise
log-likelihood is evaluated per decider, so WAIC and PSIS-LOO estimate the
accuracy of predicting the choices of new deciders. The examples use data
sets of the **AER** package [@Kleiber2008], the **mlogit** package
[@Croissant2020], and the **MASS** package [@VenablesRipley2002]. The
vignette [Get started with RprobitB][v01] explains how to fit and read a
model, and the vignettes [Model specification and variants][v02] and
[Modeling preference heterogeneity][v03] cover the specifications compared
here.

## A nested model comparison for travel mode choices

The `TravelMode` data of the **AER** package record which of four modes 210
travelers between Sydney and Melbourne had taken: air, train, bus, or car.
The vignette [Model specification and variants][v02] fits a model in which
terminal waiting time, in-vehicle cost, and travel time vary across modes,
while household income and the size of the traveling party shift the
utilities of the modes relative to air, the base alternative. Does income
change the mode choice beyond cost and time? The reduced model below omits
the two traveler characteristics and the alternative-specific constants.
`update()` rebuilds the call of the full model with the formula changed part
by part: `. ~ wait + vcost + travel | 0` keeps the first part and empties the
second. A free error covariance between four alternatives mixes slowly in a
cross section, so the chains run 20000 iterations and retain every twentieth
draw, 1000 draws in total.

```{r fit}
library(RprobitB)
set.seed(1)
data("TravelMode", package = "AER")
TravelMode$choice <- TravelMode$choice == "yes"
TravelMode$vcost <- TravelMode$vcost / 1.6196
TravelMode$income <- TravelMode$income / 1.6196
full_model <- fit(
  choice ~ wait + vcost + travel | income + size,
  data = TravelMode,
  format = "long",
  column_decider = "individual",
  column_alternative = "mode",
  iterations = 20000,
  warmup = 10000,
  thin = 20,
  chains = 2,
  progress = FALSE
)
reduced_model <- update(full_model, . ~ wait + vcost + travel | 0)
```

`logLik()` evaluates the log-likelihood at the posterior means of the
parameters. Its `df` attribute counts the free parameters, without the error
variance that is fixed to identify the utility scale, and its `nobs`
attribute counts the independent likelihood units, here the
`r nobs(full_model)` travelers with one choice each.

```{r loglik}
logLik(full_model)
logLik(reduced_model)
```

The plug-in log-likelihood ignores the posterior uncertainty about the
parameters. WAIC and PSIS-LOO use all posterior draws and are the primary
criteria.

## WAIC and PSIS-LOO

`WAIC()` and `loo()` return objects of the **loo** package [@Vehtari2026].
Both report the expected log predictive density `elpd`, an effective number
of parameters, and the criterion on the deviance scale. Lower `waic` and
`looic`, or equivalently higher `elpd`, mean better predictive accuracy.
@Watanabe2010 introduced WAIC and showed that it asymptotically approximates
Bayesian cross-validation; @Vehtari2017 and @Vehtari2024 developed the
PSIS-LOO approximation and its diagnostics.

```{r waic}
waic_full <- WAIC(full_model)
waic_full
WAIC(reduced_model)
```

```{r waic-numbers, include=FALSE}
waic_large <- sum(waic_full$pointwise[, "p_waic"] > 0.4)
waic_text <- if (waic_large > 0) {
  sprintf(
    "For %d of the %d travelers, the contribution to `p_waic` exceeds 0.4",
    waic_large, nobs(full_model)
  )
} else {
  "For no traveler does the contribution to `p_waic` exceed 0.4"
}
```

`r waic_text`, the level above which @Vehtari2017 consider the WAIC
approximation unreliable and which the **loo** package reports as a
warning. PSIS-LOO is preferable in this situation, because it comes with a
diagnostic per traveler: a Pareto-k value below the printed threshold means
that the importance sampling for that traveler is reliable, and larger
values usually belong to travelers whose choices have low probability under
the model [@Vehtari2024].

```{r loo}
loo_full <- loo(full_model)
loo_reduced <- loo(reduced_model)
loo_full
```

The **loo** package plots the Pareto-k values per traveler. Points above
the dashed line mark the travelers whose choices are hardest to predict from
the choices of the other travelers.

```{r loo-plot}
plot(loo_full)
```

`loo::loo_compare()` ranks the models by `elpd` and reports the difference to
the best model with its standard error.

```{r compare}
travel_comparison <- loo::loo_compare(loo_full, loo_reduced)
travel_comparison
```

```{r compare-numbers, include=FALSE}
travel_best <- if (travel_comparison$model[1] == "model1") {
  "full"
} else {
  "reduced"
}
travel_short <- abs(travel_comparison[2, "elpd_diff"])
travel_ratio <- travel_short / travel_comparison[2, "se_diff"]
```

The `r travel_best` model ranks first, and the other model falls short by
`r round(travel_short)`, `r round(travel_ratio)` times the standard error of
the difference: income and party size
`r if (travel_best == "full") "improve" else "do not improve"` the
prediction of the mode choice.

## Bayes factors

`bayes_factor()` estimates the marginal likelihood of each model with the
**bridgesampling** package [@Gronau2020; @Meng1996] and returns their
ratio; values above one favor the first model. The marginal likelihood
averages the likelihood over the prior, so Bayes factors depend on the
priors, and the priors of both models should be chosen deliberately when a
Bayes factor informs a substantive decision [@Kass1995]. The implementation
requires models without latent classes; their normalization does not
matter, because all draws are transformed to the same normalization before
bridge sampling.

```{r bf}
set.seed(1)
log_bf <- bayes_factor(full_model, reduced_model, log = TRUE)
log_bf
```

```{r bf-numbers, include=FALSE}
extra_parameters <- attr(logLik(full_model), "df") -
  attr(logLik(reduced_model), "df")
bf_favors <- if (log_bf$bf > 0) {
  "also favors the full model"
} else {
  "favors the reduced model"
}
bf_under <- if (log_bf$bf > 0) {
  "it than under the reduced model"
} else {
  "the reduced model than under the full model"
}
bf_agree <- if ((log_bf$bf > 0) == (travel_best == "full")) "do" else "do not"
```

The log Bayes factor of about `r round(log_bf$bf)` `r bf_favors`: the data
are $e^{`r round(abs(log_bf$bf))`}$ times more probable under `r bf_under`,
although the marginal likelihood averages over the prior and thus penalizes
the `r extra_parameters` additional parameters of the full model.
Information criteria and Bayes factors answer different questions and need
not agree; here they `r bf_agree`. Bridge sampling is a Monte Carlo method,
so the estimate should be checked before it is reported. `repetitions`
reruns the bridge estimate and reports the spread across the runs:

```{r bf-repetitions}
set.seed(2)
log_bf_repetitions <- bayes_factor(
  full_model, reduced_model, log = TRUE, repetitions = 3
)
log_bf_repetitions
```

The three repeated estimates lie within
`r signif(max(abs(log_bf_repetitions$bf - log_bf$bf)), 1)` log units of the
first one, so the conclusion does not depend on a single bridge run.

## Models with random coefficients

For models with random coefficients, the likelihood of a decider is the
joint probability of their choices on all occasions, integrated over the
distribution of the random coefficients. The result is a multivariate normal
probability with one dimension per occasion and alternative difference,
which `oeli::pmvnorm()` evaluates exactly for up to three dimensions and
with the GHK simulator on a fixed sequence of quasi-random points otherwise
[@Hajivassiliou1996; @Genz2002]. `ghk_draws` sets the number of points;
more points reduce the simulation error at a proportional cost in computing
time. Because every retained draw requires one such probability per decider,
the fits below are thinned to 100 draws in total.

Does a random price coefficient improve the train model of the vignette
[Get started with RprobitB][v01]? The fixed model is fitted first, and
`update()` gives the price coefficient a normal random effect, as in the
vignette [Posterior prediction][v04].

```{r train}
data("Train", package = "mlogit")
Train$price_A <- Train$price_A / 100 / 2.20371
Train$price_B <- Train$price_B / 100 / 2.20371
Train$time_A <- Train$time_A / 60
Train$time_B <- Train$time_B / 60
train_fixed <- fit(
  choice ~ price + time + change + factor(comfort) | 0,
  data = Train,
  column_decider = "id",
  column_occasion = "choiceid",
  iterations = 6000,
  warmup = 3000,
  thin = 60,
  chains = 2,
  progress = FALSE
)
train_random <- update(train_fixed, random_effects = c(price = "n"))
loo_fixed <- loo(train_fixed, progress = FALSE)
loo_random <- loo(train_random, progress = FALSE)
train_comparison <- loo::loo_compare(loo_fixed, loo_random)
train_comparison
```

```{r train-numbers, include=FALSE}
train_random_first <- train_comparison$model[1] == "model2"
train_short <- abs(train_comparison[2, "elpd_diff"])
train_se <- train_comparison[2, "se_diff"]
train_flagged <- length(union(
  loo::pareto_k_ids(loo_fixed), loo::pareto_k_ids(loo_random)
))
```

Both models are compared per traveler, whose likelihood unit is the joint
probability of all their choices. The model with the
`r if (train_random_first) "random" else "fixed"` price coefficient ranks
first, and the other model falls short by `r round(train_short)` with a
standard error of `r round(train_se)`, that is, by
`r round(train_short / train_se)` standard errors. Letting the price
sensitivity vary between travelers thus
`r if (train_random_first) "improves" else "does not improve"` the
prediction of a traveler's choices. The Pareto-k diagnostic flags
`r train_flagged` travelers under at least one of the two models; their
contributions are approximations with heavier-tailed importance weights, not
wrong numbers.

## Ordered and ranked responses

The criteria are not specific to unordered choices. The ordered model of the
vignette [Model specification and variants][v02] asked whether older
students smoke less; the comparison below asks whether the exercise dummies
improve the prediction.

```{r ordered-comparison}
data("survey", package = "MASS")
smoking_full <- fit(
  Smoke ~ Age + Exer | 0,
  data = survey,
  alternatives = c("Never", "Occas", "Regul", "Heavy"),
  choice_type = "ordered",
  column_decider = NULL,
  iterations = 2000,
  warmup = 1000,
  chains = 2,
  progress = FALSE
)
smoking_age <- update(smoking_full, . ~ Age | 0)
smoking_comparison <- loo::loo_compare(
  loo(smoking_full, progress = FALSE), loo(smoking_age, progress = FALSE)
)
smoking_comparison
```

```{r ordered-comparison-numbers, include=FALSE}
smoking_short <- abs(smoking_comparison[2, "elpd_diff"])
smoking_se <- smoking_comparison[2, "se_diff"]
smoking_small <- smoking_short < smoking_se
smokers <- sum(survey$Smoke != "Never", na.rm = TRUE)
```

The difference of `r round(smoking_short, 1)` is
`r if (smoking_small) "smaller" else "larger"` than its standard error of
`r round(smoking_se, 1)`, so the exercise dummies
`r if (smoking_small) "do not improve" else "improve"` the prediction of how
much a student smokes. With `r smokers` smokers among the
`r nobs(smoking_full)` students who answered the question, this is the
expected outcome. Ranked models are evaluated in the same way, with each
complete ranking as one likelihood unit.

## Further reading

The vignette [Posterior prediction][v04] computes choice probabilities and
marginal effects from a fitted model, and the vignette
[Modeling preference heterogeneity][v03] describes the random coefficient
and latent class specifications that the criteria above can compare.

[v01]: https://loelschlaeger.de/RprobitB/articles/v01_get_started.html
[v02]: https://loelschlaeger.de/RprobitB/articles/v02_model_variants.html
[v03]: https://loelschlaeger.de/RprobitB/articles/v03_heterogeneity.html
[v04]: https://loelschlaeger.de/RprobitB/articles/v04_prediction.html

## References
