---
title: "Bayesian model evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian model evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
link-citations: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
  fig.align = "center"
)
```

This vignette describes how to compare competing model specifications.
**RprobitB** provides three criteria for this purpose: the widely
applicable information criterion (WAIC), Pareto-smoothed importance sampling
leave-one-out cross-validation (PSIS-LOO), and Bayes factors. WAIC and
PSIS-LOO estimate the expected predictive accuracy of a model for new data
from the pointwise log-likelihood of the posterior draws
[@Watanabe2010; @Vehtari2017]. A Bayes factor is the ratio of the marginal
likelihoods of two models, that is, of the likelihood averaged over the
prior of each model [@Kass1995]. In panel data, the pointwise
log-likelihood is evaluated per decider, so WAIC and PSIS-LOO estimate the
accuracy of predicting the choices of new deciders. The examples use data
sets of the **AER** package [@Kleiber2008], the **mlogit** package
[@Croissant2020], and the **MASS** package [@VenablesRipley2002]. The
vignette [Get started with RprobitB][v01] explains how to fit and read a
model, and the vignettes [Model specification and variants][v02] and
[Modeling preference heterogeneity][v03] cover the specifications compared
here.

```{r setup}
library(RprobitB)
set.seed(1)
```

## A nested model comparison for travel mode choices

The `TravelMode` data of the **AER** package record which of four modes 210
travelers between Sydney and Melbourne had taken: air, train, bus, or car.
The vignette [Model specification and variants][v02] fits a model in which
terminal waiting time, in-vehicle cost, and travel time vary across modes,
while household income and the size of the traveling party shift the
utilities of the modes relative to air, the base alternative. 

Does income
change the mode choice beyond cost and time? The reduced model below omits
the two traveler characteristics and the alternative-specific constants.
`update()` rebuilds the call of the full model with the formula changed part
by part: `. ~ wait + vcost + travel | 0` keeps the first part and empties the
second.

```{r fit}
data("TravelMode", package = "AER")
TravelMode$choice <- TravelMode$choice == "yes"
TravelMode$vcost <- TravelMode$vcost / 1.6196
TravelMode$income <- TravelMode$income / 1.6196
full_model <- fit(
  choice ~ wait + vcost + travel | income + size,
  data = TravelMode,
  format = "long",
  column_decider = "individual",
  column_alternative = "mode",
  iterations = 20000,
  warmup = 10000,
  thin = 20,
  chains = 2,
  progress = FALSE
)
reduced_model <- update(full_model, . ~ wait + vcost + travel | 0)
```

`logLik()` evaluates the log-likelihood at the posterior means of the
parameters. Its `df` attribute counts the free parameters, without the error
variance that is fixed to identify the utility scale, and its `nobs`
attribute counts the independent likelihood units, here the
`r nobs(full_model)` travelers with one choice each.

```{r loglik}
logLik(full_model)
logLik(reduced_model)
```

The log-likelihood values can be used to compute AIC and BIC, but it ignores the 
posterior uncertainty about the parameters. WAIC and PSIS-LOO instead use all 
posterior draws and are the primary criteria.

## WAIC and PSIS-LOO

`WAIC()` and `loo()` return objects of the **loo** package [@Vehtari2026].
Both report the expected log predictive density `elpd`, an effective number
of parameters, and the criterion on the deviance scale. Lower `waic` and
`looic`, or equivalently higher `elpd`, mean better predictive accuracy.
@Watanabe2010 introduced WAIC and showed that it asymptotically approximates
Bayesian cross-validation; @Vehtari2017 and @Vehtari2024 developed the
PSIS-LOO approximation and its diagnostics.

```{r waic}
WAIC(full_model)
WAIC(reduced_model)
```

The **loo** package warns when the contribution of a decider to `p_waic`
exceeds 0.4, the level above which @Vehtari2017 consider the WAIC
approximation unreliable; here this concerns a few travelers. PSIS-LOO is
preferable in this situation,
because it comes with a diagnostic per traveler: a Pareto-k value below the
printed threshold means that the importance sampling for that traveler is
reliable, and larger values usually belong to travelers whose choices have
low probability under the model [@Vehtari2024].

```{r loo}
loo_full <- loo(full_model)
loo_reduced <- loo(reduced_model)
loo_full
```

The **loo** package plots the Pareto-k values per traveler. Points above
the dashed line mark the travelers whose choices are hardest to predict from
the choices of the other travelers.

```{r loo-plot}
plot(loo_full)
```

`loo::loo_compare()` ranks the models by `elpd` and reports the difference to
the best model with its standard error.

```{r compare}
loo::loo_compare(loo_full, loo_reduced)
```

The models are named in the order of the arguments, so `model1` is the full
model. It ranks first, and the reduced model falls short by several
standard errors of the difference: income and party size improve the
prediction of the mode choice.

## Bayes factors

`bayes_factor()` estimates the marginal likelihood of each model with the
**bridgesampling** package [@Gronau2020; @Meng1996] and returns their
ratio; values above one favor the first model.

```{r bf}
set.seed(1)
bayes_factor(full_model, reduced_model, log = TRUE)
```

The large positive log Bayes factor also favors the full model.

## Models with random coefficients

Does a random price coefficient improve the train model of the vignette
[Get started with RprobitB][v01]? The fixed model is fitted first, and
`update()` gives the price coefficient a normal random effect, as in the
vignette [Posterior prediction][v04].

```{r train}
data("Train", package = "mlogit")
Train$price_A <- Train$price_A / 100 / 2.20371
Train$price_B <- Train$price_B / 100 / 2.20371
Train$time_A <- Train$time_A / 60
Train$time_B <- Train$time_B / 60
train_fixed <- fit(
  choice ~ price + time + change + factor(comfort) | 0,
  data = Train,
  column_decider = "id",
  column_occasion = "choiceid",
  iterations = 6000,
  warmup = 3000,
  thin = 60,
  chains = 2,
  progress = FALSE
)
train_random <- update(train_fixed, random_effects = c(price = "n"))
loo_fixed <- loo(train_fixed, progress = FALSE)
loo_random <- loo(train_random, progress = FALSE)
loo::loo_compare(loo_fixed, loo_random)
```

The model with the random price
coefficient, `model2`, ranks first, and the fixed model falls short by
several standard errors of the difference. Letting the price
sensitivity vary between travelers thus improves the prediction of a
traveler's choices.

## Ordered and ranked responses

The criteria are not specific to unordered choices. The ordered model of the
vignette [Model specification and variants][v02] asked whether older
students smoke less; the comparison below asks whether the exercise dummies
improve the prediction.

```{r ordered-comparison}
data("survey", package = "MASS")
smoking_full <- fit(
  Smoke ~ Age + Exer | 0,
  data = survey,
  alternatives = c("Never", "Occas", "Regul", "Heavy"),
  choice_type = "ordered",
  column_decider = NULL,
  iterations = 2000,
  warmup = 1000,
  chains = 2,
  progress = FALSE
)
smoking_age <- update(smoking_full, . ~ Age | 0)
loo::loo_compare(
  loo(smoking_full, progress = FALSE), loo(smoking_age, progress = FALSE)
)
```

The difference is smaller than its standard error, so the exercise dummies 
appear to not improve the prediction of how much a student smokes.

## Further reading

The vignette [Posterior prediction][v04] computes choice probabilities and
marginal effects from a fitted model, and the vignette
[Modeling preference heterogeneity][v03] describes the random coefficient
and latent class specifications that the criteria above can compare.

[v01]: https://loelschlaeger.de/RprobitB/articles/v01_get_started.html
[v02]: https://loelschlaeger.de/RprobitB/articles/v02_model_variants.html
[v03]: https://loelschlaeger.de/RprobitB/articles/v03_heterogeneity.html
[v04]: https://loelschlaeger.de/RprobitB/articles/v04_prediction.html

## References
