---
title: "Data preparation of pairfam"
author: "Lennart Oelschläger"
date: "`r Sys.Date()`"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

The purpose of this document is to prepare the pairfam dataset for modeling contraception choice and child wish.

## The pairfam dataset

*Pairfam* means *Panel Analysis of Intimate Relationships and Family Dynamics* and is a longitudinal study of German family processes. The data is collected annually via surveys from a nationwide random sample of the three birth cohorts 1971-73, 1981-83, 1991-93. 

## Read SPSS data and save in R-format

```{r}
### save data of each wave (11) as one list element
data <- list()
wave_label <- c("2008/09","2009/10","2010/11","2011/12","2012/13","2013/14","2014/15","2015/16","2016/17","2017/18","2018/19")
for(wave in 1:11){
  ### path to achors of the 'wave'th wave
  path <- paste0("data/SPSS/English/anchor",wave,".sav")
  ### convert from SPSS to R
  data[[wave]] <- foreign::read.spss(file = path, to.data.frame = TRUE)
  ### relabel wave
  data[[wave]][,"wave"] <- wave_label[wave]
  ### save columns as character except for id column
  data[[wave]][,-1] <- dplyr::mutate_all(data[[wave]][,-1], as.character)
}
### bind rows of waves together
data <- dplyr::bind_rows(data)
### sort rows by id
data <- dplyr::arrange(data, id)
### save data
saveRDS(data, "data/data_from_spss.rds")
```

### Filter for relevant covariates

```{r}
data <- readRDS("data/data_from_spss.rds")
data <- dplyr::select(data,
  id, # person id
  wave, # wave number
  age, # age
  sex_gen, # sex
  relstat, # relation status
  sex5, # use of contraception
  sex8, # frequency of sexual intercourse
  sex6i1,sex6i2,sex6i3,sex6i4,sex6i5,sex6i6,sex6i7,sex6i8,sex6i9,sex6i10,sex6i11, # contraception methods,
  frt1, # infertility of anchor
  frt2, # infertility of partner
  frt7, # intention to become mother/father within next 2 years,
  frt11v2i1, # can afford to be a parent financially
  frt11v2i2, # child is compatible with long-term life plans
  frt11v2i3, # has access to flexible child care options
  frt11v2i4, # child is compatible with current employment situation
  frt11v2i5, # child is compatible with leisure-time interests
  frt11v2i6, # my partner is able to combine a child with his employment situation
  frt11v2i7, # my partner and I agree regarding the division of household labor and child care
  frt11v2i8 # my partner and I both feel ready to have an additional child
)
saveRDS(data,"data/data_filtered.rds")
```

### Encode covariates

```{r}
data <- readRDS("data/data_filtered.rds")

### id
data[,"id"] <- as.numeric(data[,"id"])

### wave
data[,"wave"] <- data[,"wave"]

### age
data[,"age"] <- as.numeric(data[,"age"])

### sex8: frequency of sexual intercourse
### new cov name: sex_freq
### levels: 0 (never had sex) to 7 (daily)
for(row in 1:nrow(data)){
  data[row,"sex8"] <- stringr::str_extract(data[row,"sex8"],"[[:digit:]]+")
}
data[,"sex_freq"] <- as.numeric(data[,"sex8"])
data <- dplyr::select(data, !sex8)

### sex_gen: Gender
### old levels: 1 Male and 2 Female
### new cov name: gender
### new levels: 0 (male), 1 (female)
for(row in 1:nrow(data)){
  temp <- data[row, "sex_gen"]
  if(grepl("Female", temp, ignore.case = FALSE, fixed = TRUE)) {
    data[row, "sex_gen"] <- 1
  } else if (grepl("Male", temp, ignore.case = FALSE, fixed = TRUE)) {
    data[row, "sex_gen"] <- 0
  } else {
    data[row, "sex_gen"] <- NA
  }
}
data[,"gender"] <- as.numeric(data[,"sex_gen"])
data <- dplyr::select(data, !sex_gen)

### rel_stat: Relationship status
### Original levels: 
### -  1 Never married single
### -  2 Never married living apart together
### -  3 Never married cohabiting
### -  4 Married cohabiting
### -  5 Married non-cohabiting
### -  6 Divorced/separated single
### -  7 Divorced/separated living apart together
### -  8 Divorced/separated cohabiting
### -  9 Widowed single
### - 10 Widowed living apart together
### - 11 Widowed cohabiting
### New levels: 
### -  0 single (1, 6, 7, 9)
### -  1 relationship (2, 3, 4, 5, 8, 10, 11)
for(row in 1:nrow(data)){
  x <- as.numeric(stringr::str_extract(data[row,"relstat"],"[[:digit:]]+"))
  if(x %in% c(2,3,4,5,8,10,11)){
    data[row,"relstat"] <- 1
  } else if(x %in% c(1,6,7,9)){
    data[row,"relstat"] <- 0
  } else {
    data[row,"relstat"] <- NA
  }
}
data[,"rel_stat"] <- as.numeric(data[,"relstat"])
data <- dplyr::select(data, !relstat)

### Contraception: Did you or your partner use some form of contraception in the past three months? 
### Covariate: sex5
### Levels: 1 (yes), 2 (no)
### New covariate: contr_used (contraception usage)
### New levels: 0 (no), 1 (yes)
for(row in 1:nrow(data)){
  temp <- data[row,"sex5"]
  if(grepl("Yes", temp, ignore.case = FALSE, fixed = TRUE)) {
    data[row,"sex5"]  <- 1
  } else if (grepl("No", temp, ignore.case = FALSE, fixed = TRUE)) {
    data[row,"sex5"]  <- 0
  } else {
    data[row,"sex5"]  <- NA
  }
}
data[,"contr_used"] <- as.numeric(data[,"sex5"])
data <- dplyr::select(data, !sex5)

### Contraception methods: What method(s) did you use primarily? Multiple answers are possible.
### Covariates (each level 0: not mentioned and 1: mentioned):
### - Birth control pill, mini-pill (sex6i1)
### - Condom (sex6i2)
### - Hormone preparations (sex6i3)
### - Intrauterine devise (IUD) (sex6i4)
### - Diaphragm, foam, suppository, gel (sex6i5)
### - Natural birth control (standard days method, rhythm method) (sex6i6)
### - Hysterectomy/female sterilization (sex6i7)
### - Vasectomy/male sterilization (sex6i8)
### - Withdrawal method, coitus interruptus (sex6i9)
### - The morning-after pill (sex6i10)
### - Something else (sex6i11)
### New covariates:
### - contr_method (levels: combination of '*' in 'sex6i*' and '0' for none)
### - pill (levels: 1 if 'sex6i1' is 1 and 0 else)
### - condom (levels: 1 if 'sex6i2' is 1 and 0 else)
contraception_cov <- paste0("sex6i",1:11)
for(row in 1:nrow(data)){
  for(cov in contraception_cov){
    data[row,cov] <- stringr::str_extract(data[row,cov],"[[:digit:]]+")
  }
}
for(cov in contraception_cov){
  data[,cov] <- as.numeric(data[,cov])
}
for(row in 1:nrow(data)){
  contr_method <- paste(which(data[row,contraception_cov]==1), collapse = "|")
  if(contr_method == "") contr_method <- "0"
  data[row,"contr_method"] <- contr_method
}
data[, "pill"] <- data[, "sex6i1"]
data[, "condom"] <- data[, "sex6i2"]
data <- dplyr::select(data, !all_of(contraception_cov))

### Intention to become mother/father within next 2 years
### Covariate: frt7
### Original levels: 
### -  1 Yes, definitely
### -  2 Yes, perhaps
### -  3 No, probably not
### -  4 No, definitely not
### -  7 I haven't thought about that yet
### new cov name: child_wish
### New levels: 
### -  1 yes (1, 2)
### -  0 no (3, 4)
for(row in 1:nrow(data)){
  x <- as.numeric(stringr::str_extract(data[row,"frt7"],"[[:digit:]]+"))
  if(x %in% c(1,2)){
    data[row,"frt7"] <- 1
  } else if(x %in% c(3,4)){
    data[row,"frt7"] <- 0
  } else {
    data[row,"frt7"] <- NA
  }
}
data[,"child_wish"] <- as.numeric(data[,"frt7"])
data <- dplyr::select(data, !frt7)

### frt1, frt2: infertility of anchor or partner
### new covariate name: infert
### levels: 1 (either anchor or partner is definitely infertible) 0 else
for(row in 1:nrow(data)){
  x <- as.numeric(stringr::str_extract(data[row,"frt1"],"[[:digit:]]+"))
  y <- as.numeric(stringr::str_extract(data[row,"frt2"],"[[:digit:]]+"))
  if(is.na(x) && is.na(y)){
    data[row,"infert"] <- NA
  } else if((!is.na(x) && x == 4) || (!is.na(y) && y == 4)){
    data[row,"infert"] <- 1
  } else {
    data[row,"infert"] <- 0
  }
}
data <- dplyr::select(data, !frt1)
data <- dplyr::select(data, !frt2)

### frt11v2i1:8: different conditions to become a child
### new cov names: child_cond1:8
### old levels: 1 (yes) 2 (no)
### new levels: 1 (yes) 0 (no)
vars <- paste0("frt11v2i",1:8)
for(row in 1:nrow(data)){
  for(i in 1:length(vars)){
    temp <- data[row,vars[i]]
    if(grepl("Yes", temp, ignore.case = FALSE, fixed = TRUE)) {
      data[row,vars[i]] <- 1
    } else if (grepl("No", temp, ignore.case = FALSE, fixed = TRUE)) {
      data[row,vars[i]] <- 0
    } else {
      data[row,vars[i]] <- NA
    }
  }
}
for(i in 1:length(vars)){
  data[,paste0("child_cond",i)] <- as.numeric(data[,vars[i]])
}
data <- dplyr::select(data, !all_of(vars))

saveRDS(data,"data/data_encoded.rds")
```
